### 2022-08-19 tpd-22.tex
...

\documentclass[10pt,journal,compsoc]{IEEEtran}

\usepackage{verbatim}
\usepackage{url}
\usepackage{graphicx}

\usepackage{xspace}
\newcommand{\FC}       {Freechains\xspace}
\newcommand{\reps}     {\emph{reps}\xspace}
\newcommand{\onerep}   {\emph{1~rep}\xspace}
\newcommand{\nreps}[1] {\emph{#1~reps\xspace}}
\newcommand{\code}[1]  {\texttt{\footnotesize{#1}}}
\newcommand{\Xon} {$1{\rightarrow}N$\xspace}
\newcommand{\Xno} {$1{\leftarrow}N$\xspace}
\newcommand{\Xnn} {$N{\leftrightarrow}N$\xspace}
\newcommand{\Xoo} {$1{\leftrightarrow}1$\xspace}
\newcommand{\Xo}  {$1{\hookleftarrow}$\xspace}

\renewcommand{\theenumi}{\alph{enumi}}

\hyphenation{off-line}

\begin{document}

\title{
    Symmetric Peer-to-Peer Applications
}

\author{
    Francisco Sant'Anna~\IEEEmembership{Department of Computer Science, Rio de Janeiro State University}
}

\IEEEtitleabstractindextext{%
\begin{abstract}
In multi-node collaborative applications, users can interact asynchronously
and share the exact same experience remotely.
%
In this work, we propose a middleware for \emph{symmetric peer-to-peer
applications} in which decentralized instances can broadcast asynchronous
events and yet conform to identical behavior.
%
Nodes are allowed to join and leave the network at any time.
Application developers must adhere to a restricted API, which is
deterministic, stateless, and only supports pre-allocated memory.
%
The middleware is responsible for synchronizing the events in a global shared
timeline across the network.
Based on memory snapshots and deterministic simulation, a ``time machine''
rolls back and resynchronizes the state of conflicting nodes.
%
TODO: application, results
\end{abstract}

\begin{IEEEkeywords}
collaboration, determinism, peer-to-peer, time machine
\end{IEEEkeywords}}

\maketitle

% TOTAL: 12 pages

\section{Introduction}
\label{sec.introduction}

Collaborative networked applications allow multiple users to interact
remotely and yet share the same experience in real time.
Examples of such \emph{symmetric distributed applications}~\cite{gals}
are shared documents, watch parties, and multi-player games.
%
In order to reproduce the exact behavior in multiple devices, the application
must be able to synchronize its time and execution across the network.
In addition, since users can interact asynchronously with the application,
event occurrences must somehow be synchronized back across devices.

A common approach towards symmetric applications is to introduce a middleware
to orchestrate events and time in the network~\cite{gals,croquet}.
Applications developers rely on middleware primitives to trigger events, which
are intercepted and synchronized in a global shared timeline across the
network.
Developers must also restrict themselves to deterministic and stateless calls
only, such that execution can be equally reproduced in all nodes according to
the shared timeline.
However, current solutions depend on a central server to interconnect network
nodes and determine the shared timeline.

In this work, we propose \emph{symmetric peer-to-peer applications}, which
do not rely on central servers for coordination.
Nodes in the application form a dynamic peer-to-peer network and communicate
only with their neighbours directly.
Events are flooded in the network and are triggered locally with a small
delay to cope with latency.
Our main contribution is a time machine that can rollback nodes to a previous
state to apply events received out of order or too late.

TODO: application, results

In Section~\ref{sec.related}, we revisit existing solutions for symmetric
distributed applications.
In Section~\ref{sec.tml}, we detail our proposed middleware.
In Section~\ref{sec.app}, ...
In Section~\ref{sec.eval}, ...
In Section~\ref{sec.conclusion}, ...

- assumptions
    - correct (non-malicious) nodes

- local first
- p2p
- less restrictive (compared to CRDTs)
- semantic events
    - limitation b/c of rollbacks
    - features b/c of network traffic

\section{Related Work}
\label{sec.related}

In this section, we revisit existing solutions for symmetric distributed
applications.
We focus on (i) how the network is organized, (ii) how events are propagated,
and (iii) how global time is determined.
Figure~\ref{fig.related} compares selected works regarding these aspects.

Croquet%
\footnote{Croquet.io: https://croquet.io/}~\cite{TODO} guarantees
bit-identical real-time behavior for every user in a collaborative
distributed environment.
%
As illustrated in Figure~\ref{fig.croquet}, a centralized \emph{reflector}
issues periodic ticks, such that all nodes in the network advance together
according to a synchronized global shared clock.
If an user generates an event, it is delayed and sent back to the reflector,
which broadcasts the event in the next tick, which all nodes apply in sync
(including the originating node).
%
Croquet takes periodic snapshots of the whole application state in order to
support late joins to a running session.
The new node just needs to restore the latest snapshot and simulate the
remaining events to reach the current running state.
%
As indicated in Figure~\ref{fig.related}, Croquet relies on a centralized
network, in which nodes advance in sync, and in which event outcomes depend
on the originating node to be online.

GALS~\cite{TODO} shares the same goals with Croquet, but with some tradeoffs,
mostly favoring clients with slow connections.
As illustrated in Figure~\ref{fig.gals}, instead of advancing ticks in sync
with the server, clients have their own local clocks.
Also, event generation is delayed dynamically according to the slowest client.
%
On the one hand, ticks do not generate any traffic and clients have smooth
frame transitions, even those with poor connections.
On the other hand, event generation requires two round trips to the server and
clients may experience occasional freezes.
The syncing protocol also requires bookkeeping to deal with clock drifts and
client disconnections.
%
As indicated in Figure~\ref{fig.related}, GALS relies on a centralized
network, in which nodes advance independently, and in which event outcomes
depend on the originating node to be online.

In both solutions, the application advances in real time, with a total order
among events, which is determined by a centralized server that must be
permanently online.

An antagonistic approach is

- two strong restrictions:
   - very restrictive API
   - not real time animations b/c no consensus on timestamps

- offline, local first
- DAG, no consensus, partial order
    - branches in parallel w/ T,T+d not required to be applied in order

- all still deterministic, stateless

architecture and dy-
namic programming environment with a synchronization
protocol called TeaTime, which more recently became a prod-
uct8 . Croquet uses a similar centralized architecture to our
work, but takes advantage of the cloud to place the server as
close to clients as possible for better performance. In contrast
to our work, Croquet relies on the server to issue tick events
to clients, taking full control over the timeline. The server
also timestamps events coming from the clients and redirect
them to all participants. This results in a simpler protocol
that supports joins and leaves transparently, and does not
require roundtrips to determine timestamps. However, since
all tick messages need to be transmitted over the network,
the FPS rate of applications is proportional to the network
latency. For instance, a latency of 100ms supports at most 10
FPS. The transmission of ticks at higher FPS rates also incur
in higher network traffic and server load. Another distinc-
tion between the protocols is that Croquet favors instances
with better latency, which observe optimal experience. In
contrast, our protocol adjusts event latency dynamically to
accommodate the slower instance. In addition, instances ex-
ecute at the same tick rate respecting local times, regardless
of network latency.

Croquet OS is a.
- centralized
    - synchronous
        - croquet
	- PALS
    - asynchronous
        - gals
- decentralized
    - CRDTs (asynchronous) (not realtime) (pause would not be bit identical)
    - rollback (async)

- local-first + CRDTs + pushpin

- time machine
    - paper que eu achei ha um tempo
    - braid

Regarding the structure of messages, append-only Merkle~DAGs have been proposed
as self-certified archives, and as CRDTs that provide strong eventual
consistency~\cite{fed.matrix,p2p.byz}.
However, when these DAGs are open to permissionless writes, they are subject to
abuse, which may degrade the content and performance of the network.
Another aspect is that a DAG itself has no attached semantics, which weakens
its consistency property at the application level, which may interpret the DAG
in conflicting orders.
The proposed consensus mechanism addresses permissionless writes and provides
a total order for applications (at the cost of rollbacks).

A similar dVCS have been recently proposed~\cite{p2p.dvcs}, with a DAG
representation and merging policies.
To resolve \emph{competition conflicts}, they propose to use an external
scoring function, such as users' reputations.
Our proposed consensus mechanism internalizes such reputation scoring function.
Integrated reputation also prevents SPAM and abusive behavior from byzantine
nodes, which could otherwise generate very large graphs that take forever to
synchronize (as discussed for Byzantine Causal Broadcast~\cite{p2p.dag.sync}).

\section{The API \& Middleware}
\label{sec.conclusion}

\section{Evaluation}
\label{sec.conclusion}

\section{Conclusion}
\label{sec.conclusion}

\bibliographystyle{IEEEtran}
\bibliography{tpd-22}

%\begin{comment}
\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{chico}}]{Francisco Sant'Anna}
received his PhD degree in Computer Science from PUC-Rio, Brazil in
2013.
In 2016, he joined the Faculty of Computer Science at the Rio de Janeiro State
University, Brazil.
His research interests include Programming Languages and Concurrent \&
Distributed Systems.
\end{IEEEbiography}
%\end{comment}

\end{document}

### 2022-08-19 tpd-22.tex
- related

@@
 and (iii) how global time is determined.
-Figure~\ref{fig.related} compares selected works regarding these aspects.
+Figure~\ref{fig.table} compares selected works regarding these aspects.
 
@@
 %
-As illustrated in Figure~\ref{fig.croquet}, a centralized \emph{reflector}
-issues periodic ticks, such that all nodes in the network advance together
-according to a synchronized global shared clock.
+As detailed in Figure~\ref{fig.algos}, a centralized \emph{reflector} issues
+periodic ticks, such that all nodes in the network advance together according
+to a synchronized global shared clock.
 If an user generates an event, it is delayed and sent back to the reflector,
@@
 %
-As indicated in Figure~\ref{fig.related}, Croquet relies on a centralized
+As indicated in Figure~\ref{fig.table}, Croquet relies on a centralized
 network, in which nodes advance in sync, and in which event outcomes depend
@@
 mostly favoring clients with slow connections.
-As illustrated in Figure~\ref{fig.gals}, instead of advancing ticks in sync
-with the server, clients have their own local clocks.
+As detailed in Figure~\ref{fig.algos}, instead of advancing ticks in sync with
+the server, clients have their own local clocks.
 Also, event generation is delayed dynamically according to the slowest client.
@@
 %
-As indicated in Figure~\ref{fig.related}, GALS relies on a centralized
+As indicated in Figure~\ref{fig.table}, GALS relies on a centralized
 network, in which nodes advance independently, and in which event outcomes
@@
 permanently online.
+All clients compute the events in sequence, respecting their timestamps, and
+using only deterministic and stateless calls.
+This way, it is guaranteed that all clients reach the same state, going
+through bit-identical steps.
 
-An antagonistic approach is
-
-- two strong restrictions:
-   - very restrictive API
-   - not real time animations b/c no consensus on timestamps
+An antagonistic approach to computing a series of event occurrences
+deterministically, is to model the application data structures with
+conflict-free replicated data types (CRDTs)~\cite{TODO}.
+CRDTs must be designed in such a way that all operations are commutative, so
+that the order in which they are applied does not affect the final outcome.
+%
+As detailed in Figure~\ref{fig.algos}, peers can communicate operations
+directly to each other with no central authority.
+Also, since operations need not to be ordered, they can be applied at the
+very first moment they are generated or received, even if the peers are
+offline.
+%
+Since operations must be commutative, it is not possible to timestamp them
+in a global shared clock.
+Hence, there is no notion of a timeline with peers going through bit-identical
+steps, but they do eventually reach the same state.
+%
+On the one hand, the main advantage of CRDTs is they support the concept of
+local-first software~\cite{TODO}, since they work offline.
+On the other hand, they provide very restrictive APIs (the CRDTs themselves),
+and do not support real-time consensus among peers.
+%
+Automerge~\cite{TODO} along with its accompanying Hypermerge peer-to-peer
+infrastructure is an example of an implementation of this approach.
 
@@
 - all still deterministic, stateless
-
-architecture and dy-
-namic programming environment with a synchronization
-protocol called TeaTime, which more recently became a prod-
-uct8 . Croquet uses a similar centralized architecture to our
-work, but takes advantage of the cloud to place the server as
-close to clients as possible for better performance. In contrast
-to our work, Croquet relies on the server to issue tick events
-to clients, taking full control over the timeline. The server
-also timestamps events coming from the clients and redirect
-them to all participants. This results in a simpler protocol
-that supports joins and leaves transparently, and does not
-require roundtrips to determine timestamps. However, since
-all tick messages need to be transmitted over the network,
-the FPS rate of applications is proportional to the network
-latency. For instance, a latency of 100ms supports at most 10
-FPS. The transmission of ticks at higher FPS rates also incur
-in higher network traffic and server load. Another distinc-
-tion between the protocols is that Croquet favors instances
-with better latency, which observe optimal experience. In
-contrast, our protocol adjusts event latency dynamically to
-accommodate the slower instance. In addition, instances ex-
-ecute at the same tick rate respecting local times, regardless
-of network latency.
 
@@
     - braid
-
-Regarding the structure of messages, append-only Merkle~DAGs have been proposed
-as self-certified archives, and as CRDTs that provide strong eventual
-consistency~\cite{fed.matrix,p2p.byz}.
-However, when these DAGs are open to permissionless writes, they are subject to
-abuse, which may degrade the content and performance of the network.
-Another aspect is that a DAG itself has no attached semantics, which weakens
-its consistency property at the application level, which may interpret the DAG
-in conflicting orders.
-The proposed consensus mechanism addresses permissionless writes and provides
-a total order for applications (at the cost of rollbacks).
-
-A similar dVCS have been recently proposed~\cite{p2p.dvcs}, with a DAG
-representation and merging policies.
-To resolve \emph{competition conflicts}, they propose to use an external
-scoring function, such as users' reputations.
-Our proposed consensus mechanism internalizes such reputation scoring function.
-Integrated reputation also prevents SPAM and abusive behavior from byzantine
-nodes, which could otherwise generate very large graphs that take forever to
-synchronize (as discussed for Byzantine Causal Broadcast~\cite{p2p.dag.sync}).
 

### 2022-08-19 tpd-22.tex
- algos
- table.png

@@
 
+% docs.google.com/spreadsheets/d/1CpMmEgabJq2XQeTDW_BNCV4JWrcMlELuaEjDtw7kwg0/
+\begin{figure}[t]
+  \centering
+  \includegraphics[width=\linewidth]{table}
+  \caption{
+    \label{fig.table}
+  }
+\end{figure}
+
+\begin{figure*}[t]
+  \centering
+  \includegraphics[width=\linewidth]{algos}
+  \caption{
+    \label{fig.algos}
+  }
+\end{figure*}
+
 In this section, we revisit existing solutions for symmetric distributed
 applications.
-We focus on (i) how the network is organized, (ii) how events are propagated,
-and (iii) how global time is determined.
+We focus on (i) how the network is organized, (ii) how global time is
+determined, and (iii) how events are propagated.
 Figure~\ref{fig.table} compares selected works regarding these aspects.
@@
 
-An antagonistic approach to computing a series of event occurrences
-deterministically, is to model the application data structures with
-conflict-free replicated data types (CRDTs)~\cite{TODO}.
-CRDTs must be designed in such a way that all operations are commutative, so
-that the order in which they are applied does not affect the final outcome.
+An antagonistic approach to deterministic reactions to events is to model the
+application data structures with conflict-free replicated data types
+(CRDTs)~\cite{TODO}.
+CRDTs are designed in such a way that all operations are commutative, so that
+the order in which they are applied does not affect the final outcome.
 %
@@
 infrastructure is an example of an implementation of this approach.
+
+In this work, we try to .
 
@@
 \section{The API \& Middleware}
-\label{sec.conclusion}
+\label{sec.api}
 
 \section{Evaluation}
-\label{sec.conclusion}
+\label{sec.eval}
 
@@
 \bibliographystyle{IEEEtran}
-\bibliography{tpd-22}
+%\bibliography{tpd-22}
 
-%\begin{comment}
+\begin{comment}
 \begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{chico}}]{Francisco Sant'Anna}
@@
 \end{IEEEbiography}
-%\end{comment}
+\end{comment}
 

### 2022-08-19 tpd-22.tex
...

@@
 %
+As indicated in Figure~\ref{fig.table}, CRDTs work on decentralized networks,
+in which nodes advance independently, and in which operations can be applied
+immediately, even while offline.
+%
 Automerge~\cite{TODO} along with its accompanying Hypermerge peer-to-peer
@@
 
-In this work, we try to .
+In this work, our goal is to support peer-to-peer real-time symmetric
+execution, while still tolerating offline and out-of-order event generation.
+%
+As detailed in Figure~\ref{fig.algos}, peers have their own timeline and can
+communicate events directly to each other.
+Events are delayed to be able to reach other peers in time.
+In the case deadlines are missed, the peer rolls back in time and simulates
+the execution in fast forward until it reaches real time again.
+
+, in which nodes advance independently, and in which event outcomes
+depend on the originating node to be online.
+
+we try to .
 

### 2022-08-19 tpd-22.tex
...

--- 2022-08-19 tpd-22.tex
@@
 
-Collaborative networked applications allow multiple users to interact
-remotely and yet share the same experience in real time.
+\IEEEPARstart{C}{ollaborative} networked applications allow multiple users to
+interact remotely and yet share the same experience in real time.
 Examples of such \emph{symmetric distributed applications}~\cite{gals}
@@
 Croquet%
-\footnote{Croquet.io: https://croquet.io/}~\cite{TODO} guarantees
+\footnote{Croquet.io: https://croquet.io/}~\cite{croquet} guarantees
 bit-identical real-time behavior for every user in a collaborative
@@
 
-GALS~\cite{TODO} shares the same goals with Croquet, but with some tradeoffs,
+GALS~\cite{gals} shares the same goals with Croquet, but with some tradeoffs,
 mostly favoring clients with slow connections.
@@
 application data structures with conflict-free replicated data types
-(CRDTs)~\cite{TODO}.
+(CRDTs)~\cite{crdts}.
 CRDTs are designed in such a way that all operations are commutative, so that
@@
 On the one hand, the main advantage of CRDTs is they support the concept of
-local-first software~\cite{TODO}, since they work offline.
+local-first software~\cite{local}, since they work offline.
 On the other hand, they provide very restrictive APIs (the CRDTs themselves),
@@
 %
-Automerge~\cite{TODO} along with its accompanying Hypermerge peer-to-peer
-infrastructure is an example of an implementation of this approach.
+As indicated in Figure~\ref{fig.table}, CRDTs work on decentralized networks,
+in which nodes advance independently, and in which operations can be applied
+immediately, even while offline.
+%
+Automerge along with its accompanying Hypermerge peer-to-peer infrastructure
+is an example of an implementation of this approach~\cite{p2p.automerge,p2p.pushpin}.
 
-In this work, we try to .
+In this work, as indicated in Figure~\ref{fig.table}, our goal is to support
+peer-to-peer real-time symmetric execution, while still tolerating offline and
+out-of-order event generation.
+%
+As detailed in Figure~\ref{fig.algos}, peers have independent timelines and
+can communicate events directly to each other.
+Events are delayed to be able to reach other peers in time.
+In the case a deadline is missed, the peer rolls back in time and fast
+forwards execution until it reaches real time again.
+%
+Like Croquet and GALS, peers compute events in sequence using only
+deterministic and stateless calls.
 
-- offline, local first
-- DAG, no consensus, partial order
-    - branches in parallel w/ T,T+d not required to be applied in order
-
-- all still deterministic, stateless
-
-Croquet OS is a.
-- centralized
-    - synchronous
-        - croquet
-	- PALS
-    - asynchronous
-        - gals
-- decentralized
-    - CRDTs (asynchronous) (not realtime) (pause would not be bit identical)
-    - rollback (async)
-
-- local-first + CRDTs + pushpin
-
+TODO
+- photon unreal
 - time machine
@@
 \bibliographystyle{IEEEtran}
-%\bibliography{tpd-22}
+\bibliography{tpd-22}
 

### 2022-08-20 tpd-22.tex
...

@@
 Events are delayed to be able to reach other peers in time.
-In the case a deadline is missed, the peer rolls back in time and fast
+In the case a deadline is missed, the peer rolls back in time and then fast
 forwards execution until it reaches real time again.
@@
 
-TODO
-- photon unreal
+Fusion%
+\footnote{Fusion: \url{https://doc.photonengine.com/en-us/fusion}}
+is a state synchronization networking library for the Unity game engine.
+It provides networked tick-based simulation in two modes: client-server
+(\emph{hosted mode}) or peer-to-peer (\emph{shared mode}).
+The hosted mode is based on continuous memory snapshots, which allows for
+full rollbacks when clients diverge.
+However, the shared mode is less powerful and only supports eventual
+consistency among clients.
+We presume that continuous memory snapshots would be too expensive without a
+central server.
+
+Regarding single-user time machines, the video game Braid%
+\footnote{\url{Braid: http://braid-game.com/}}
+is designed on the assumption that players have unrestricted power to go back
+in time.
+Like Fusion, it's implementation is based on continuous memory snapshots, and
+not deterministic simulation as we propose in this work.
+
 - time machine
@@
 \section{The API \& Middleware}
-\label{sec.api}
+\label{sec.tml}
+
+\section{Application}
+\label{sec.app}
 

### 2022-08-20 tpd-22.tex
...

--- 2022-08-19 tpd-22.tex
@@
 \begin{abstract}
-In multi-node collaborative applications, users can interact asynchronously
-and share the exact same experience remotely.
+In multi-node collaborative applications, users can interact remotely and
+share the exact same experience as if they were in a single machine.
 %
@@
 timeline across the network.
-Based on memory snapshots and deterministic simulation, a ``time machine''
-rolls back and resynchronizes the state of conflicting nodes.
+Based on memory snapshots and deterministic simulation, a ``time machine'' can
+rollback conflicting nodes to resynchronize their state.
 %
@@
 interact remotely and yet share the same experience in real time.
-Examples of such \emph{symmetric distributed applications}~\cite{gals}
-are shared documents, watch parties, and multi-player games.
+Examples of these \emph{symmetric distributed applications}~\cite{gals} are
+shared documents, watch parties, and multi-player games.
 %
 In order to reproduce the exact behavior in multiple devices, the application
-must be able to synchronize its time and execution across the network.
+must be able to synchronize time and execution across the network.
 In addition, since users can interact asynchronously with the application,
@@
 to orchestrate events and time in the network~\cite{gals,croquet}.
-Applications developers rely on middleware primitives to trigger events, which
-are intercepted and synchronized in a global shared timeline across the
-network.
+This way, applications developers can rely on middleware primitives to trigger
+events, which are intercepted and synchronized in a global shared timeline
+across the network.
 Developers must also restrict themselves to deterministic and stateless calls
@@
 do not rely on central servers for coordination.
-Nodes in the application form a dynamic peer-to-peer network and communicate
-only with their neighbours directly.
-Events are flooded in the network and are triggered locally with a small
-delay to cope with latency.
-Our main contribution is a time machine that can rollback nodes to a previous
-state to apply events received out of order or too late.
+Nodes in the application form a dynamic graph network and communicate only
+with direct neighbours.
+Events are flooded in the graph and are triggered locally with a small delay
+to consider the network latency.
+To deal with events received out of order or too late, a distributed time
+machine can rollback nodes to a previous state and reapply events in order and
+in time.
 
@@
 We focus on (i) how the network is organized, (ii) how global time is
-determined, and (iii) how events are propagated.
+determined, and (iii) how events are propagated and applied.
 Figure~\ref{fig.table} compares selected works regarding these aspects.
@@
 network, in which nodes advance in sync, and in which event outcomes depend
-on the originating node to be online.
+on the central server to be online.
 
@@
 The syncing protocol also requires bookkeeping to deal with clock drifts and
-client disconnections.
+client disconnections~\cite{gals}.
 %
@@
 network, in which nodes advance independently, and in which event outcomes
-depend on the originating node to be online.
+depend on the central server to be online.
 
-In both solutions, the application advances in real time, with a total order
-among events, which is determined by a centralized server that must be
-permanently online.
+In both solutions, the network advances together as a whole in real time, with
+a total order among events, which is determined by a central server that must
+be permanently online.
 All clients compute the events in sequence, respecting their timestamps, and
 using only deterministic and stateless calls.
-This way, it is guaranteed that all clients reach the same state, going
+This way, it is guaranteed that all clients reach the same state, also going
 through bit-identical steps.
@@
 An antagonistic approach to deterministic reactions to events is to model the
-application data structures with conflict-free replicated data types
-(CRDTs)~\cite{crdts}.
+application with conflict-free replicated data types (CRDTs)~\cite{crdts}.
 CRDTs are designed in such a way that all operations are commutative, so that
@@
 in a global shared clock.
-Hence, there is no notion of a timeline with peers going through bit-identical
-steps, but they do eventually reach the same state.
+Hence, there is no notion of a timeline in which peers go through
+bit-identical steps, but they do eventually reach the same state.
 %
-On the one hand, the main advantage of CRDTs is they support the concept of
-local-first software~\cite{local}, since they work offline.
+On the one hand, the main advantage of CRDTs is the support for local-first
+software~\cite{local}, since they can work offline in the same way as online.
 On the other hand, they provide very restrictive APIs (the CRDTs themselves),
@@
 Automerge along with its accompanying Hypermerge peer-to-peer infrastructure
-is an example of an implementation of this approach~\cite{p2p.automerge,p2p.pushpin}.
+is an example of an implementation of decentralized
+CRDTs~\cite{p2p.automerge,p2p.pushpin}.
 
@@
 Events are delayed to be able to reach other peers in time.
-In the case a deadline is missed, the peer rolls back in time and fast
-forwards execution until it reaches real time again.
+In the case a deadline is missed, the peer rolls back in time and then fast
+forwards execution until it catches real-time behavior again.
 %
@@
 
-TODO
-- photon unreal
-- time machine
-    - paper que eu achei ha um tempo
-    - braid
+Regarding decentralized time machines, Fusion%
+\footnote{Fusion: \url{https://doc.photonengine.com/en-us/fusion}}
+is a state synchronization networking library for the Unity game engine.
+It provides networked tick-based simulation in two modes: client-server
+(\emph{hosted mode}) and peer-to-peer (\emph{shared mode}).
+The hosted mode is based on continuous memory snapshots, which allows for
+full rollbacks when clients diverge.
+However, the shared mode is less powerful and only supports eventual
+consistency among clients.
+We presume that continuous memory snapshots would be too costly without a
+central server.
+
+Regarding single-user time machines, the video game Braid%
+\footnote{\url{Braid: http://braid-game.com/}}
+is designed on the assumption that players have unrestricted rollbacks as part
+of the game mechanics.
+Like Fusion, the implementation is based on continuous memory snapshots,
+instead of deterministic simulation as we propose in this work.
+%
+Schuster~\cite{tml.js} proposes a simple API for time travelling in
+event-driven JavaScript:
+    \code{init()} to return an initial state;
+    \code{handle(evt,mem)} to process events and modify state; and
+    \code{render(mem)} to output the current state.
+As we describe next, we use a similar approach by separating mutation from
+rendering~\cite{tml.alive}.
 
 \section{The API \& Middleware}
-\label{sec.api}
+\label{sec.tml}
+
+\section{Application}
+\label{sec.app}
 

### 2022-08-20 tpd-22.tex
...

@@
 \begin{abstract}
-In multi-node collaborative applications, users can interact remotely and
-share the exact same experience as if they were in a single machine.
+In multi-node collaborative applications, such as shared online documents,
+users can interact remotely and yet share the exact same experience as if they
+were in a single machine.
 %
@@
 
+In this section, we describe the architecture of our middleware and also the
+API in which developers need to conform.
+%
+We guide our discussion through a didactic example in which multiple users use
+the arrow keys to change the direction of a moving rectangle.
+
+- same keys, same behavior, pause at the same time
+
+we show how a
+
+- sym
+- ids
+- walk back-forth
+- vector clocks
+- dynamic net
+- re-connect
+- late join
+
 \section{Application}

### 2022-08-21 tpd-22.tex
...

@@
 Based on memory snapshots and deterministic simulation, a ``time machine'' can
-rollback conflicting nodes to resynchronize their state.
+rollback conflicting peers to resynchronize their state.
 %
@@
 To deal with events received out of order or too late, a distributed time
-machine can rollback nodes to a previous state and reapply events in order and
+machine can rollback peers to a previous state and reapply events in order and
 in time.
@@
     \label{fig.algos}
+    Four approaches to symmetric distributed applications: Croquet, GALS,
+    CRDTs, and this work.
+    The vertical arrows in parallel represent the timelines in nodes in the
+    network.
+    The arrows crossing nodes represent communication between them.
   }
@@
 As indicated in Figure~\ref{fig.table}, CRDTs work on decentralized networks,
-in which nodes advance independently, and in which operations can be applied
+in which peers advance independently, and in which operations can be applied
 immediately, even while offline.
@@
 
-\section{The API \& Middleware}
+\section{The Middleware \& Programming API}
 \label{sec.tml}
@@
 In this section, we describe the architecture of our middleware and also the
-API in which developers need to conform.
+API in which developers have to conform with.
+
+We guide our discussion through the didactic example of Figure~\ref{fig.move},
+in which multiple users use the arrow keys to control the direction of a
+moving rectangle.
+The six instances are connected in a peer-to-peer network, such that some
+peers can only communicate indirectly.
 %
-We guide our discussion through a didactic example in which multiple users use
-the arrow keys to change the direction of a moving rectangle.
+The application is symmetric in the sense that if any peer presses a key, an
+event is broadcast and applied in all peers simultaneously, as if they were a
+single mirrored instance.
+At any moment, if a peer pauses the application, all peers are guaranteed to
+pause exactly at the same frame with the rectangle at the same position.
+%
+A video
+
+\begin{figure}[t]
+    \centering
+    \includegraphics[width=\linewidth]{move}
+    \caption[XXX] {
+	A moving rectangle is controlled by peers connected indirectly.
+	Users can press \code{↑ ↓ → ←} to change the direction of the
+	rectangle or \code{SPACE} to pause the application.
+	The inputs are applied simultaneously in all peers, as if they were
+	mirrors of a single application.
+        Video: \url{http://youtube.com/TODO}
+        \label{fig.move}
+    }
+\end{figure}
 

### 2022-08-21 tpd-22.tex
...

@@
 %
-The application is symmetric in the sense that if any peer presses a key, an
-event is broadcast and applied in all peers simultaneously, as if they were a
-single mirrored instance.
+The application is symmetric in the sense that if any peer presses a key, a
+corresponding event is broadcast and applied in all peers simultaneously, as
+if they were a single mirrored instance.
 At any moment, if a peer pauses the application, all peers are guaranteed to
-pause exactly at the same frame with the rectangle at the same position.
+pause exactly at the same frame with the rectangle ``bit-identically'' at the
+same position.
+
+The main job of the middleware is to deal with the uncertainties of the
+network in order to preserve the symmetric behavior across peers.
+More specifically, the middleware ensures that all peers
+    (i)   meet event deadlines;
+    (ii)  advance in sync in real time;
+    (iii) can leave and join the network and remain symmetric.
+We detail how the middleware preserves these properties further.
+
+\subsection{The Programming API}
+\label{sec.tml.api}
+
+The middleware expects to take full control of the application execution in
+order to control its timeline, and also communicate events in the network.
+For this reason, the basic API is to call an \code{init} function passing a
+set of callbacks, which the middleware uses to communicate with the
+application, and then call \code{loop} to yield control to the middleware:
+
+{\footnotesize
+\begin{verbatim}
+int main (void) {
+    ...  // initialize the UI
+    p2p_init(1, 50, sizeof(G), &G, cb_sim, cb_eff, cb_rec);
+    p2p_link("182.168.0.2", 5000, 2);
+    p2p_loop();
+    p2p_quit();
+    ...  // finalize the UI
+}
+\end{verbatim}
+}
+
+The middleware relies on the library \emph{SDL} to deal with the network and
+input events, such as timers, mouse clicks and key presses.
+The code that follows is a typical:
+
+{\footnotesize
+\begin{verbatim}
+// preallocated full application state
+struct {
+    ...
+} G;
+
+// executes a simulation step (handles ticks or events)
+void cb_sim (p2p_evt evt) {
+    switch (evt.id) {
+        ...
+    }
+}
+
+// executes the simulation effects (e.g., redraw the screen)
+void cb_eff (int trv) {
+    ...
+}
+
+// generates events
+int cb_rec (SDL_Event* sdl, p2p_evt* evt) {
+    switch (sdl->type) {
+    	...
+    }
+}
+
+\end{verbatim}
+}
+
+for computer graphics, and
+\emph{Arduino} for embedded systems.%
+\footnote{SDL: \url{www.libsdl.org}. Arduino: \url{www.arduino.cc}.}
+This allows for an easier integration of our middleware with practical systems,
+and also reinforces how, in our proposal, programming distributed versions
+becomes similar to their local counterparts.
+
+{\footnotesize
+\begin{verbatim}
+void p2p_init (
+    uint8_t me, int port, int fps,
+    int mem_n, void* mam_app,
+    void (*cb_sim) (p2p_evt),           // simulation callback
+    void (*cb_eff) (int trv),           // effects callback
+    int (*cb_rec) (SDL_Event*,p2p_evt*) // event recording callback
+);
+void p2p_link (char* host, int port, uint8_t me);
+void p2p_quit (void);
+void p2p_loop (void);
+
+void p2p_bcast (uint32_t tick, p2p_evt* evt);
+void p2p_dump  (void);
+
+ > freechains '#p2p.md' dislike 3_AE3A1B.. --sign=USR1
+ > freechains '#p2p.md' dislike 3_AE3A1B.. --sign=USR2
+ > freechains '#p2p.md' dislike 3_AE3A1B.. --sign=USR3
+ > freechains-vcs '#p2p.md' checkout p2p.md
+ > cat p2p.md
+ P2p networking is...
+ The [USENET](#usenet.md), ...  <-- no 3rd line
+\end{verbatim}
+}
+
 %
-A video
+\begin{itemize}
+\item \textbf{Event Deadlines:}
+Due to the inherent latency of networks, peers generate events with deliberate
+delays so that they reach other peers in time to be applied in sync in a
+common deadline.
+However, a peer that is locally at time \code{T} may receive an event \code{E}
+that should have been applied at time \code{T-t}.
+In this case, the middleware automatically rolls back the peer \code{t} units
+of time and applies \code{E} at the correct time.
+It then fast forwards the peer \code{t} units of time, re-applying all events
+that occurred in between \code{T-t} and \code{T}.
+%
+\item \textbf{Time Synchronization:}
+Each peer has its own timeline initiated as soon as the application is
+launched.
+Also, the internal clock of each peer may diverge over time.
+Hence, the timelines will inevitably be out of sync eventually.
+The middleware synchronizes clocks continuously as follows:
+Every time a
+
+		Each peer also
+%
+\item[Peer Churn] Arrival and Departure of peers. received events, time sync, all deadlines were missed, departure no problem
+\end{itemize}
+- time sync
+
+The middleware handles:
+- time sync
+- disc / conn
+- late event
 
@@
 
-- same keys, same behavior, pause at the same time
-
-we show how a
-
-- sym
-- ids
 - walk back-forth
@@
 - late join
+
+We make some assumptions about the network, which are assigned externally to
+the middleware:
+%
+\begin{itemize}
+\item Peers have contiguous numeric unique identifiers.
+\item Peers identifiers are assigned externally to the middleware.
+\item Topology is dynamic, but also assigned externally
+
+\end{itemize}
+%
+Except for the last item, we claim that the others can be solved externally,
+without affecting the middleware.
+For instance,
+- orthogonal and would not affect the other algorithms in the middleware
+
+, which we claim that could be
+solved further:
+- peers have unique identifiers
+- ids are outside
+- topology, dynamic but outside
+- fixed maximum nodes
+- fixed latency
+We claim that Use external protocols
+For instance, latency is only a constant, no problem being a variable
+
+- non-malicious nodes
 

### 2022-08-22 tpd-22.tex
...

@@
 The middleware expects to take full control of the application execution in
-order to control its timeline, and also communicate events in the network.
-For this reason, the basic API is to call an \code{init} function passing a
-set of callbacks, which the middleware uses to communicate with the
-application, and then call \code{loop} to yield control to the middleware:
+order to control its timeline, and also to communicate events in the network.
+For this reason, the basic programming API is to call a \code{loop} function
+passing a set of callbacks, which the middleware uses to communicate with the
+application:
 
@@
 int main (void) {
-    ...  // initialize the UI
-    p2p_init(1, 50, sizeof(G), &G, cb_sim, cb_eff, cb_rec);
-    p2p_link("182.168.0.2", 5000, 2);
-    p2p_loop();
-    p2p_quit();
-    ...  // finalize the UI
+    p2p_loop (
+        1,              // peer identifier
+        50,             // simulation FPS
+        sizeof(G), &G,  // pre-allocated full state
+        cb_ini,         // init/quit callback
+        cb_sim,         // simulation callback
+        cb_ren,         // rendering callback
+        cb_evt          // event triggering callback
+    );
 }
@@
 }
+
+The middleware assumes that each peer is assigned a unique identifier in a
+contiguous range.
+The FPS must be the same in all peers to ensure that the simulation executes
+exactly the same.
+The variable \code{G} must hold the full application state.
+If dynamic allocation is required, it is necessary to hold a finite memory
+pool in \code{G} with a custom allocator.
+For instance, the moving rectangle application only requires to hold the
+current position and direction:
+
+{\footnotesize
+\begin{verbatim}
+struct {
+    int x,  y;   // position
+    int dx, dy;  // direction speed
+} G;
+int main (void) { ... }
+\end{verbatim}
+}
+
+The middleware calls \code{cb\_ini} twice: at the beginning and at the end of
+the loop.
+It should be used to initialize (and finalize) the network topology, as well
+as static globals that live outside the simulation memory, such as the window
+and image textures.
+
+{\footnotesize
+\begin{verbatim}
+void cb_ini (int ini) {
+    if (ini) {
+        // create the window
+        // open arrow images
+        // create peer links (peer-id, ip, port)
+        p2p_link(2, "192.168.1.2", 5000);
+    } else {
+        // create the window
+        // open arrow images
+        // destroy a peer links
+        p2p_unlink(2);
+    }
+}
+\end{verbatim}
+}
+
+The middleware calls \code{cb\_sim} on every frame to affect the simulation
+state \code{G} deterministically.
+The callback must never perform any side effects, such as rendering the frame.
+The calls normally happen during real-time simulation, i.e., while the user is
+interacting with the application.
+However, if the peer is out of sync, the middleware may ``time travel'' and
+call the simulation many times to go back and forth and catch up with real
+time.
+This is the reason why \code{cb\_sim} should
+
+This is why it is important that the simulation is detached from the view.
+
+\subsection{The Time Machine}
+\label{sec.tml.time}
+
+at the same time as the user
+is interacting with the application
+
+twice: at the beginning and at the end of
+the loop.
+It should be used to initialize (and finalize) the network topology, as well
+as static globals that live outside the simulation memory, such as the window
+and image textures.
+
+The core of the API are the 4 callbacks
+
+The network topology is dynamic
+
+%\item Peers have contiguous numeric unique identifiers.
 

### 2022-08-22 tpd-22.tex
...

@@
 
+Under the hood, the middleware relies on the library \emph{SDL}%
+\footnote{SDL: \url{www.libsdl.org}}
+to deal with the network and input events, such as timers, mouse clicks and
+key presses.
+
 The middleware assumes that each peer is assigned a unique identifier in a
@@
 It should be used to initialize (and finalize) the network topology, as well
-as static globals that live outside the simulation memory, such as the window
-and image textures.
+as static globals that can live outside the simulation memory, such as the
+window and image textures.
 
@@
     if (ini) {
-        // create the window
-        // open arrow images
+        // create the SDL window
+        // open the arrow PNG images
         // create peer links (peer-id, ip, port)
@@
     } else {
-        // create the window
-        // open arrow images
-        // destroy a peer links
+        // destroy the SDL window
+        // destroy the arrow PNG images
+        // destroy peer links
         p2p_unlink(2);
@@
 The middleware calls \code{cb\_sim} on every frame to affect the simulation
-state \code{G} deterministically.
-The callback must never perform any side effects, such as rendering the frame.
+state of \code{G} deterministically.
+The callback must never perform any side effects, such as stateful calls or
+rendering the frame.
 The calls normally happen during real-time simulation, i.e., while the user is
@@
 time.
-This is the reason why \code{cb\_sim} should
+This is the reason why \code{cb\_sim} should be split from the rendering
+callback \code{cb\_ren}~\cite{tml.js}.
+%, which the middleware calls only on real-time frames.
 
-This is why it is important that the simulation is detached from the view.
+cbsim
+
+In our example, this callback just needs to redraw the rectangle at the
+current position:
+
+{\footnotesize
+\begin{verbatim}
+void cb_eff (void) {
+    // clears the SDL window
+    // redraws the rectangle at the current position
+    SDL_Rect r = { G.x, G.y, 20, 20 };
+    SDL_DrawRect(&r);
+}
+\end{verbatim}
+}
+
+Finally, the \code{cb\_evt} callback is called by the middleware in real time,
+whenever a local SDL event occurs.
+This callback has two goals:
+    (i)  broadcast application events, and
+    (ii) provide instant feedback to the user.
+Not all local low-level events may generate application events to be broadcast
+in the network.
+The callback is allowed to perform side effects, such as modifying the network
+topology, or exhibiting some feedback on the screen.
+In our example, we only generate application events for the 5 key events, and
+we also exhibit the pressed key in real time:
+
+{\footnotesize
+\begin{verbatim}
+int cb_evt (SDL_Event* sdl, p2p_evt* evt) {
+    if (sdl->type == SDL_KEYDOWN) {
+        int key = sdl->key.keysym.sym;
+	if (key == <any-of-the-keys>) {
+            SDL_DrawImage(<img-of-the-key>);
+            *evt = (p2p_evt) { P2P_EVT_KEY, key };
+            return 1; // generate application event
+	}
+    }
+    return 0; // otherwise, does not generate any event
+}
+\end{verbatim}
+}
 
@@
 
-at the same time as the user
-is interacting with the application
-
-twice: at the beginning and at the end of
-the loop.
-It should be used to initialize (and finalize) the network topology, as well
-as static globals that live outside the simulation memory, such as the window
-and image textures.
-
-The core of the API are the 4 callbacks
-
-The network topology is dynamic
+- p2p links por fora, como um servico TCP local
+- The network topology is dynamic
 
 %\item Peers have contiguous numeric unique identifiers.
-
-The middleware relies on the library \emph{SDL} to deal with the network and
-input events, such as timers, mouse clicks and key presses.
-The code that follows is a typical:
-
-{\footnotesize
-\begin{verbatim}
-// preallocated full application state
-struct {
-    ...
-} G;
-
-// executes a simulation step (handles ticks or events)
-void cb_sim (p2p_evt evt) {
-    switch (evt.id) {
-        ...
-    }
-}
-
-// executes the simulation effects (e.g., redraw the screen)
-void cb_eff (int trv) {
-    ...
-}
-
-// generates events
-int cb_rec (SDL_Event* sdl, p2p_evt* evt) {
-    switch (sdl->type) {
-    	...
-    }
-}
-
-\end{verbatim}
-}
-
-for computer graphics, and
-\emph{Arduino} for embedded systems.%
-\footnote{SDL: \url{www.libsdl.org}. Arduino: \url{www.arduino.cc}.}
-This allows for an easier integration of our middleware with practical systems,
-and also reinforces how, in our proposal, programming distributed versions
-becomes similar to their local counterparts.
-
-{\footnotesize
-\begin{verbatim}
-void p2p_init (
-    uint8_t me, int port, int fps,
-    int mem_n, void* mam_app,
-    void (*cb_sim) (p2p_evt),           // simulation callback
-    void (*cb_eff) (int trv),           // effects callback
-    int (*cb_rec) (SDL_Event*,p2p_evt*) // event recording callback
-);
-void p2p_link (char* host, int port, uint8_t me);
-void p2p_quit (void);
-void p2p_loop (void);
-
-void p2p_bcast (uint32_t tick, p2p_evt* evt);
-void p2p_dump  (void);
-
- > freechains '#p2p.md' dislike 3_AE3A1B.. --sign=USR1
- > freechains '#p2p.md' dislike 3_AE3A1B.. --sign=USR2
- > freechains '#p2p.md' dislike 3_AE3A1B.. --sign=USR3
- > freechains-vcs '#p2p.md' checkout p2p.md
- > cat p2p.md
- P2p networking is...
- The [USENET](#usenet.md), ...  <-- no 3rd line
-\end{verbatim}
-}
 

### 2022-08-24 tpd-22.tex
...

@@
 applications.
-We focus on (i) how the network is organized, (ii) how global time is
-determined, and (iii) how events are propagated and applied.
+We focus on (a) how the network is organized, (b) how global time is
+determined, and (c) how events are propagated and applied.
 Figure~\ref{fig.table} compares selected works regarding these aspects.
@@
 At any moment, if a peer pauses the application, all peers are guaranteed to
-pause exactly at the same frame with the rectangle ``bit-identically'' at the
-same position.
+pause exactly at the same frame with the rectangle being ``bit-identically'' at
+the same position.
 
@@
 More specifically, the middleware ensures that all peers
-    (i)   meet event deadlines;
-    (ii)  advance in sync in real time;
-    (iii) can leave and join the network and remain symmetric.
-We detail how the middleware preserves these properties further.
+    (a) meet event deadlines,
+    (b) advance in sync in real time, and
+    (c) can leave and join the network and remain symmetric:
+%
+\begin{enumerate}
+\item \textbf{Event Deadlines:}
+Due to the inherent latency of networks, peers generate events with a
+deliberate delay so that they reach other peers in time to be applied in sync
+in a common deadline.
+%We first assume a predetermined delay for the sake of simplicity.
+However, a peer that is locally ahead of time may receive an event that should
+have been applied in the past, event considering the delay.
+In this case, the middleware automatically rolls back the peer, applies the
+event at the correct time, and then fast forwards the peer, re-applying all
+remaining events up to the real time.
+%
+\item \textbf{Time Synchronization:}
+Since peers run in different machines, their timelines will inevitably be out
+of sync because:
+    (a) applications are launched locally at different times, and
+    (b) internal clocks may diverge over time.
+The middleware assumes the maximum time among all peers the be the
+``correct real time'', so that other peers must fast forward to catch up with
+it.
+Therefore, in order to determine the maximum time and synchronize the clocks,
+the middleware proceeds as follows:
+    (a) peers broadcast \code{SYNC} events periodically with their local
+        times, and
+    (b) if a peer is behind a received \code{SYNC}, it advances its frames
+        proportionally faster to catch up in time.
+%
+\item \textbf{Peer Churn:}
+Considering that we target peer-to-peer networks, it is important that the
+middleware can handle arrival and departure of peers seamlessly.
+TODO
+\end{enumerate}
 
@@
 exactly the same.
-The variable \code{G} must hold the full application state.
+The global variable \code{G} must hold the full application state.
 If dynamic allocation is required, it is necessary to hold a finite memory
@@
 as static globals that can live outside the simulation memory, such as the
-window and image textures.
+window and image textures:
 
@@
 The callback must never perform any side effects, such as stateful calls or
-rendering the frame.
-The calls normally happen during real-time simulation, i.e., while the user is
-interacting with the application.
+rendering the frame:
+The callback receives the occurring input event if any.
+There are two predetermined events:
+    \code{P2P\_EVT\_INIT} called at the beginning of the simulation, and
+    \code{P2P\_EVT\_TICK} called at every frame.
+The application-specific events must be declared by the programmer in an
+enumeration.
+In our application, we define a new event \code{P2P\_EVT\_KEY} to represent key
+presses:
+
+{\footnotesize
+\begin{verbatim}
+enum {
+    P2P_EVT_KEY = P2P_EVT_NEXT
+};
+\end{verbatim}
+}
+
+Our callback \code{cb\_sim}
+    (a) resets the rectangle position and speed on \code{INIT},
+    (b) increments the position based on the current speed on \code{TICK}, and
+    (c) modifies the speed based on \code{KEY}:
+
+{\footnotesize
+\begin{verbatim}
+void cb_sim (p2p_evt evt) {
+    switch (evt.id) {
+        case P2P_EVT_INIT:  // resets position and speed
+            G.x = G.y = G.dx = G.dy = 0;
+            break;
+        case P2P_EVT_TICK:  // increases position
+            G.x += G.dx * VEL;
+            G.y += G.dy * VEL;
+            break;
+        case P2P_EVT_KEY:   // modifies speed
+            switch (evt.pay.i1) {
+                case SDLK_UP:
+                    G.dx = 0;
+                    G.dy = -1;
+                    break;
+                <...> // similar to DOWN/LEFT/RIGHT/SPACE
+            }
+            break;
+    }
+}
+\end{verbatim}
+}
+
+The calls to \code{cb\_sim} normally happen during real-time simulation, i.e.,
+while the user is interacting with the application.
 However, if the peer is out of sync, the middleware may ``time travel'' and
@@
 time.
+This is the reason why we define \code{INIT} as an event: time travelling
+requires to simulate, since the
+
 This is the reason why \code{cb\_sim} should be split from the rendering
@@
 %, which the middleware calls only on real-time frames.
-
-cbsim
-
 In our example, this callback just needs to redraw the rectangle at the
@@
 This callback has two goals:
-    (i)  broadcast application events, and
-    (ii) provide instant feedback to the user.
+    (a) broadcast application events, and
+    (b) provide instant feedback to the user.
 Not all local low-level events may generate application events to be broadcast
@@
     if (sdl->type == SDL_KEYDOWN) {
-        int key = sdl->key.keysym.sym;
-	if (key == <any-of-the-keys>) {
+        if (sdl->key == <any-of-the-keys>) {
             SDL_DrawImage(<img-of-the-key>);
-            *evt = (p2p_evt) { P2P_EVT_KEY, key };
+            *evt = (p2p_evt) { P2P_EVT_KEY, sdl->key };
             return 1; // generate application event
@@
 }
+
+\subsection{TODO}
+
+TODO: simpler than vector clocks, absolute times are correct
+TODO: piggyback SYNC in normal inputs
 
@@
 %
-\begin{itemize}
-\item \textbf{Event Deadlines:}
-Due to the inherent latency of networks, peers generate events with deliberate
-delays so that they reach other peers in time to be applied in sync in a
-common deadline.
-However, a peer that is locally at time \code{T} may receive an event \code{E}
-that should have been applied at time \code{T-t}.
-In this case, the middleware automatically rolls back the peer \code{t} units
-of time and applies \code{E} at the correct time.
-It then fast forwards the peer \code{t} units of time, re-applying all events
-that occurred in between \code{T-t} and \code{T}.
-%
-\item \textbf{Time Synchronization:}
-Each peer has its own timeline initiated as soon as the application is
-launched.
-Also, the internal clock of each peer may diverge over time.
-Hence, the timelines will inevitably be out of sync eventually.
-The middleware synchronizes clocks continuously as follows:
-Every time a
-
-		Each peer also
-%
-\item[Peer Churn] Arrival and Departure of peers. received events, time sync, all deadlines were missed, departure no problem
-\end{itemize}
 - time sync

### 2022-08-24 tpd-22.tex
...

@@
 API in which developers have to conform with.
+The middleware is written in C and relies on the library \emph{SDL}%
+\footnote{SDL: \url{www.libsdl.org}}
+to deal with the network and input events, such as timers, mouse clicks and
+key presses.
+The full source code%
+\footnote{TODO}
+is under 500 lines of code.
 
@@
 middleware can handle arrival and departure of peers seamlessly.
-TODO
+When a peer leaves the network, nothing needs to be done.
+The middleware just needs to ensure that a peer that joins the network receives
+all events in order, which is trivial since they all hold an absolute
+timestamp.
 \end{enumerate}
@@
 The middleware expects to take full control of the application execution in
-order to control its timeline, and also to communicate events in the network.
+order to control its timeline and disseminate events in the network.
 For this reason, the basic programming API is to call a \code{loop} function
@@
 }
-
-Under the hood, the middleware relies on the library \emph{SDL}%
-\footnote{SDL: \url{www.libsdl.org}}
-to deal with the network and input events, such as timers, mouse clicks and
-key presses.
 
@@
     if (ini) {
-        // create the SDL window
-        // open the arrow PNG images
+        <...> // create the SDL window
+        <...> // open the arrow PNG images
         // create peer links (peer-id, ip, port)
@@
     } else {
-        // destroy the SDL window
-        // destroy the arrow PNG images
+        <...> // destroy the SDL window
+        <...> // destroy the arrow PNG images
         // destroy peer links
@@
     switch (evt.id) {
-        case P2P_EVT_INIT:  // resets position and speed
+        case P2P_EVT_INIT: // resets pos and speed
             G.x = G.y = G.dx = G.dy = 0;
@@
                     break;
-                <...> // similar to DOWN/LEFT/RIGHT/SPACE
+                <...> // same for other keys
             }
@@
 void cb_eff (void) {
-    // clears the SDL window
+    <...> // clears the SDL window
     // redraws the rectangle at the current position

### 2022-08-25 tpd-22.tex
...

@@
 Nodes are allowed to join and leave the network at any time.
-Application developers must adhere to a restricted API, which is
-deterministic, stateless, and only supports pre-allocated memory.
+Application developers must adhere to a restricted API, which is deterministic,
+stateless, and only supports pre-allocated memory.
 %
@@
 remaining events up to the real time.
+As we detail further, rolling back a peer requires to simulate its whole
+execution from the beginning (or from a snapshot in between).
 %
@@
         cb_ren,         // rendering callback
-        cb_evt          // event triggering callback
+        cb_evt          // events callback
     );
@@
 exactly the same.
+
+\subsubsection{Application State}
+
 The global variable \code{G} must hold the full application state.
@@
 }
+
+\subsubsection{Application Events}
+
+The application generates events that need to be broadcast to the network so
+that all peers behave the same.
+In this sense, application events may differ from low-level local system
+events, which may or may not be mapped 1-to-1 in the callback \code{cb\_evt}.
+
+There middleware predefines two events for all applications:
+    \code{P2P\_EVT\_START} represents the beginning of the simulation, and
+    \code{P2P\_EVT\_TICK} called at every frame.
+The application-specific events must be declared by the programmer in an
+enumeration.
+In our application, we define a new event \code{P2P\_EVT\_KEY} to represent key
+presses:
+
+{\footnotesize
+\begin{verbatim}
+enum {
+    P2P_EVT_KEY = P2P_EVT_NEXT
+};
+\end{verbatim}
+}
+
+\subsubsection{Initialization Callback}
 
@@
 
-The middleware calls \code{cb\_sim} on every frame to affect the simulation
-state of \code{G} deterministically.
+\subsubsection{Simulation Callback}
+
+The middleware calls \code{cb\_sim} on every frame, passing an application
+event, to affect the simulation state \code{G} deterministically.
 The callback must never perform any side effects, such as stateful calls or
-rendering the frame:
-The callback receives the occurring input event if any.
-There are two predetermined events:
-    \code{P2P\_EVT\_INIT} called at the beginning of the simulation, and
-    \code{P2P\_EVT\_TICK} called at every frame.
-The application-specific events must be declared by the programmer in an
-enumeration.
-In our application, we define a new event \code{P2P\_EVT\_KEY} to represent key
-presses:
-
-{\footnotesize
-\begin{verbatim}
-enum {
-    P2P_EVT_KEY = P2P_EVT_NEXT
-};
-\end{verbatim}
-}
-
-Our callback \code{cb\_sim}
-    (a) resets the rectangle position and speed on \code{INIT},
-    (b) increments the position based on the current speed on \code{TICK}, and
-    (c) modifies the speed based on \code{KEY}:
+rendering frames.
+In our example, we modify the state of the rectangle as follows:
+    (a) reset its position and speed on \code{START},
+    (b) increment its position based on the current speed on \code{TICK}, and
+    (c) modify its speed based on \code{KEY}:
 
@@
     switch (evt.id) {
-        case P2P_EVT_INIT: // resets pos and speed
+        case P2P_EVT_START: // resets pos and speed
             G.x = G.y = G.dx = G.dy = 0;
@@
 time.
-This is the reason why we define \code{INIT} as an event: time travelling
-requires to simulate, since the
+This is the reason why we define \code{START} as an event like any other, since
+travelling requires to simulate the execution from the beginning, event by
+event.
 
-This is the reason why \code{cb\_sim} should be split from the rendering
-callback \code{cb\_ren}~\cite{tml.js}.
+\subsubsection{Rendering Callback}
+
+Time travelling is also the reason why \code{cb\_sim} must be split from the
+rendering callback \code{cb\_ren}~\cite{tml.js}, which would otherwise render
+the screen multiple times while travelling.
 %, which the middleware calls only on real-time frames.
-In our example, this callback just needs to redraw the rectangle at the
+In our example, \code{cb\_ren} just needs to redraw the rectangle at the
 current position:
@@
 
-Finally, the \code{cb\_evt} callback is called by the middleware in real time,
-whenever a local SDL event occurs.
+\subsubsection{Events Callback}
+
+The \code{cb\_evt} callback is called by the middleware in real time, whenever
+a local SDL event occurs.
 This callback has two goals:
-    (a) broadcast application events, and
+    (a) map and broadcast application events, and
     (b) provide instant feedback to the user.
-Not all local low-level events may generate application events to be broadcast
-in the network.
+As mentioned previously, not all local low-level events may generate
+application events to be broadcast in the network.
 The callback is allowed to perform side effects, such as modifying the network

### 2022-08-27 tpd-22.tex
...

@@
 
-\begin{figure*}[t]
+\begin{figure*}[p]
   \centering
@@
 API in which developers have to conform with.
+%
 The middleware is written in C and relies on the library \emph{SDL}%
@@
 the same position.
+
+\begin{figure}[t]
+  \centering
+  \includegraphics[width=\linewidth]{middleware}
+  \caption{
+    Applications generate asynchronous events and communicate through an API
+    with the middleware.
+    The middleware orchestrates the peers, synchronizing their ticks locally
+    and broadcasting events with a delay.
+    \label{fig.middleware}
+  }
+\end{figure}
+
+In Figure~\ref{fig.middleware}, a peer is depicted as an application that uses
+an API to communicate with the middleware.
+The application itself is not aware of the network and has no direct access to
+the other peers, which communicate transparently through the middleware.
+The application generates asynchronous events that go through the middleware,
+which synchronizes them with a timestamp in the future, such that all peers can
+agree.
+The middleware controls the execution of the application by issuing ticks and
+synchronized events.
 
@@
 } G;
-int main (void) { ... }
 \end{verbatim}
@@
             SDL_DrawImage(<img-of-the-key>);
-            *evt = (p2p_evt) { P2P_EVT_KEY, sdl->key };
+            *evt = (p2p_evt){ P2P_EVT_KEY,sdl->key };
             return 1; // generate application event
@@
     }
-    return 0; // otherwise, does not generate any event
+    return 0; // otherwise, do not generate any event
 }
@@
 
-\subsection{The Time Machine}
-\label{sec.tml.time}
+\subsection{Middleware Orchestration}
+\label{sec.tml.middleware}
+
+We now describe in detail how the middleware orchestrates an application and
+ensures that all peers (a) meet event deadlines, (b) advance in sync in real
+time, and (c) can leave and join the network and remain symmetric.
 

### 2022-08-28 tpd-22.tex
...

@@
   \caption{
+    Related work regarding
+        (a) how the network is organized,
+        (b) how global time is determined, and
+        (c) how events are propagated and applied.
     \label{fig.table}
@@
 applications.
-We focus on (a) how the network is organized, (b) how global time is
-determined, and (c) how events are propagated and applied.
+We focus on
+    (a) how the network is organized,
+    (b) how global time is determined, and
+    (c) how events are propagated and applied.
 Figure~\ref{fig.table} compares selected works regarding these aspects.
@@
 
-The middleware assumes that each peer is assigned a unique identifier in a
-contiguous range.
+The middleware assumes that each peer assigns to itself a unique identifier in
+a contiguous range.
 The FPS must be the same in all peers to ensure that the simulation executes
@@
 
-\subsection{TODO}
+\subsection{Middleware Orchestration}
+\label{sec.tml.middleware}
 
+We now detail how the middleware orchestrates an application and ensures that
+all peers
+    (a) meet event deadlines,
+    (b) advance in sync in real time, and
+    (c) can leave and join the network and remain symmetric.
+
+\subsubsection{Events Broadcasting}
+\label{sec.tml.middleware.events}
+
+As illustrated in Figure~\ref{fig.middleware}, events are broadcast between
+peers with a timestamp in the future such that all peers are able to apply them
+in sync.
+%
+To prevent broadcast cycles, event are tagged with a source peer and sequence
+number.
+Each peer keeps a vector with the highest sequence number it received from each
+other peer, such that lower numbers can be ignored when received.
+Higher numbers update the vector and trigger a broadcast to neighbours.
+Since we rely on TCP connections, it is not possible to receive out-of-order
+sequence numbers from the same source peer.
+%
+As mentioned in Section~\ref{sec.tml.api}, we assume that peers assign
+themselves unique identifiers in a contiguous range.
+The reasons are
+    (a) to fit a \code{uint8\_t} size, and
+    (b) to use a contiguous vector of sequence numbers.
+Therefore, this is not a serious limitation, and a separate mechanism could be
+used for assignment, with few modifications to core of the middleware.
+
+When receiving an unseen event, the middleware also enqueues it in timestamp
+order and calls the application callback \code{cb\_sim} at the appropriate
+time.
+The peers must keep the full queue in memory for two reasons:
+    (a) receiving a late event may reorder the queue and require a time
+        travel, and
+    (b) peers that join the network need to receive all events.
+%
+Therefore, the middleware keeps a queue of ``packets'', which includes the
+event and other metadata:
+
+{\footnotesize
+\begin{verbatim}
+typedef struct {
+    uint8_t  src;       // source peer
+    uint32_t seq;       // sequence number
+    uint32_t tick;      // tick to apply
+    p2p_evt  evt;       // event id + payload
+} p2p_pak;
+
+struct {
+    int i;              // next event to apply
+    int n;              // number of events
+    p2p_pak buf[MAX];   // queue of packets
+} PAKS;
+\end{verbatim}
+}
+
+\subsubsection{The Main Loop}
+\label{sec.tml.middleware.loop}
+
+The middleware main loop \code{p2p\_loop} is responsible for calling the
+application callbacks, and also controlling its timeline.
+To achieve this, the loop continuously checks for network packets, local
+inputs, and time ticks, as follows:
+
+{\footnotesize
+\begin{verbatim}
+01  void p2p_loop (...,cb_ini,cb_sim,cb_ren,cb_evt) {
+01    cb_ini(1);
+01    cb_sim(<P2P_EVT_INIT>);
+01
+01    while (<app-running>) {
+01      if (<next-packet>) {
+01        // fast forward if too late
+01        if (<packet-sync-time-future>) {
+01          for (<step-by-step-sleep>) {
+01            p2p_travel(<packet-sync-time>);
+01            cb_ren();
+01          }
+01        }
+01
+01        if (<packet-evt>) {
+01          if (<packet-evt-time-past>) {
+01            for (<step-by-step-sleep>) {
+01              p2p_travel(<packet-evt-time>);
+01              cb_ren();
+01            }
+01            for (<step-by-step-sleep>) {
+01              p2p_travel(<now>);
+01              cb_ren();
+01            }
+01          } else {
+01            cb_sim(<packet-evt>);
+01            cb_eff();
+01          }
+01        }
+01      }
+01
+01      if (<next-input>) {
+01        p2p_evt evt;
+01        if (cb_rec(&evt)) {
+01          p2p_bcast(<future>, &evt);
+01        }
+01      }
+01
+01      if (<next-tick>) {
+01        cb_sim(<P2P_EVT_TICK>);
+01        <take-snapshot-every-XXXms>
+01        cb_ren();
+01      }
+01    }
+01
+01    cb_ini(0);
+01  }
+\end{verbatim}
+}
+
+- async events
+- sync events
+- real-time sim
+- sync time
+
+advance time locally
+
+TODO: no final falar de a,b,c
 TODO: simpler than vector clocks, absolute times are correct
@@
 
-\subsection{Middleware Orchestration}
-\label{sec.tml.middleware}
-
-We now describe in detail how the middleware orchestrates an application and
-ensures that all peers (a) meet event deadlines, (b) advance in sync in real
-time, and (c) can leave and join the network and remain symmetric.
-
 - p2p links por fora, como um servico TCP local
 - The network topology is dynamic
-
-%\item Peers have contiguous numeric unique identifiers.
 
@@
 \begin{itemize}
-\item Peers have contiguous numeric unique identifiers.
 \item Peers identifiers are assigned externally to the middleware.
@@
 solved further:
-- peers have unique identifiers
 - ids are outside

### 2022-08-28 tpd-22.tex
...

@@
 \newcommand{\Xo}  {$1{\hookleftarrow}$\xspace}
+\newcommand{\lin}[1]{(\emph{ln. #1}\xspace)}
 
@@
 \subsubsection{Application State}
+\label{sec.tml.api.state}
 
@@
 \subsubsection{Application Events}
+\label{sec.tml.api.events}
 
@@
 \subsubsection{Initialization Callback}
+\label{sec.tml.api.cb_ini}
 
@@
 \subsubsection{Simulation Callback}
+\label{sec.tml.api.cb_sim}
 
@@
 application callbacks, and also controlling its timeline.
-To achieve this, the loop continuously checks for network packets, local
-inputs, and time ticks, as follows:
+The loop continuously checks for network packets, local inputs, and time ticks,
+as follows:
 
@@
 01  void p2p_loop (...,cb_ini,cb_sim,cb_ren,cb_evt) {
-01    cb_ini(1);
-01    cb_sim(<P2P_EVT_INIT>);
-01
-01    while (<app-running>) {
-01      if (<next-packet>) {
-01        // fast forward if too late
-01        if (<packet-sync-time-future>) {
-01          for (<step-by-step-sleep>) {
-01            p2p_travel(<packet-sync-time>);
-01            cb_ren();
-01          }
-01        }
-01
-01        if (<packet-evt>) {
-01          if (<packet-evt-time-past>) {
-01            for (<step-by-step-sleep>) {
-01              p2p_travel(<packet-evt-time>);
-01              cb_ren();
-01            }
-01            for (<step-by-step-sleep>) {
-01              p2p_travel(<now>);
-01              cb_ren();
-01            }
-01          } else {
-01            cb_sim(<packet-evt>);
-01            cb_eff();
-01          }
-01        }
-01      }
-01
-01      if (<next-input>) {
-01        p2p_evt evt;
-01        if (cb_rec(&evt)) {
-01          p2p_bcast(<future>, &evt);
-01        }
-01      }
-01
-01      if (<next-tick>) {
-01        cb_sim(<P2P_EVT_TICK>);
-01        <take-snapshot-every-XXXms>
-01        cb_ren();
-01      }
-01    }
-01
-01    cb_ini(0);
-01  }
+02    cb_ini(1);
+03    cb_sim(<P2P_EVT_START>);
+04
+05    while (<app-running>) {
+06      if (<next-packet>) {
+07        // fast forward if I'm late
+08        if (<packet-sync-future>) {
+09          p2p_travel(<now>, <packet-sync-time>);
+10        }
+11
+12        // travel back & forth if I'm early
+13        if (<packet-evt-past>) {
+14          p2p_travel(<now>, <packet-evt-time>);
+15          p2p_travel(<packet-evt-time>, <now>);
+16        }
+17
+18        // simulate in real time if I'm on time
+19        if (<packet-evt-ok>) {
+20          cb_sim(<packet-evt>);
+21          cb_ren();
+22        }
+23      }
+24
+25      if (<next-input>) {
+26        // broadcast event from local
+27        p2p_evt evt;
+28        if (cb_evt(&evt)) {
+29          p2p_bcast(<future>, &evt);
+30        }
+31      }
+32
+33      // simulate tick in real time
+34      if (<next-tick>) {
+35        cb_sim(<P2P_EVT_TICK>);
+36        <take-snapshot-every-second>
+37        cb_ren();
+38      }
+39    }
+40
+41    cb_ini(0);
+42  }
 \end{verbatim}
 }
+
+The function receives the application callbacks as arguments \lin{1}
+(Section~\ref{sec.tml.api}).
+It first (and last) calls \code{cb\_ini} \lin{2,41} to initialize (and
+finalize) the application.
+Then, it calls \code{cb\_sim} \lin{3}, passing the event \code{START} to start
+the application in real time
+(Sections~\ref{sec.tml.api.events}~and~\ref{sec.tml.api.cb_sim}).
+The main loop \lin{5--39} checks continuously for network packets \lin{6--23},
+local input events \lin{25--31}, and time ticks \lin{33--38}.
+%
+If we receive a time synchronization packet \code{SYNC} from the future
+\lin{7--10}, then the we fast forward the application to catch up in time
+(Section~\ref{sec.tml}.b).
+If we receive an event that should have been applied in the past \lin{12--16},
+then we travel back and forth (Section~\ref{sec.tml}.a).
+Otherwise, we just apply the event in real time \lin{18--22}
+(Section~\ref{sec.tml.api.cb_sim}).
+
+We detail the function \code{p2p\_travel} further
+
+\subsubsection{The Time Machine}
+\label{sec.tml.middleware.travel}
 

### 2022-08-30 tpd-22.tex
...

@@
 in time.
+Our main contribution is the design of a middleware that runs on each peer to
+ensure that all peers
+    (a) meet event deadlines,
+    (b) advance in sync in real time, and
+    (c) can leave and join the network and remain symmetric.
 
@@
 
+\begin{figure}[t]
+    \centering
+    \includegraphics[width=\linewidth]{move}
+    \caption[XXX] {
+        A moving rectangle is controlled by peers connected indirectly.
+        Users can press \code{↑ ↓ → ←} to change the direction of the
+        rectangle or \code{SPACE} to pause the application.
+        The inputs are applied simultaneously in all peers, as if they were
+        mirrors of a single application.
+        Video: \url{http://youtube.com/TODO}
+        \label{fig.move}
+    }
+\end{figure}
+
 We guide our discussion through the didactic example of Figure~\ref{fig.move},
@@
 network in order to preserve the symmetric behavior across peers.
-More specifically, the middleware ensures that all peers
+As described in the Introduction, the middleware ensures that all peers
     (a) meet event deadlines,
@@
 The global variable \code{G} must hold the full application state.
+This allows the middleware to take memory snapshots and rollback to a previous
+state.
 If dynamic allocation is required, it is necessary to hold a finite memory
@@
 \subsubsection{Rendering Callback}
+\label{sec.tml.api.cb_ren}
 
@@
 \subsubsection{Events Callback}
+\label{sec.tml.api.cb_evt}
 
@@
 35        cb_sim(<P2P_EVT_TICK>);
-36        <take-snapshot-every-second>
+36        <application-snapshot-every-second>
 37        cb_ren();
@@
 (Section~\ref{sec.tml.api}).
-It first (and last) calls \code{cb\_ini} \lin{2,41} to initialize (and
-finalize) the application.
-Then, it calls \code{cb\_sim} \lin{3}, passing the event \code{START} to start
+We first (and last) call \code{cb\_ini} \lin{2,41} to initialize (and finalize)
+the application.
+Then, we call \code{cb\_sim} \lin{3}, passing the event \code{START} to start
 the application in real time
@@
 local input events \lin{25--31}, and time ticks \lin{33--38}.
-%
-If we receive a time synchronization packet \code{SYNC} from the future
-\lin{7--10}, then the we fast forward the application to catch up in time
-(Section~\ref{sec.tml}.b).
+
+Regarding network packets, if we receive a time synchronization packet
+\code{SYNC} from the future \lin{7--10}, then the we fast forward the
+application to catch up in time (Section~\ref{sec.tml}.b).
 If we receive an event that should have been applied in the past \lin{12--16},
@@
 (Section~\ref{sec.tml.api.cb_sim}).
-
-We detail the function \code{p2p\_travel} further
+We detail the function \code{p2p\_travel} in
+Section~\ref{sec.tml.middleware.travel}.
+%
+Regarding local inputs, we call \code{cb\_evt} \lin{26--30} to signal if the
+event should be broadcast, also transforming it to an application event
+(Section~\ref{sec.tml.api.cb_evt}).
+%
+Regarding time tick, we call \code{cb\_sim} and \code{cb\_ren} in real time
+\lin{33--38}.
+The middleware also takes periodic snapshots of the application state to
+rollback and assist time travelling (Section~\ref{sec.tml.api.state}).
 
@@
 
-- async events
-- sync events
-- real-time sim
-- sync time
+The last important mechanism of the middleware is the time machine, which
+allows to move the application back and forth in time:
 
-advance time locally
+{\footnotesize
+\begin{verbatim}
+01  void p2p_travel (int from, int to) {
+02    for (i=from..to) {
+03      int bef = <tick-of-snapshot-before-i>
+04      <restore-snapshot-at-bef>
+05      for (j=bef..i) {
+06        cb_sim(<tick-or-evt-at-j>)
+07      }
+08      cb_ren();
+09      <delay-ms>
+10    }
+11  }
+\end{verbatim}
+}
 
-TODO: no final falar de a,b,c
+The function \code{p2p\_travel} time travels, starting at tick \code{from},
+tick by tick, until it reaches tick \code{to} \lin{2--10}.
+The loop can go in any direction in time, i.e., \code{from} may be higher than
+\code{to}.
+At each step, we need to find the tick \code{bef} with the closest past
+snapshot \lin{3}, restore it \lin{4}, and than simulate the remaining steps
+from the snapshot to the desired tick \lin{5--7}.
+We also exhibit each step with a small delay \lin{8--9} so that users
+experience smooth time transitions.
+
+\subsubsection{Summary}
+\label{sec.tml.middleware.summary}
+
+This section provided an overview of the middleware implementation to answer,
+in concrete terms, how it ensures the desired properties enumerated in the
+Introduction and at the beginning of Section~\ref{sec.tml.middleware}:
+    (a) peers meet event deadlines either in real time \lin{18--21} or
+        travelling back and forth in time \lin{12--16}
+        (Section~\ref{sec.tml.middleware.loop});
+    (b) peers synchronize their clocks with periodic \code{SYNC} packets that
+        fast forwards late peers \lin{7--10}
+        (Section~\ref{sec.tml.middleware.loop}); and
+    (c) peers can join the network at any time and still catch up in time by
+        receiving the full event history
+        (Section~\ref{sec.tml.middleware.events}).
+
+\section{Evaluation}
+\label{sec.eval}
+
+In this section, we evaluate if our design meets its goals to ensure that all
+peers (a) meet event deadlines, (b) advance in sync in real time, and (c) can
+leave and join the network and remain symmetric.
+%
+For each of these goals, we simulate different network scenarios varying the
+(i) topology, (j) latency, (k) churn, and (k) frequency and (m) deadlines of
+events.
+For each scenario, we measure (x) the recurrence of time travels, (y) the
+real-time pace of peers, and (z) the correctness of unstable peers, such that
+they validate goals (a), (b), and (c), respectively.
+
+- VARIABLES
+- number of nodes
+- topology
+    - distance b/w nodes
+    - number of edges
+- latency b/w nodes
+- freq of events
+- ~FPS~
+- future delta
+
+- RESULTS
+- time travels
+- ~real time~
+- ~enter leave~ (catches up?)
+
+\section{TODO}
+
 TODO: simpler than vector clocks, absolute times are correct
-TODO: piggyback SYNC in normal inputs
 
 - p2p links por fora, como um servico TCP local
-- The network topology is dynamic
-
-%
-- time sync
-
-The middleware handles:
-- time sync
-- disc / conn
-- late event
-
-\begin{figure}[t]
-    \centering
-    \includegraphics[width=\linewidth]{move}
-    \caption[XXX] {
-        A moving rectangle is controlled by peers connected indirectly.
-        Users can press \code{↑ ↓ → ←} to change the direction of the
-        rectangle or \code{SPACE} to pause the application.
-        The inputs are applied simultaneously in all peers, as if they were
-        mirrors of a single application.
-        Video: \url{http://youtube.com/TODO}
-        \label{fig.move}
-    }
-\end{figure}
-
-- walk back-forth
-- vector clocks
-- dynamic net
-- re-connect
-- late join
 
@@
 
-\section{Evaluation}
-\label{sec.eval}
-
 \section{Conclusion}

### 2022-09-04 tpd-22.tex
...

@@
 
-In this section, we evaluate if our design meets its goals to ensure that all
+In this section, we simulate an application varying the network topology,
+latency, and churn, as well as the frequency and deadlines of events.
+We use a single machine .
+We then evaluate if our design meets its goals as follows:
+%
+\begin{enumerate}
+\item \textbf{Event Deadlines:}
+We measure the recurrence of time travels among peers.
+If a peer experiences a time travel, it means that it received an event after
+its deadline.
+For instance, that would be the case when the source peer generates an event
+with a small deadline in comparison to the latency to the farthest destination
+peer.
+More concretely, we teak the middleware implementation to count time travel
+ticks.
+We then measure the percentage of the sum of all time travels ticks over the to
+the total simulation ticks (e.g., $1\%$, if all peers sum $1000$ time travel
+ticks over $100k$ accumulated simulation ticks).
+%
+\item \textbf{Time Synchronization:}
+We measure if peers advance in sync in real time using as a reference the peer
+most advanced in time.
+We start each peer with a random delay to force a time mismatch, and then
+verify if they synchronize and remain synchronized during the whole simulation.
+More concretely, we teak the middleware to .
+
+More concretely, we measure the percentage o ticks diverging from the reference
+peer (e.g., $1\%$, if .
+
+converge
+
+, and (c) can
+Since peers run in different machines, their timelines will inevitably be out
+of sync because:
+    (a) applications are launched locally at different times, and
+    (b) internal clocks may diverge over time.
+The middleware assumes the maximum time among all peers the be the
+``correct real time'', so that other peers must fast forward to catch up with
+it.
+Therefore, in order to determine the maximum time and synchronize the clocks,
+the middleware proceeds as follows:
+    (a) peers broadcast \code{SYNC} events periodically with their local
+        times, and
+    (b) if a peer is behind a received \code{SYNC}, it advances its frames
+        proportionally faster to catch up in time.
+%
+\item \textbf{Peer Churn:}
+Considering that we target peer-to-peer networks, it is important that the
+middleware can handle arrival and departure of peers seamlessly.
+When a peer leaves the network, nothing needs to be done.
+The middleware just needs to ensure that a peer that joins the network receives
+all events in order, which is trivial since they all hold an absolute
+timestamp.
+\end{enumerate}
+
+ to ensure that all
 peers (a) meet event deadlines, (b) advance in sync in real time, and (c) can
@@
 %
+For each of these goals, we measure (a) the recurrence of time travels, (b) the
+real-time pace of peers, and (c) the correctness of unstable peers.
+
+, respectively.
+
+, such that
+they validate goals (a), (b), and (c), respectively.
+
 For each of these goals, we simulate different network scenarios varying the
@@
 they validate goals (a), (b), and (c), respectively.
+
+- start with random deltas
 
@@
 - ~enter leave~ (catches up?)
+
+- use of space for snapshots
+    - limit to go back?
+    - super nodes? (when connecting, pass tick, may close if tick<last)
 

### 2022-09-04 tpd-22.tex
...

--- 2022-08-30 tpd-22.tex
@@
 %
-For each of these goals, we simulate different network scenarios varying the
-(i) topology, (j) latency, (k) churn, and (k) frequency and (m) deadlines of
-events.
-For each scenario, we measure (x) the recurrence of time travels, (y) the
-real-time pace of peers, and (z) the correctness of unstable peers, such that
-they validate goals (a), (b), and (c), respectively.
+We simulate an application varying the network topology, latency, and churn, as
+well as the frequency and deadlines of events.
+%
+We then measure (a) the recurrence of time travels, (b) the real-time pace of
+peers, and (c) the correctness of unstable peers, such that they validate our
+goals above, respectively.
+
+Regarding (a), if a peer experiences a time travel, it means that it received
+an event after its deadline.
+We can count all time travel occurrences during the simulation.
+%
+Regarding (b), we can use an absolute simulation clock to track, in real time,
+the pace of each peer in comparison to a reference peer most advanced in time.
+%
+Regarding (c), we can TODO specific peers and very that they TODO within a
+short period of time.
+
+For instance, if the reference peer signals $T$ and then a peer signals $T-dt$,
+than we.
+That would be the case when a source peer generates an event with a small
+deadline in comparison to the latency to the farthest destination peer.
+
+\begin{figure*}[t]
+  \centering
+  \includegraphics[width=\linewidth]{topo}
+  \caption{
+    \label{fig.topo}
+    TODO
+  }
+\end{figure*}
+
+We simulate the application with $21$ peers using the mixed topology of
+Figure~\ref{fig.topo}, with straight lines, long cycles, and short cycles.
+%
+In order to have an absolute clock to compare peers, we use a single machine
+to simulate all peers, and output their logs to a single file.
+We TODO the network latency and clock skew using TODO and TODO.
+
+We now detail why we measure the properties above and how we instrument the
+middleware implementation:
+%
+\begin{enumerate}
+%
+\item \textbf{Recurrence of time travels:}
+We teak the middleware implementation to count time travel ticks.
+We then measure the percentage of the sum of all time travels ticks over the to
+the total simulation ticks (e.g., $1\%$, if all peers sum $1000$ time travel
+ticks over $100k$ accumulated simulation ticks).
+%
+\item \textbf{Real-time pace of peers:}
+We start each peer with a random delay of up to $10s$ to force a time mismatch,
+and then verify if they synchronize and remain synchronized during the whole
+simulation.
+We teak the middleware to output its current tick every $100ms$.
+
+More concretely, we measure the percentage o ticks diverging from the reference
+peer (e.g., $1\%$, if .
+
+%
+\end{enumerate}
+
+We run the simulation for 5 minutes and, since we start the measures after 1 minute
 
@@
 
+-- use of space for snapshots
+-    - limit to go back?
+-    - super nodes? (when connecting, pass tick, may close if tick<last)
+
+-- start with random deltas
 \section{TODO}

### 2022-09-05 tpd-22.tex
...

@@
 
-In this section, we evaluate if our design meets its goals to ensure that all
-peers (a) meet event deadlines, (b) advance in sync in real time, and (c) can
-leave and join the network and remain symmetric.
+In this section, we evaluate if (and to what extent) our design meets its goals
+to ensure that all peers (a) meet event deadlines, (b) advance in sync in real
+time, and (c) can leave and join the network and remain symmetric.
 %
@@
 an event after its deadline.
-We can count all time travel occurrences during the simulation.
+We count all time travel occurrences during the simulation.
 %
-Regarding (b), we can use an absolute simulation clock to track, in real time,
-the pace of each peer in comparison to a reference peer most advanced in time.
+Regarding (b), we use an absolute simulation clock to track, in real time, the
+execution pace of each peer in comparison to a reference peer most advanced in
+time.
 %
-Regarding (c), we can TODO specific peers and very that they TODO within a
-short period of time.
-
-For instance, if the reference peer signals $T$ and then a peer signals $T-dt$,
-than we.
-That would be the case when a source peer generates an event with a small
-deadline in comparison to the latency to the farthest destination peer.
+Regarding (c), we designate some peers to leave and join the network
+periodically to assert that they catch up with the other peers in real time.
 
@@
 We simulate the application with $21$ peers using the mixed topology of
-Figure~\ref{fig.topo}, with straight lines, long cycles, and short cycles.
+Figure~\ref{fig.topo}, with peers organized in straight lines, long cycles, and
+short cycles.
 %
 In order to have an absolute clock to compare peers, we use a single machine
-to simulate all peers, and output their logs to a single file.
-We TODO the network latency and clock skew using TODO and TODO.
+to simulate all peers, and merge their logs into a single file constituting a
+shared timeline.
+%
+We simulate the network latency and clock drift in software with random delays.
 
-We now detail why we measure the properties above and how we instrument the
-middleware implementation:
+We now detail how we instrument the middleware implementation measure the
+properties above:
 %
@@
 \item \textbf{Recurrence of time travels:}
-We teak the middleware implementation to count time travel ticks.
-We then measure the percentage of the sum of all time travels ticks over the to
-the total simulation ticks (e.g., $1\%$, if all peers sum $1000$ time travel
-ticks over $100k$ accumulated simulation ticks).
+Every time a peer is ahead of time and needs to travel back, we output its id
+and how much ticks it needs to travel back.
+At the end, we measure the percentage of the sum of all time travel ticks over
+the total simulation ticks.
+As an example, if all peers sum $1000$ time travel ticks over $100k$
+accumulated simulation ticks, then the final measure is $1\%$.
 %
 \item \textbf{Real-time pace of peers:}
-We start each peer with a random delay of up to $10s$ to force a time mismatch,
+We start each peer with a random delay of up to $10s$ to force a time mismatch.
+We also force exaggerated clock drifts in random peers ($1ms$ every $1h$).
 and then verify if they synchronize and remain synchronized during the whole
@@
 \end{enumerate}
+
+For instance, if the reference peer signals $T$ and then a peer signals $T-dt$,
+than we.
+That would be the case when a source peer generates an event with a small
+deadline in comparison to the latency to the farthest destination peer.
 

### 2022-09-10 tpd-22.tex
...

@@
     \label{fig.topo}
-    TODO
+    Network topology: 21 peers organized as a straight line ($0--3$), a simple
+    cycle ($3--8$), a second straight line ($8--11$), a full connected cycle
+    ($12--17$), and a third straight line ($17--20$).
+    An optional link ($0,20$) creates a long cycle in the network.
   }
@@
 We simulate the application with $21$ peers using the mixed topology of
-Figure~\ref{fig.topo}, with peers organized in straight lines, long cycles, and
-short cycles.
+Figure~\ref{fig.topo}, with peers organized in straight lines and cycles.
 %
-In order to have an absolute clock to compare peers, we use a single machine
-to simulate all peers, and merge their logs into a single file constituting a
+In order to have an absolute clock, we use a single machine to simulate all
+peers, allowing us to merge their logs into a single file with a comparable
 shared timeline.
 %
-We simulate the network latency and clock drift in software with random delays.
+We simulate the network latency in software with random delays.
+
+Figure~\ref{TODO} results...
 
@@
 We start each peer with a random delay of up to $10s$ to force a time mismatch.
-We also force exaggerated clock drifts in random peers ($1ms$ every $1h$).
+Each peer outputs its current tick continuously, allowing us to compare them
+in the shared timeline.
+Each time we see a greater tick, we count how many smaller ticks from other
+peers appear in the timeline, which means that they are ...
+\item \textbf{Correctness of unstable peers:}
+- state (even for stable peers)
+- time
+%
+\end{enumerate}
+
+After $Xs$, .
 and then verify if they synchronize and remain synchronized during the whole
@@
 peer (e.g., $1\%$, if .
-
-%
-\end{enumerate}
 

### 2022-09-11 tpd-22.tex
...

@@
 do not rely on central servers for coordination.
-Nodes in the application form a dynamic graph network and communicate only
-with direct neighbours.
+Peers in the application form a dynamic graph network and communicate only
+with direct neighbours, as a typical unstructured peer-to-peer network.
 Events are flooded in the graph and are triggered locally with a small delay
@@
     (a) applications are launched locally at different times, and
-    (b) internal clocks may diverge over time.
+    (b) internal clocks may diverge over time (e.g., from rollbacks or timer
+        polling inaccuracies).
 The middleware assumes the maximum time among all peers the be the
@@
 middleware can handle arrival and departure of peers seamlessly.
-When a peer leaves the network, nothing needs to be done.
-The middleware just needs to ensure that a peer that joins the network receives
-all events in order, which is trivial since they all hold an absolute
-timestamp.
+Note that high churn may lead to intermittent network partitions.
+Given our unstructured approach, nothing needs to be done when a peer leaves
+the network.
+However, when a peer joins the network, the middleware just needs to ensure
+that it receives all events in order, which is trivial since they all hold an
+absolute timestamp.
 \end{enumerate}
@@
 Figure~\ref{TODO} results...
-
-We now detail how we instrument the middleware implementation measure the
+- colors indicate feasibility
+%
+We now detail how we instrument the middleware implementation to measure the
 properties above:
@@
 At the end, we measure the percentage of the sum of all time travel ticks over
-the total simulation ticks.
+the total simulation ticks (\code{TODO} in Figure~\ref{TODO}).
 As an example, if all peers sum $1000$ time travel ticks over $100k$
@@
 \item \textbf{Real-time pace of peers:}
-We start each peer with a random delay of up to $10s$ to force a time mismatch.
+Every time a peer receives \code{SYNC} packets and is behind, we output its id
+and how much ticks it needs to fast forward.
+At the end, we measure the percentage of the sum of all forward ticks over the
+total simulation ticks (\code{TODO} in Figure~\ref{TODO}).
+As an example, if all peers sum $1000$ forward ticks over $100k$ accumulated
+simulation ticks, then the final measure is $1\%$.
+%
+We also start each peer with a random delay of up to $10s$ to force a time
+mismatch.
+%
 Each peer outputs its current tick continuously, allowing us to compare them
 in the shared timeline.
-Each time we see a greater tick, we count how many smaller ticks from other
-peers appear in the timeline, which means that they are ...
+Each time we see a greater tick never seen in the timeline before, we count
+smaller ticks from other peers (considering latency), which would constitute
+time violations (\code{TODO} in Figure~\ref{TODO}).
+We expect this counter to be close to zero considering the whole execution.
 \item \textbf{Correctness of unstable peers:}
@@
 We run the simulation for 5 minutes and, since we start the measures after 1 minute
+
+- check if all ticks of all peers are in the same state
 

### 2022-09-13 tpd-22.tex
...

@@
 We then measure (a) the recurrence of time travels, (b) the real-time pace of
-peers, and (c) the correctness of unstable peers, such that they validate our
+peers, and (c) the recoverability from churns, such that they validate our
 goals above, respectively.
@@
 Regarding (c), we designate some peers to leave and join the network
-periodically to assert that they catch up with the other peers in real time.
+periodically, possibly creating partitions, to assert that they all catch up in
+real time.
 
@@
 time violations (\code{TODO} in Figure~\ref{TODO}).
-We expect this counter to be close to zero considering the whole execution.
+We expect this counter to be close to zero even considering the whole
+simulation.
+%
 \item \textbf{Correctness of unstable peers:}
@@
 - time
+- remove 13 (nao partition) e 10 (partition)
+- qto tempo leva para sync
 %

### 2022-10-01 tpd-22.tex
- ate antes de 3.2

@@
 \begin{abstract}
-In multi-node collaborative applications, such as shared online documents,
+In collaborative networked applications, such as shared online documents,
 users can interact remotely and yet share the exact same experience as if they
@@
 %
-Nodes are allowed to join and leave the network at any time.
+Peers are allowed to join and leave the network at any time.
 Application developers must adhere to a restricted API, which is deterministic,
@@
 timeline across the network.
-Based on memory snapshots and deterministic simulation, a ``time machine'' can
-rollback conflicting peers to resynchronize their state.
+Also, ased on memory snapshots and deterministic simulation, a ``time machine''
+can rollback conflicting peers to resynchronize their state.
 %
@@
 do not rely on central servers for coordination.
-Peers in the application form a dynamic graph network and communicate only
-with direct neighbours, as a typical unstructured peer-to-peer network.
+Peers in the application form a dynamic network graph and communicate only
+with direct neighbours, as in typical unstructured peer-to-peer networks.
 Events are flooded in the graph and are triggered locally with a small delay
-to consider the network latency.
+to compensate the network latency.
 To deal with events received out of order or too late, a distributed time
@@
 to a synchronized global shared clock.
-If an user generates an event, it is delayed and sent back to the reflector,
-which broadcasts the event in the next tick, which all nodes apply in sync
-(including the originating node).
+If a user generates an event, it is delayed and sent back to the reflector,
+which broadcasts the event in the next tick, which all nodes apply in sync.
 %
@@
 As detailed in Figure~\ref{fig.algos}, instead of advancing ticks in sync with
-the server, clients have their own local clocks.
-Also, event generation is delayed dynamically according to the slowest client.
+the server, clients have their own independent local clocks.
+Event generation requires two round trips (\emph{when→ask→ack→tick}) and is
+delayed dynamically according to the slowest client.
 %
@@
 frame transitions, even those with poor connections.
-On the other hand, event generation requires two round trips to the server and
-clients may experience occasional freezes.
+On the other hand, events are slower and clients may experience occasional
+freezes.
 The syncing protocol also requires bookkeeping to deal with clock drifts and
@@
 %
-As indicated in Figure~\ref{fig.table}, GALS relies on a centralized
-network, in which nodes advance independently, and in which event outcomes
-depend on the central server to be online.
+As indicated in Figure~\ref{fig.table}, GALS also relies on a centralized
+network, but in which nodes advance time independently, and in which event
+outcomes still depend on the central server to be online.
 
@@
 using only deterministic and stateless calls.
-This way, it is guaranteed that all clients reach the same state, also going
-through bit-identical steps.
+This way, it is guaranteed that all clients reach the same state, and
+observe bit-identical steps.
 
@@
 Since operations must be commutative, it is not possible to timestamp them
-in a global shared clock.
+in a unique global shared clock.
 Hence, there is no notion of a timeline in which peers go through
@@
 %
-On the one hand, the main advantage of CRDTs is the support for local-first
+On the one hand, the main advantage of CRDTs is to support local-first
 software~\cite{local}, since they can work offline in the same way as online.
@@
 \footnote{SDL: \url{www.libsdl.org}}
-to deal with the network and input events, such as timers, mouse clicks and
+to deal with the network, timers, and input events, such as mouse clicks and
 key presses.
@@
 %
-The application is symmetric in the sense that if any peer presses a key, a
+The application is symmetric in the sense that if any user presses a key, a
 corresponding event is broadcast and applied in all peers simultaneously, as
-if they were a single mirrored instance.
-At any moment, if a peer pauses the application, all peers are guaranteed to
+if they were a single mirrored machine.
+At any moment, if a user pauses the application, all peers are guaranteed to
 pause exactly at the same frame with the rectangle being ``bit-identically'' at
@@
 which synchronizes them with a timestamp in the future, such that all peers can
-agree.
+satisfy.
 The middleware controls the execution of the application by issuing ticks and
@@
 Due to the inherent latency of networks, peers generate events with a
-deliberate delay so that they reach other peers in time to be applied in sync
-in a common deadline.
+deliberate delay so that they can reach other peers in time to be applied in
+sync.
 %We first assume a predetermined delay for the sake of simplicity.
-However, a peer that is locally ahead of time may receive an event that should
-have been applied in the past, event considering the delay.
+Nevertheless, a peer that is locally ahead of time may receive an event that
+should have been applied in the past, even considering the delay.
 In this case, the middleware automatically rolls back the peer, applies the
@@
 remaining events up to the real time.
-As we detail further, rolling back a peer requires to simulate its whole
-execution from the beginning (or from a snapshot in between).
+As we detail further, rolling back a peer requires to simulate the whole
+application execution from the beginning.
+As an optimization, the middleware takes periodic snapshots locally to avoid
+full simulation.
 %
@@
 Since peers run in different machines, their timelines will inevitably be out
-of sync because:
-    (a) applications are launched locally at different times, and
-    (b) internal clocks may diverge over time (e.g., from rollbacks or timer
+of sync because applications are launched locally at different times, and
+because internal clocks may diverge over time (e.g., from rollbacks or timer
 polling inaccuracies).
@@
 the network.
-However, when a peer joins the network, the middleware just needs to ensure
-that it receives all events in order, which is trivial since they all hold an
-absolute timestamp.
+However, when a peer joins the network, the middleware needs to ensure that the
+peer receives all events that ever happened in order, which is trivial since
+they all hold an absolute timestamp in the shared timeline.
 \end{enumerate}
@@
 The middleware expects to take full control of the application execution in
-order to control its timeline and disseminate events in the network.
+order to manipulate its timeline and disseminate events in the network.
 For this reason, the basic programming API is to call a \code{loop} function
@@
 
-The middleware assumes that each peer assigns to itself a unique identifier in
-a contiguous range.
-The FPS must be the same in all peers to ensure that the simulation executes
-exactly the same.
+The middleware assumes that each peer assigns itself a unique identifier in a
+contiguous range.
+The FPS must be the same in all peers to ensure bit-identical simulation.
 
@@
 
-The global variable \code{G} must hold the full application state.
-This allows the middleware to take memory snapshots and rollback to a previous
+The variable \code{G} passed to the middleware holds the full application
 state.
+This allows the middleware to take memory snapshots and rollback to previous
+states.
 If dynamic allocation is required, it is necessary to hold a finite memory
 pool in \code{G} with a custom allocator.
-For instance, the moving rectangle application only requires to hold the
-current position and direction:
+In our guiding example of Figure~\ref{fig.move}, the moving rectangle
+application only requires to hold the current position and direction:
 
@@
 that all peers behave the same.
-In this sense, application events may differ from low-level local system
-events, which may or may not be mapped 1-to-1 in the callback \code{cb\_evt}.
-
-There middleware predefines two events for all applications:
+Note that the application events may (or may not) differ from low-level local
+system events.
+%which may or may not be mapped 1-to-1 in the callback \code{cb\_evt}.
+%
+The middleware predefines two events for all applications:
     \code{P2P\_EVT\_START} represents the beginning of the simulation, and
-    \code{P2P\_EVT\_TICK} called at every frame.
+    \code{P2P\_EVT\_TICK} is generated every frame.
 The application-specific events must be declared by the programmer in an
@@
 the loop.
-It should be used to initialize (and finalize) the network topology, as well
-as static globals that can live outside the simulation memory, such as the
-window and image textures:
+The callback initializes (and finalizes) the network topology, as well as
+static immutable globals that can live outside the simulation memory, such as
+the window and image textures in our example:
 
@@
 
-The middleware calls \code{cb\_sim} on every frame, passing an application
-event, to affect the simulation state \code{G} deterministically.
-The callback must never perform any side effects, such as stateful calls or
-rendering frames.
+The middleware calls \code{cb\_sim} every frame, which affects the simulation
+state \code{G} deterministically.
+The occurring event is passed as an argument to the callback, which must never
+perform any side effects, such as stateful calls or rendering frames.
 In our example, we modify the state of the rectangle as follows:
@@
 time.
-This is the reason why we define \code{START} as an event like any other, since
+This is the reason why we define \code{START} as an event like any other:
 travelling requires to simulate the execution from the beginning, event by
@@
 
-Time travelling is also the reason why \code{cb\_sim} must be split from the
-rendering callback \code{cb\_ren}~\cite{tml.js}, which would otherwise render
-the screen multiple times while travelling.
+Time travelling is also the reason why \code{cb\_sim} must be separated from
+the rendering callback \code{cb\_ren}~\cite{tml.js}, which would otherwise
+render the screen multiple times while travelling.
 %, which the middleware calls only on real-time frames.
@@
 \begin{verbatim}
-void cb_eff (void) {
+void cb_ren (void) {
     <...> // clears the SDL window
@@
 
-The \code{cb\_evt} callback is called by the middleware in real time, whenever
+The callback \code{cb\_evt} is called by the middleware in real time, whenever
 a local SDL event occurs.
@@
     (b) provide instant feedback to the user.
-As mentioned previously, not all local low-level events may generate
-application events to be broadcast in the network.
+%As mentioned previously,
+Not all local low-level events need to broadcast application events in the
+network.
 The callback is allowed to perform side effects, such as modifying the network
-topology, or exhibiting some feedback on the screen.
+topology, or exhibiting instant feedback on the screen.
 In our example, we only generate application events for the 5 key events, and
-we also exhibit the pressed key in real time:
+we also exhibit the pressed key in real time as a visual feedback:
 
@@
 Regarding (b), we use an absolute simulation clock to track, in real time, the
-execution pace of each peer in comparison to a reference peer most advanced in
-time.
+execution pace of each peer in comparison to the reference peer, which is the
+one most advanced in time.
 %
-Regarding (c), we designate some peers to leave and join the network
-periodically, possibly creating partitions, to assert that they all catch up in
-real time.
+Regarding (c), we make peers leave and join the network periodically, possibly
+creating partitions, and assert that they all catch up in real time.
 
@@
 - colors indicate feasibility
-%
+
 We now detail how we instrument the middleware implementation to measure the
@@
 %
-We also start each peer with a random delay of up to $10s$ to force a time
-mismatch.
+We also start each peer with a random delay of up to $10s$ to force a
+synchronization mismatch from the beginning.
 %
@@
 \item \textbf{Correctness of unstable peers:}
+We run a separate simulation in which, for every $100s$, each peer remains
+$40s$ online and $20s$ offline, including peers \code{9--11} in the middle of
+the network.
+This creates partitions in which sides advance in time and generate events out
+of sync, which need to be merged and resynchronized constantly.
+We measure
     - state (even for stable peers)
-- time
-- remove 13 (nao partition) e 10 (partition)
     - qto tempo leva para sync
@@
 \end{enumerate}
+
+- ate 20
+- 30: bugs
+- 40: -> <- O(n), [500|500] future, (-5s) delay links
+
+- qual a natureza do baksS alto e baksI baixo?
 

### 2022-10-01 tpd-22.tex
...

@@
 \newcommand{\lin}[1]{(\emph{ln. #1}\xspace)}
+\newcommand{\linx}[1]{\emph{ln. #1}\xspace}
 
@@
 The application generates asynchronous events that go through the middleware,
-which synchronizes them with a timestamp in the future, such that all peers can
-satisfy.
+which synchronizes them with a timestamp scheduled to the future, such that all
+peers can satisfy.
 The middleware controls the execution of the application by issuing ticks and
@@
 
-\subsubsection{Events Broadcasting}
+\subsubsection{Event Broadcasting}
 \label{sec.tml.middleware.events}
@@
 As illustrated in Figure~\ref{fig.middleware}, events are broadcast between
-peers with a timestamp in the future such that all peers are able to apply them
-in sync.
+peers with a timestamp scheduled to the future such that all peers are able to
+apply them in sync.
 %
-To prevent broadcast cycles, event are tagged with a source peer and sequence
-number.
-Each peer keeps a vector with the highest sequence number it received from each
-other peer, such that lower numbers can be ignored when received.
-Higher numbers update the vector and trigger a broadcast to neighbours.
+To prevent broadcast cycles, each peer keeps a vector with the highest
+timestamp it received from each other peer, such that lower numbers can be
+ignored when received.
 Since we rely on TCP connections, it is not possible to receive out-of-order
-sequence numbers from the same source peer.
+timestamps attached to a given source peer.
+When the middleware receives a higher number from a given peer, it updates its
+vector position and triggers a broadcast to neighbours.
 %
 As mentioned in Section~\ref{sec.tml.api}, we assume that peers assign
-themselves unique identifiers in a contiguous range.
-The reasons are
-    (a) to fit a \code{uint8\_t} size, and
-    (b) to use a contiguous vector of sequence numbers.
-Therefore, this is not a serious limitation, and a separate mechanism could be
-used for assignment, with few modifications to core of the middleware.
+themselves unique identifiers in a contiguous range, such that they fit in a
+contiguous vector of timestamps.
+Therefore, this is not a severe limitation, and a separate mechanism could be
+used for assignments, with few modifications to core of the middleware.
 
@@
     uint8_t  src;       // source peer
-    uint32_t seq;       // sequence number
     uint32_t tick;      // tick to apply
@@
 01  void p2p_loop (...,cb_ini,cb_sim,cb_ren,cb_evt) {
-02    cb_ini(1);
+02    cb_ini(1);    // initialization
 03    cb_sim(<P2P_EVT_START>);
@@
 05    while (<app-running>) {
-06      if (<next-packet>) {
+06      if (<next-network-packet>) {
 07        // fast forward if I'm late
 08        if (<packet-sync-future>) {
-09          p2p_travel(<now>, <packet-sync-time>);
+09          p2p_travel(<now>, <packet-sync-time>, <speed>);
 10        }
@@
 13        if (<packet-evt-past>) {
-14          p2p_travel(<now>, <packet-evt-time>);
-15          p2p_travel(<packet-evt-time>, <now>);
+14          p2p_travel(<now>, <packet-evt-time>, <speed>);
+15          p2p_travel(<packet-evt-time>, <now>, <speed>);
 16        }
 17
-18        // simulate in real time if I'm on time
+18        // real time if I'm on time
 19        if (<packet-evt-ok>) {
@@
 24
-25      if (<next-input>) {
-26        // broadcast event from local
+25      if (<next-local-input>) {
+26        // broadcast local event
 27        p2p_evt evt;
@@
 33      // simulate tick in real time
-34      if (<next-tick>) {
+34      if (<next-local-tick>) {
 35        cb_sim(<P2P_EVT_TICK>);
@@
 40
-41    cb_ini(0);
+41    cb_ini(0);    // finalization
 42  }
@@
 Regarding network packets, if we receive a time synchronization packet
-\code{SYNC} from the future \lin{7--10}, then the we fast forward the
-application to catch up in time (Section~\ref{sec.tml}.b).
+\code{SYNC} from the future \lin{7--10}, then we fast forward the application
+to catch up in time (Section~\ref{sec.tml}.b).
 If we receive an event that should have been applied in the past \lin{12--16},
@@
 %
-Regarding time tick, we call \code{cb\_sim} and \code{cb\_ren} in real time
+Regarding time ticks, we call \code{cb\_sim} and \code{cb\_ren} in real time
 \lin{33--38}.
 The middleware also takes periodic snapshots of the application state to
-rollback and assist time travelling (Section~\ref{sec.tml.api.state}).
+optimize time travelling (Section~\ref{sec.tml.api.state}).
 
@@
 \begin{verbatim}
-01  void p2p_travel (int from, int to) {
+01  void p2p_travel (int from, int to, int ms) {
 02    for (i=from..to) {
@@
 
-The function \code{p2p\_travel} time travels, starting at tick \code{from},
-tick by tick, until it reaches tick \code{to} \lin{2--10}.
-The loop can go in any direction in time, i.e., \code{from} may be higher than
+The function \code{p2p\_travel} time travels the application, starting at tick
+\code{from}, tick by tick, until it reaches tick \code{to} \lin{2--10}.
+The loop can travel in both directions, i.e., \code{from} may be higher than
 \code{to}.
@@
 from the snapshot to the desired tick \lin{5--7}.
-We also exhibit each step with a small delay \lin{8--9} so that users
+We also exhibit each step with a small \code{ms} delay \lin{8--9} so that users
 experience smooth time transitions.
@@
     (a) peers meet event deadlines either in real time \lin{18--21} or
-        travelling back and forth in time \lin{12--16}
-        (Section~\ref{sec.tml.middleware.loop});
+        travelling back and forth in time
+        (\linx{12--16} / \ref{sec.tml.middleware.loop});
     (b) peers synchronize their clocks with periodic \code{SYNC} packets that
-        fast forwards late peers \lin{7--10}
-        (Section~\ref{sec.tml.middleware.loop}); and
+        fast forwards late peers
+        (\linx{7--10} / \ref{sec.tml.middleware.loop}); and
     (c) peers can join the network at any time and still catch up in time by
@@
 %
-We simulate an application varying the network topology, latency, and churn, as
-well as the frequency and deadlines of events.
+We simulate an application varying the network latency and churn, as well as
+the frequency and deadlines of events.
 %
@@
 an event after its deadline.
-We count all time travel occurrences during the simulation.
+We then count all time travel occurrences during the simulation.
 %
 Regarding (b), we use an absolute simulation clock to track, in real time, the
-execution pace of each peer in comparison to the reference peer, which is the
-one most advanced in time.
+execution pace of each peer.
+% in comparison to the reference peer, which is the one most advanced in time.
+We then count the mismatches between the clocks.
 %
 Regarding (c), we make peers leave and join the network periodically, possibly
-creating partitions, and assert that they all catch up in real time.
+creating partitions.
+We then assert that they all catch up in real time.
 
@@
     \label{fig.topo}
-    Network topology: 21 peers organized as a straight line ($0--3$), a simple
-    cycle ($3--8$), a second straight line ($8--11$), a full connected cycle
-    ($12--17$), and a third straight line ($17--20$).
-    An optional link ($0,20$) creates a long cycle in the network.
+    Network topology: 21 peers organized as a straight line ($0-3$), a simple
+    cycle ($3-8$), a second straight line ($8-11$), a full connected cycle
+    ($12-17$), and a third straight line ($17-20$).
+    %An optional link ($0,20$) creates a long cycle in the network.
   }
@@
 accumulated simulation ticks, then the final measure is $1\%$.
+TODO: in the figure, we consider range a-b, c-d...
 %
@@
 simulation.
+TODO: in the figure, we consider range a-b, c-d...
+TODO: in comparison to the reference peer, which is the one most advanced in time.
 %
@@
     - qto tempo leva para sync
+TODO: in the figure, we consider range a-b, c-d...
 %

### 2022-10-13 tpd-22.tex
- baks figure

@@
 %
-We simulate an application varying the network latency and churn, as well as
-the frequency and deadlines of events.
+We simulate an application varying the network latency and peer churn, as well
+as the frequency and deadlines of events.
 %
@@
 creating partitions.
-We then assert that they all catch up in real time.
+We then assert that they all catch up in real time within a short period.
 
@@
 Figure~\ref{fig.topo}, with peers organized in straight lines and cycles.
+The topology has an average of 5 hops between peers (e.g., \code{4--16} are 7
+hops away).
 %
@@
 %
-We simulate the network latency in software with random delays.
+We use \emph{NetEm}~\cite{netem} to simulate the network latency with a normal
+distribution (e.g., \code{tc qdisc add dev lo root netem delay 200ms 40ms
+distribution normal}).
+%
+We vary
+    (i)   the network latency (\code{5--500ms} between each peer),
+    (ii)  the rate of events (\code{5--200 evt/sec} in the whole network), and
+    (iii) the delay multiplier (\code{1--5x} considering the latency and the
+          average hops between peers).
+Regarding item (c), a \code{50ms} latency with a \code{2x} multiplier results
+in \code{500ms} event delays, i.e., each event is triggered with this delay to
+compensate the given latency and fixed average of 5 hops.
+%
+These variations result in $108$ combinations.
+We executed each combination $3$ times for $5$ minutes each.
+Therefore, the experiments take $27$ hours to complete.
 
-Figure~\ref{TODO} results...
-- colors indicate feasibility
+\begin{figure*}[t]
+  \centering
+  \includegraphics[width=\linewidth]{baks}
+  \caption{
+    \label{fig.baks}
+The \code{5x}, \code{2x}, and \code{1x} delay regions each contains $36$
+measures for multiple configurations of event rate (\code{5--200evt/min} and
+network latency (\code{5--500ms}).
+%
+The color thresholds (\emph{white}, \emph{yellow}, \emph{red}, \emph{black})
+are arbitrary but help to distinguish between regions of interest.
+%
+The experiments uses $1$ process for each peer in a single Linux machine
+(\emph{i7/16GB}).
+\end{figure*}
 
-We now detail how we instrument the middleware implementation to measure the
-properties above:
+We now detail the results and how we instrument the middleware implementation
+to measure the properties above:
 %
@@
 At the end, we measure the percentage of the sum of all time travel ticks over
-the total simulation ticks (\code{TODO} in Figure~\ref{TODO}).
+the total simulation ticks.
 As an example, if all peers sum $1000$ time travel ticks over $100k$
 accumulated simulation ticks, then the final measure is $1\%$.
-TODO: in the figure, we consider range a-b, c-d...
+%
+Figure~\ref{fig.baks} presents the results considering all combinations of
+\code{(i)}, \code{(ii)}, and \code{(iii)} above.
+The colors indicate the feasibility of each tested scenario
+    (\emph{white=good}, \emph{yellow=moderate}, \emph{red=poor},
+    \emph{black=faulty}).
+%
+For instance, the three rectangles labelled \code{(1)} represent a scenario
+with $50ms$ network delay, generating $25evt/min$.
+Each rectangle uses a different delay multiplier, showing a good performance
+for \code{5x}, moderate for \code{2x}, and poor for \code{1x}.
+%
+The color patterns indicate clear boundaries for the scenarios that can meet
+event deadlines with the desired performance.
 %
 \item \textbf{Real-time pace of peers:}
-Every time a peer receives \code{SYNC} packets and is behind, we output its id
-and how much ticks it needs to fast forward.
-At the end, we measure the percentage of the sum of all forward ticks over the
-total simulation ticks (\code{TODO} in Figure~\ref{TODO}).
-As an example, if all peers sum $1000$ forward ticks over $100k$ accumulated
-simulation ticks, then the final measure is $1\%$.
+All peers output their current ticks continuously, allowing us to compare them
+in the shared timeline.
+Every time we see a greater tick never seen in the timeline before, we count
+smaller ticks from other peers, which constitute time violations.
+As an example, if all peers sum $1000$ of such smaller ticks over $100k$
+accumulated simulation ticks, then the final measure is $1\%$.
 %
 We also start each peer with a random delay of up to $10s$ to force a
-synchronization mismatch from the beginning.
+synchronization mismatch at the beginning.
 %
-Each peer outputs its current tick continuously, allowing us to compare them
-in the shared timeline.
-Each time we see a greater tick never seen in the timeline before, we count
-smaller ticks from other peers (considering latency), which would constitute
-time violations (\code{TODO} in Figure~\ref{TODO}).
-We expect this counter to be close to zero even considering the whole
-simulation.
-TODO: in the figure, we consider range a-b, c-d...
-TODO: in comparison to the reference peer, which is the one most advanced in time.
+Considering all scenarios, the average is negligibly under $0.2\%$, with no
+variations.
+Therefore, we do not show the resulting table (such that of
+Figure~\ref{fig.baks}), which would not provide further insights.
 %
@@
 
-- ate 20
-- 30: bugs
-- 40: -> <- O(n), [500|500] future, (-5s) delay links
-
-- qual a natureza do baksS alto e baksI baixo?
+This could represent a...
+As an example of a concrete scenario, an interactive class over the Internet
+could have
+    an RTT of \emph{50--200ms} (\emph{Latency}),
+    interacting every \emph{5s} (\emph{Rate}).
+%e.g., a slide advance).
 

### 2022-10-18 tpd-22.tex
...

@@
 \begin{abstract}
-In collaborative networked applications, such as shared online documents,
-users can interact remotely and yet share the exact same experience as if they
-were in a single machine.
+In real-time networked applications, such as shared documents, users can
+interact remotely and yet share the exact same experience as if they were in a
+single machine.
 %
@@
 timeline across the network.
-Also, ased on memory snapshots and deterministic simulation, a ``time machine''
+Also, based on memory snapshots and deterministic simulation, a ``time machine''
 can rollback conflicting peers to resynchronize their state.
 %
-TODO: application, results
+We show that the middleware can handle applications with over 20 peers in a
+heterogeneous topology under high churn.
+For instance, in a simulation scenario of 25 events per minute with a network
+latency of $50ms$, the peers need to roll back and resynchronize only $3\%$ of
+the time.
 \end{abstract}
@@
 
-\IEEEPARstart{C}{ollaborative} networked applications allow multiple users to
-interact remotely and yet share the same experience in real time.
+\IEEEPARstart{R}{eal-time} networked applications allow multiple users to
+interact remotely and yet share the same experience.
 Examples of these \emph{symmetric distributed applications}~\cite{gals} are
@@
 However, current solutions depend on a central server to interconnect network
-nodes and determine the shared timeline.
+nodes and determine a shared timeline.
 
@@
 in time.
-Our main contribution is the design of a middleware that runs on each peer to
-ensure that all peers
+Our main contribution is the design of a middleware to ensure that all peers
     (a) meet event deadlines,
@@
 
-TODO: application, results
+We perform simulations with over 20 peers in a 5-hop mixed topology with cycles
+and straight lines.
+We vary the network latency and peer churn, as well as the frequency and
+deadlines of events.
+%
+We then measure (a) the recurrence of time travels, (b) the real-time pace of
+peers, and (c) the recoverability from churns, such that they validate our
+goals above, respectively.
+%
+As results, we show that the last two goals are met even under extreme
+conditions: peers exhibit a time mismatch of under $0.2\%$ and take less than
+$1.5s$ to recover from churns.
+%
+For the first goal, we identify the scenarios in which the middleware can meet
+event deadlines with the desired performance.
+As an example, in a scenario of 25 events per minute with a network latency of
+$50ms$, the peers need to roll back only $3\%$ of the time.
 
@@
     (c) how events are propagated and applied.
-Figure~\ref{fig.table} compares selected works regarding these aspects.
+Figure~\ref{fig.table} compares three selected works regarding these aspects.
 
@@
 \footnote{Croquet.io: https://croquet.io/}~\cite{croquet} guarantees
-bit-identical real-time behavior for every user in a collaborative
-distributed environment.
+bit-identical real-time behavior for users in collaborative distributed
+environments.
 %
@@
 to a synchronized global shared clock.
-If a user generates an event, it is delayed and sent back to the reflector,
-which broadcasts the event in the next tick, which all nodes apply in sync.
+If a user generates an event, it is sent back to the reflector, which
+broadcasts the event in the next tick, which all nodes apply in sync.
 %
@@
 As indicated in Figure~\ref{fig.table}, Croquet relies on a centralized
-network, in which nodes advance in sync, and in which event outcomes depend
-on the central server to be online.
+network (a), in which nodes advance in sync (b), and in which event outcomes
+depend on the central server to be online (c).
 
@@
 frame transitions, even those with poor connections.
-On the other hand, events are slower and clients may experience occasional
-freezes.
-The syncing protocol also requires bookkeeping to deal with clock drifts and
-client disconnections~\cite{gals}.
+On the other hand, events take longer to apply and clients may experience
+occasional freezes.
+The syncing protocol also requires extra bookkeeping to deal with clock drifts
+and client disconnections~\cite{gals}.
 %
 As indicated in Figure~\ref{fig.table}, GALS also relies on a centralized
-network, but in which nodes advance time independently, and in which event
-outcomes still depend on the central server to be online.
+network (a), but in which nodes advance time independently (b), and in which
+event outcomes still depend on the central server to be online (c).
 
@@
 be permanently online.
-All clients compute the events in sequence, respecting their timestamps, and
-using only deterministic and stateless calls.
+All clients compute events in sequence, respecting timestamps, and using only
+deterministic and stateless calls.
 This way, it is guaranteed that all clients reach the same state, and
@@
 An antagonistic approach to deterministic reactions to events is to model the
-application with conflict-free replicated data types (CRDTs)~\cite{crdts}.
+application with conflict-free replicated data types~\cite{crdts}.
 CRDTs are designed in such a way that all operations are commutative, so that
@@
 %
-As indicated in Figure~\ref{fig.table}, CRDTs work on decentralized networks,
-in which peers advance independently, and in which operations can be applied
-immediately, even while offline.
+As indicated in Figure~\ref{fig.table}, CRDTs work on decentralized networks
+(a), in which peers advance independently (b), and in which operations can be
+applied immediately, even while offline (c).
 %
 Automerge along with its accompanying Hypermerge peer-to-peer infrastructure
-is an example of an implementation of decentralized
-CRDTs~\cite{p2p.automerge,p2p.pushpin}.
+is an example of decentralized CRDTs~\cite{p2p.automerge,p2p.pushpin}.
 
 In this work, as indicated in Figure~\ref{fig.table}, our goal is to support
-peer-to-peer real-time symmetric execution, while still tolerating offline and
-out-of-order event generation.
+peer-to-peer real-time symmetric execution (a), while still tolerating
+out-of-order (b) and offline (c) event generation.
 %
@@
 %
-Like Croquet and GALS, peers compute events in sequence using only
-deterministic and stateless calls.
-
-Regarding decentralized time machines, Fusion%
+Like Croquet and GALS, peers compute events in sequence using deterministic and
+stateless calls only.
+
+We now discuss related work on time machines.
+%
+Regarding decentralized machines, Fusion%
 \footnote{Fusion: \url{https://doc.photonengine.com/en-us/fusion}}
@@
 central server.
-
-Regarding single-user time machines, the video game Braid%
-\footnote{\url{Braid: http://braid-game.com/}}
+%
+Regarding single-user local machines, the video game Braid%
+\footnote{Braid: \url{http://braid-game.com/}}
 is designed on the assumption that players have unrestricted rollbacks as part
@@
     \code{render(mem)} to output the current state.
-As we describe next, we use a similar approach by separating mutation from
-rendering~\cite{tml.alive}.
+As we describe next, we use a similar approach by separating state modification
+from rendering~\cite{tml.alive}.
 
@@
 
-In this section, we describe the architecture of our middleware and also the
-API in which developers have to conform with.
-%
 The middleware is written in C and relies on the library \emph{SDL}%
-\footnote{SDL: \url{www.libsdl.org}}
+\footnote{SDL: \url{http://www.libsdl.org}}
 to deal with the network, timers, and input events, such as mouse clicks and
@@
 moving rectangle.
-The six instances are connected in a peer-to-peer network, such that some
-peers can only communicate indirectly.
+The six instances are connected in a peer-to-peer network, such that most
+peers can only communicate indirectly to each other.
 %
@@
     The middleware orchestrates the peers, synchronizing their ticks locally
-    and broadcasting events with a delay.
+    and broadcasting events with a $\Delta$ delay.
     \label{fig.middleware}
@@
 The application itself is not aware of the network and has no direct access to
-the other peers, which communicate transparently through the middleware.
+other peers, which communicate transparently through the middleware.
 The application generates asynchronous events that go through the middleware,
-which synchronizes them with a timestamp scheduled to the future, such that all
-peers can satisfy.
+which synchronizes them with a timestamp with a $\Delta$ delay scheduled to the
+future, such that all peers can satisfy.
 The middleware controls the execution of the application by issuing ticks and
@@
 Due to the inherent latency of networks, peers generate events with a
-deliberate delay so that they can reach other peers in time to be applied in
+deliberate $\Delta$ so that they can reach other peers in time to be applied in
 sync.
 %We first assume a predetermined delay for the sake of simplicity.
-Nevertheless, a peer that is locally ahead of time may receive an event that
-should have been applied in the past, even considering the delay.
+Nevertheless, a peer ahead of time may receive an event that should have been
+applied in the past, even considering the delay.
 In this case, the middleware automatically rolls back the peer, applies the
@@
 the network.
-However, when a peer joins the network, the middleware needs to ensure that the
-peer receives all events that ever happened in order, which is trivial since
-they all hold an absolute timestamp in the shared timeline.
+However, when a peer joins the network, the middleware needs to ensure that it
+receives from other peer all events that ever happened, in order.
+This is trivial since all peers already need to hold the full event history (as
+described in item (a)), including their absolute timestamps in the shared
+timeline.
 \end{enumerate}
@@
 
-The application generates events that need to be broadcast to the network so
-that all peers behave the same.
-Note that the application events may (or may not) differ from low-level local
+When an application generates events, they need to be broadcast to the network
+so that all peers behave the same.
+Note that application events may (or may not) differ from low-level local
 system events.
@@
     \code{P2P\_EVT\_TICK} is generated every frame.
-The application-specific events must be declared by the programmer in an
+The other application-specific events must be declared by the programmer in an
 enumeration.
-In our application, we define a new event \code{P2P\_EVT\_KEY} to represent key
+In our example, we define a new event \code{P2P\_EVT\_KEY} to represent key
 presses:
@@
 enum {
-    P2P_EVT_KEY = P2P_EVT_NEXT
+    P2P_EVT_KEY = P2P_EVT_NEXT // next to predefined
 };
@@
 the loop.
-The callback initializes (and finalizes) the network topology, as well as
+The callback should initialize (and finalize) the network topology, as well as
 static immutable globals that can live outside the simulation memory, such as
@@
 
-The middleware calls \code{cb\_sim} every frame, which affects the simulation
-state \code{G} deterministically.
-The occurring event is passed as an argument to the callback, which must never
+The middleware calls \code{cb\_sim} every frame or event occurrence, which
+should affect the simulation state \code{G} deterministically.
+An occurring event may be passed as argument to the callback, which must never
 perform any side effects, such as stateful calls or rendering frames.
@@
 topology, or exhibiting instant feedback on the screen.
-In our example, we only generate application events for the 5 key events, and
-we also exhibit the pressed key in real time as a visual feedback:
+In our example, we only generate application events for the 5 key events of
+interest, but we also exhibit the pressed key in real time as a visual
+feedback:
 
@@
 08        if (<packet-sync-future>) {
-09          p2p_travel(<now>, <packet-sync-time>, <speed>);
+09          p2p_travel(<now>, <pak-sync-time>, <vel>);
 10        }
@@
 13        if (<packet-evt-past>) {
-14          p2p_travel(<now>, <packet-evt-time>, <speed>);
-15          p2p_travel(<packet-evt-time>, <now>, <speed>);
+14          p2p_travel(<now>, <pak-evt-time>, <vel>);
+15          p2p_travel(<pak-evt-time>, <now>, <vel>);
 16        }
@@
 %
-We simulate an application varying the network latency and peer churn, as well
-as the frequency and deadlines of events.
+We perform simulation varying the network latency and peer churn, as well as
+the frequency and deadlines of events.
 %
@@
 
-We simulate the application with $21$ peers using the mixed topology of
+\begin{figure}[t]
+  \centering
+  \includegraphics[width=2in]{mult}
+  \caption{
+    \label{fig.mult}
+Event delays in $ms$ considering the network latency, delay multiplier, and the
+average of 5 hops between peers from Figure~\ref{fig.topo}.
+    }
+\end{figure}
+
+We perform simulations with 21 peers using the mixed topology of
 Figure~\ref{fig.topo}, with peers organized in straight lines and cycles.
@@
 We use \emph{NetEm}~\cite{netem} to simulate the network latency with a normal
-distribution (e.g., \code{tc qdisc add dev lo root netem delay 200ms 40ms
-distribution normal}).
+distribution.
+% (e.g., \code{tc qdisc add dev lo root netem delay 200ms 40ms distribution normal}).
 %
@@
           average hops between peers).
-Regarding item (c), a \code{50ms} latency with a \code{2x} multiplier results
-in \code{500ms} event delays, i.e., each event is triggered with this delay to
+%
+Figure~\ref{fig.mult} shows the event delays considering the network latency
+and chosen multiplier.
+For instance, a \code{50ms} latency with a \code{2x} multiplier results in
+\code{500ms} event delays, i.e., each event is triggered with this delay to
 compensate the given latency and fixed average of 5 hops.
@@
 We executed each combination $3$ times for $5$ minutes each.
-Therefore, the experiments take $27$ hours to complete.
+We also make some modifications to evaluate peer churn, which requires to
+re-execute the simulation.
+The experiments take around $40$ hours to complete.
 
@@
 (\emph{i7/16GB}).
+    }
 \end{figure*}
@@
 Figure~\ref{fig.baks} presents the results considering all combinations of
-\code{(i)}, \code{(ii)}, and \code{(iii)} above.
+parameters (i), (ii), and (iii) above.
 The colors indicate the feasibility of each tested scenario
@@
 %
-For instance, the three rectangles labelled \code{(1)} represent a scenario
-with $50ms$ network delay, generating $25evt/min$.
+For instance, the three framed rectangles ($0.1$, $3.1$, and $7.2$) represent
+the scenarios with $50ms$ network delay, generating $25evt/min$.
 Each rectangle uses a different delay multiplier, showing a good performance
-for \code{5x}, moderate for \code{2x}, and poor for \code{1x}.
-%
-The color patterns indicate clear boundaries for the scenarios that can meet
-event deadlines with the desired performance.
+for $5x$, moderate for $2x$, and poor for $1x$.
 %
@@
 Considering all scenarios, the average is negligibly under $0.2\%$, with no
-variations.
-Therefore, we do not show the resulting table (such that of
-Figure~\ref{fig.baks}), which would not provide further insights.
+significant variance.
+Therefore, we omit the resulting table akin to Figure~\ref{fig.baks}), which
+would not provide further insights.
 %
 \item \textbf{Correctness of unstable peers:}
-We run a separate simulation in which, for every $100s$, each peer remains
-$40s$ online and $20s$ offline, including peers \code{9--11} in the middle of
-the network.
-This creates partitions in which sides advance in time and generate events out
-of sync, which need to be merged and resynchronized constantly.
-We measure
-    - state (even for stable peers)
-    - qto tempo leva para sync
-TODO: in the figure, we consider range a-b, c-d...
+We tweak our simulation to make each peer remain $40s$ online and $20s$
+offline in the average, including peers \code{9--11} in the middle of the
+network.
+This creates partitions in the network, which generate events out of sync that
+require constant resynchronization between peers.
+%
+Then, to evaluate the network correctness under such high churn, we count how
+much time the recovering peers take to catch up in real time:
+When a peer becomes online, we take the current maximum tick considering all
+reachable peers, and then count the time the recovering peer takes to catch up.
+We also assert that they all reach the same final state eventually.
+%
+Considering all scenarios, the average is $1.3$ seconds of recovering time,
+with no significant variance.
+Therefore, we also omit the resulting table akin to Figure~\ref{fig.baks}).
 %
@@
 
-This could represent a...
-As an example of a concrete scenario, an interactive class over the Internet
-could have
-    an RTT of \emph{50--200ms} (\emph{Latency}),
-    interacting every \emph{5s} (\emph{Rate}).
-%e.g., a slide advance).
-
-After $Xs$, .
-and then verify if they synchronize and remain synchronized during the whole
-simulation.
-We teak the middleware to output its current tick every $100ms$.
-
-More concretely, we measure the percentage o ticks diverging from the reference
-peer (e.g., $1\%$, if .
-
-For instance, if the reference peer signals $T$ and then a peer signals $T-dt$,
-than we.
-That would be the case when a source peer generates an event with a small
-deadline in comparison to the latency to the farthest destination peer.
-
-We run the simulation for 5 minutes and, since we start the measures after 1 minute
-
-- check if all ticks of all peers are in the same state
-
-- VARIABLES
-- number of nodes
-- topology
-    - distance b/w nodes
-    - number of edges
-- latency b/w nodes
-- freq of events
-- ~FPS~
-- future delta
-
-- RESULTS
-- time travels
-- ~real time~
-- ~enter leave~ (catches up?)
-
--- use of space for snapshots
--    - limit to go back?
--    - super nodes? (when connecting, pass tick, may close if tick<last)
-
--- start with random deltas
+\begin{comment}
+$ vi test.c: // #define TEST_OFFLINE
+$ sudo ./test-ana.sh 2>&1 | tee all.log
+$ ./test-sheet.sh > sheet.log
+$ ./test-sheet-2.sh > sheet-2.log
+$ lua5.3 test-sheet-2.lua sheet-2.log > /tmp/x
+$ gedit /tmp/x
+
+$ vi test.c: #define TEST_OFFLINE
+$ sudo ./test2-ana.sh 2>&1 | tee all2.log
+$ lua5.3 test2-sheet.lua
+\end{comment}
+
+Our results show that goals (b) and (c) are met in all scenarios consistently,
+and we consider them nearly optimal:
+Regardless of the network latency, event rate, and delay multiplier,
+    the variation in the pace of peers is under $0.2\%$ (goal b), and
+    recovering peers take $1.3$ seconds to catch up in time (goal c).
+Regarding goal (c), note that the middleware is configured to make small steps
+towards smooth time transitions, which take $1$ second to complete.
+Therefore, the average of $1.3$ seconds shows that peers need no more than 2
+time travels while recovering.
+
+Considering goal (a), we need to take a closer look at Figure~\ref{fig.baks}.
+The color patterns indicate clear boundaries for the scenarios that can meet
+event deadlines with the desired performance.
+%
+For instance, high event delays (column $5x$) make all scenarios viable,
+even with high latency and event frequency.
+As an application example, chats are not affected by high event delays.
+Nevertheless, the higher are the event delays, the less applications suit the
+middleware.
+%
+For the lowest possible delay (column $1x$), we see that the middleware can
+either handle low latency with high event frequency (e.g., $5ms$ with
+$100evt/min$) or high lantency with low event frequency (e.g., $100ms$ with
+$5evt/min$).
+
 \section{TODO}

### 2022-10-20 tpd-22.tex
...

@@
 As illustrated in Figure~\ref{fig.middleware}, events are broadcast between
-peers with a timestamp scheduled to the future such that all peers are able to
-apply them in sync.
+peers with a timestamp scheduled to the future with a $\Delta$ such that all
+peers are able to apply them in sync.
 %
@@
 themselves unique identifiers in a contiguous range, such that they fit in a
-contiguous vector of timestamps.
+simple vector of timestamps.
 Therefore, this is not a severe limitation, and a separate mechanism could be
-used for assignments, with few modifications to core of the middleware.
+used for assignments, with few modifications to the core of the middleware.
 
@@
 time.
-The peers must keep the full queue in memory for two reasons:
-    (a) receiving a late event may reorder the queue and require a time
-        travel, and
+Peers must keep the full event queue in memory for two reasons:
+    (a) receiving a late event may reorder it and require a time travel, and
     (b) peers that join the network need to receive all events.
@@
 Therefore, the middleware keeps a queue of ``packets'', which includes the
-event and other metadata:
+event and additional metadata:
 
@@
 struct {
-    int i;              // next event to apply
-    int n;              // number of events
+    int i;  // next event to apply in real time
+    int n;  // number of events in the queue
     p2p_pak buf[MAX];   // queue of packets
@@
 (Section~\ref{sec.tml.api}).
-We first (and last) call \code{cb\_ini} \lin{2,41} to initialize (and finalize)
-the application.
+We first (and last) call \code{cb\_ini} to initialize (and finalize) the
+application \lin{2,41}.
 Then, we call \code{cb\_sim} \lin{3}, passing the event \code{START} to start
@@
 %
-Regarding local inputs, we call \code{cb\_evt} \lin{26--30} to signal if the
-event should be broadcast, also transforming it to an application event
+Regarding local inputs, we call \code{cb\_evt} to signal if the event should be
+broadcast \lin{26--30}, also transforming it to an application event
 (Section~\ref{sec.tml.api.cb_evt}).
@@
 The function \code{p2p\_travel} time travels the application, starting at tick
-\code{from}, tick by tick, until it reaches tick \code{to} \lin{2--10}.
+\code{from}, tick by tick \lin{2--10}, until it reaches tick \code{to}.
 The loop can travel in both directions, i.e., \code{from} may be higher than
@@
 
-This section provided an overview of the middleware implementation to answer,
-in concrete terms, how it ensures the desired properties enumerated in the
-Introduction and at the beginning of Section~\ref{sec.tml.middleware}:
-    (a) peers meet event deadlines either in real time \lin{18--21} or
-        travelling back and forth in time
-        (\linx{12--16} / \ref{sec.tml.middleware.loop});
+This section provided an overview of the middleware implementation to answer
+how it ensures the desired properties in concrete terms as follows:
+    (a) peers meet event deadlines either travelling back and forth or in real
+        time (\linx{12--16} or \lin{18--22} / \ref{sec.tml.middleware.loop});
     (b) peers synchronize their clocks with periodic \code{SYNC} packets that
-        fast forwards late peers
+        fast forward late peers
         (\linx{7--10} / \ref{sec.tml.middleware.loop}); and
@@
 an event after its deadline.
-We then count all time travel occurrences during the simulation.
+We then count all time travel occurrences in the evaluation.
 %
@@
 % in comparison to the reference peer, which is the one most advanced in time.
-We then count the mismatches between the clocks.
+We then count the mismatches between the clocks in the evaluation.
 %
@@
 creating partitions.
-We then assert that they all catch up in real time within a short period.
+We then assert that they all catch up in real time within a short period in the
+evaluation.
 
@@
 Figure~\ref{fig.topo}, with peers organized in straight lines and cycles.
-The topology has an average of 5 hops between peers (e.g., \code{4--16} are 7
+The topology has an average of 5 hops between peers (e.g., \code{4-16} are 7
 hops away).
@@
 We vary
-    (i)   the network latency (\code{5--500ms} between each peer),
-    (ii)  the rate of events (\code{5--200 evt/sec} in the whole network), and
-    (iii) the delay multiplier (\code{1--5x} considering the latency and the
+    (i)   the network latency (\code{5-500ms} between each peer),
+    (ii)  the rate of events (\code{5-200 evt/sec} in the whole network), and
+    (iii) the delay multiplier (\code{1-5x} considering the latency and the
           average hops between peers).
@@
 The \code{5x}, \code{2x}, and \code{1x} delay regions each contains $36$
-measures for multiple configurations of event rate (\code{5--200evt/min} and
-network latency (\code{5--500ms}).
+measures for multiple configurations of event rate (\code{5-200evt/min} and
+network latency (\code{5-500ms}).
 %
@@
 We tweak our simulation to make each peer remain $40s$ online and $20s$
-offline in the average, including peers \code{9--11} in the middle of the
+offline in the average, including peers \code{9-11} in the middle of the
 network.

### 2022-10-24 tpd-22.tex
...

@@
 
+%\newpage
 \section{Evaluation}
@@
 
-In this section, we evaluate if (and to what extent) our design meets its goals
-to ensure that all peers (a) meet event deadlines, (b) advance in sync in real
-time, and (c) can leave and join the network and remain symmetric.
-%
-We perform simulation varying the network latency and peer churn, as well as
+In this section, we evaluate if our design meets the goals to ensure that all
+peers
+    (a) meet event deadlines,
+    (b) advance in sync in real time, and
+    (c) can leave and join the network and remain symmetric.
+%
+We perform simulations varying the network latency and peer churn, as well as
 the frequency and deadlines of events.
 %
-We then measure (a) the recurrence of time travels, (b) the real-time pace of
-peers, and (c) the recoverability from churns, such that they validate our
-goals above, respectively.
+We then measure
+    (a) the recurrence of time travels,
+    (b) the real-time pace of peers, and
+    (c) the recoverability from churns,
+such that they validate our goals above, respectively.
 
@@
 an event after its deadline.
-We then count all time travel occurrences in the evaluation.
+We then count all time travel occurrences in our evaluation.
 %
@@
 % in comparison to the reference peer, which is the one most advanced in time.
-We then count the mismatches between the clocks in the evaluation.
+We then count the mismatches between the clocks in our evaluation.
 %
@@
 creating partitions.
-We then assert that they all catch up in real time within a short period in the
+We then assert that they all catch up in real time within a short period in our
 evaluation.
@@
 %
+We run the application main loop at $50 FPS$ ($20ms$ per frame) and calculate
+the middleware overhead at each iteration, which is $9us$ in the average,
+corresponding to less than $1/2000$ of the CPU time.
+%
 In order to have an absolute clock, we use a single machine to simulate all
@@
 We vary
-    (i)   the network latency (\code{5-500ms} between each peer),
-    (ii)  the rate of events (\code{5-200 evt/sec} in the whole network), and
-    (iii) the delay multiplier (\code{1-5x} considering the latency and the
-          average hops between peers).
+    (i)   the network latency ($5-500ms$ between each peer),
+    (ii)  the rate of events ($5-200~evt/sec$), and
+    (iii) the delay multiplier ($1-5x$).
 %
@@
 and chosen multiplier.
-For instance, a \code{50ms} latency with a \code{2x} multiplier results in
-\code{500ms} event delays, i.e., each event is triggered with this delay to
-compensate the given latency and fixed average of 5 hops.
+For instance, a $50ms$ latency with a $2x$ multiplier results in $500ms$ event
+delays, i.e., each event is triggered with this delay to compensate the given
+latency and fixed average of 5 hops.
 %
@@
     \label{fig.baks}
-The \code{5x}, \code{2x}, and \code{1x} delay regions each contains $36$
-measures for multiple configurations of event rate (\code{5-200evt/min} and
-network latency (\code{5-500ms}).
+The $5x$, $2x$, and $x$ delay regions each contains $36$ measures for multiple
+configurations of event rate ($5-200~evt/min$ and network latency ($5-500ms$).
 %
@@
 \item \textbf{Recurrence of time travels:}
-Every time a peer is ahead of time and needs to travel back, we output its id
+Every time a peer is ahead in time and needs to travel back, we output its id
 and how much ticks it needs to travel back.
@@
 %
-For instance, the three framed rectangles ($0.1$, $3.1$, and $7.2$) represent
-the scenarios with $50ms$ network delay, generating $25evt/min$.
+For instance, the three framed rectangles ($0.1$, $3.1$, and $7.2$) in the
+figure represent the scenarios with $50ms$ network delay, generating
+$25~evt/min$.
 Each rectangle uses a different delay multiplier, showing a good performance
@@
 Then, to evaluate the network correctness under such high churn, we count how
-much time the recovering peers take to catch up in real time:
+much time a recovering peer takes to catch up in real time:
 When a peer becomes online, we take the current maximum tick considering all
@@
 with no significant variance.
-Therefore, we also omit the resulting table akin to Figure~\ref{fig.baks}).
+Therefore, we also omit the resulting table (akin to Figure~\ref{fig.baks}).
 %
@@
 Regardless of the network latency, event rate, and delay multiplier,
-    the variation in the pace of peers is under $0.2\%$ (goal b), and
+    the variation in the pace of peers is under $0.2\%$ (goal b), while
     recovering peers take $1.3$ seconds to catch up in time (goal c).
-Regarding goal (c), note that the middleware is configured to make small steps
-towards smooth time transitions, which take $1$ second to complete.
+Also regarding goal (c), note that the middleware makes smooth time transitions
+in small steps that take $1$ second to complete.
 Therefore, the average of $1.3$ seconds shows that peers need no more than 2
@@
 $5evt/min$).
-
-\section{TODO}
-
-TODO: simpler than vector clocks, absolute times are correct
-
-- p2p links por fora, como um servico TCP local
-
-We make some assumptions about the network, which are assigned externally to
-the middleware:
-%
-\begin{itemize}
-\item Peers identifiers are assigned externally to the middleware.
-\item Topology is dynamic, but also assigned externally
-
-\end{itemize}
-%
-Except for the last item, we claim that the others can be solved externally,
-without affecting the middleware.
-For instance,
-- orthogonal and would not affect the other algorithms in the middleware
-
-, which we claim that could be
-solved further:
-- ids are outside
-- topology, dynamic but outside
-- fixed maximum nodes
-- fixed latency
-We claim that Use external protocols
-For instance, latency is only a constant, no problem being a variable
-
-- non-malicious nodes
-
-\section{Application}
-\label{sec.app}
+%
+The color patterns in the table provide the necessary information to confront
+an application with its feasibility or desired performance.
+
+%\section{Application}
+%\label{sec.app}
 

### 2022-10-28 tpd-22.tex
...

@@
 distributed applications.
-In Section~\ref{sec.tml}, we detail our proposed middleware.
-In Section~\ref{sec.app}, ...
-In Section~\ref{sec.eval}, ...
-In Section~\ref{sec.conclusion}, ...
-
-- assumptions
-    - correct (non-malicious) nodes
-
-- local first
-- p2p
-- less restrictive (compared to CRDTs)
-- semantic events
-    - limitation b/c of rollbacks
-    - features b/c of network traffic
+In Section~\ref{sec.tml}, we detail the architecture of our middleware, its
+programming API, and how it orchestrates peers.
+%In Section~\ref{sec.app}, ...
+In Section~\ref{sec.eval}, we evaluate our design under a number of scenarios.
+In Section~\ref{sec.conclusion}, we conclude this work.
 
@@
 
-In this section, we evaluate if our design meets the goals to ensure that all
-peers
-    (a) meet event deadlines,
-    (b) advance in sync in real time, and
-    (c) can leave and join the network and remain symmetric.
-%
-We perform simulations varying the network latency and peer churn, as well as
-the frequency and deadlines of events.
-%
-We then measure
-    (a) the recurrence of time travels,
-    (b) the real-time pace of peers, and
-    (c) the recoverability from churns,
-such that they validate our goals above, respectively.
-
-Regarding (a), if a peer experiences a time travel, it means that it received
-an event after its deadline.
-We then count all time travel occurrences in our evaluation.
-%
-Regarding (b), we use an absolute simulation clock to track, in real time, the
-execution pace of each peer.
-% in comparison to the reference peer, which is the one most advanced in time.
-We then count the mismatches between the clocks in our evaluation.
-%
-Regarding (c), we make peers leave and join the network periodically, possibly
-creating partitions.
-We then assert that they all catch up in real time within a short period in our
-evaluation.
-
 \begin{figure*}[t]
@@
 
+In this section, we evaluate if our design meets the goals to ensure that all
+peers
+    (a) meet event deadlines,
+    (b) advance in sync in real time, and
+    (c) can leave and join the network and remain symmetric.
+%
+We perform simulations varying the network latency and peer churn, as well as
+the frequency and deadlines of events.
+%
+We then measure
+    (a) the recurrence of time travels,
+    (b) the real-time pace of peers, and
+    (c) the recoverability from churns,
+such that they validate our goals above, respectively.
+
+Regarding (a), if a peer experiences a time travel, it means that it received
+an event after its deadline.
+We then count all time travel occurrences in our evaluation.
+%
+Regarding (b), we use an absolute simulation clock to track, in real time, the
+execution pace of each peer.
+% in comparison to the reference peer, which is the one most advanced in time.
+We then count the mismatches between the clocks in our evaluation.
+%
+Regarding (c), we make peers leave and join the network periodically, possibly
+creating partitions.
+We then assert that they all catch up in real time within a short period in our
+evaluation.
+
 \begin{figure}[t]
@@
 %
-We run the application main loop at $50 FPS$ ($20ms$ per frame) and calculate
+We run the application main loop at $50~FPS$ ($20ms$ per frame) and calculate
 the middleware overhead at each iteration, which is $9us$ in the average,
@@
 The experiments take around $40$ hours to complete.
+
+% https://docs.google.com/spreadsheets/d/1xK5_wF9MoR-SmPGWVj0sGLrHxhpPBlLjVoS8U-Y7zTQ/edit#gid=63016757
 
@@
 to measure the properties above:
-%
-\begin{enumerate}
-%
-\item \textbf{Recurrence of time travels:}
+
+\textbf{(a) Recurrence of time travels:}
 Every time a peer is ahead in time and needs to travel back, we output its id
@@
 for $5x$, moderate for $2x$, and poor for $1x$.
-%
-\item \textbf{Real-time pace of peers:}
+
+\textbf{(b) Real-time pace of peers:}
 All peers output their current ticks continuously, allowing us to compare them
@@
 would not provide further insights.
-%
-\item \textbf{Correctness of unstable peers:}
+
+\textbf{(c) Correctness of unstable peers:}
 We tweak our simulation to make each peer remain $40s$ online and $20s$
@@
 Therefore, we also omit the resulting table (akin to Figure~\ref{fig.baks}).
-%
-\end{enumerate}
 
@@
     recovering peers take $1.3$ seconds to catch up in time (goal c).
-Also regarding goal (c), note that the middleware makes smooth time transitions
-in small steps that take $1$ second to complete.
-Therefore, the average of $1.3$ seconds shows that peers need no more than 2
-time travels while recovering.
+Note that the middleware makes smooth time transitions in small steps that take
+$1$ second to complete, resulting in no more than 2 time travels to recover
+peers.
 
@@
 \label{sec.conclusion}
+
+We propose a middleware to program \emph{symmetric peer-to-peer applications}
+in which interacting decentralized instances conform to identical behavior.
+%
+The middleware API shares similar limitations with centralized
+alternatives~\cite{gals,croquet}, being restricted to deterministic and
+stateless calls, and only supports pre-allocated memory.
+%
+Our main contribution is a ``time machine'' based on memory snapshots and
+deterministic simulation, allowing to decentralize the network timeline such
+that conflicting peers can rollback and resynchronize their state.
+%
+We assume that peers are non-malicious and that they can assign themselves
+unique identifiers in a contiguous range.
+
+The middleware can handle applications with over 20 peers in a heterogeneous
+topology under high churn.
+%
+We show that peers can
+    (a) broadcast events and meet deadlines globally, 
+    (b) advance in sync in real time, and
+    (c) leave and join the network and remain symmetric.
+%
+In our experiments, we vary the network latency and the rate and deadline of
+events.
+We than measure the recurrence of time travels in the network as a
+representative of quality.
+%
+As a result, we present a feasibility table to indicate the application
+scenarios that the middleware can handle.
+For instance, in a simulation scenario of 25 events per minute with a network
+latency of $50ms$, the peers need to roll back and resynchronize only $3\%$ of
+the time.
 

### 2022-10-28 p2p-22.tex
...

--- 2022-10-28 tpd-22.tex
@@
-\documentclass[10pt,journal,compsoc]{IEEEtran}
+\documentclass[sn-mathphys,iicol]{sn-jnl}% Math and Physical Sciences Reference Style 
+%\documentclass[default,iicol]{sn-jnl}% Default with double column layout
+
+\jyear{2022}%
+
+\raggedbottom
+%%\unnumbered% uncomment this for unnumbered level heads
 
@@
 
-\title{
+\title[Article Title]{
     Symmetric Peer-to-Peer Applications
@@
 
-\author{
-    Francisco Sant'Anna~\IEEEmembership{Department of Computer Science, Rio de Janeiro State University}
-}
-
-\IEEEtitleabstractindextext{%
-\begin{abstract}
+\author{\fnm{Francisco} \sur{Sant'Anna} \email{francisco@ime.uerj.br}}
+\affil{\orgdiv{Department of Computer Science}, \orgname{Rio de Janeiro State University}, \orgaddress{\country{Brazil}}}
+
+\abstract{
 In real-time networked applications, such as shared documents, users can
@@
 the time.
-\end{abstract}
-
-\begin{IEEEkeywords}
+}
+
+\keywords{
 collaboration, determinism, peer-to-peer, time machine
-\end{IEEEkeywords}}
+}
+
+%%\pacs[JEL Classification]{D8, H51}
+
+%%\pacs[MSC Classification]{35A01, 65L10, 65L12, 65L20, 65L70}
 
 \maketitle
-
-% TOTAL: 12 pages
 
@@
 
-\IEEEPARstart{R}{eal-time} networked applications allow multiple users to
-interact remotely and yet share the same experience.
+Real-time networked applications allow multiple users to interact remotely and
+yet share the same experience.
 Examples of these \emph{symmetric distributed applications}~\cite{gals} are
@@
 % docs.google.com/spreadsheets/d/1CpMmEgabJq2XQeTDW_BNCV4JWrcMlELuaEjDtw7kwg0/
-\begin{figure}[t]
+\begin{figure}
   \centering
@@
 
-\begin{figure*}[p]
+\begin{figure*}
   \centering
@@
 
-\begin{figure}[t]
+\begin{figure}
     \centering
@@
 
-\begin{figure}[t]
+\begin{figure}
   \centering
@@
 
-\begin{figure*}[t]
+In this section, we evaluate if our design meets the goals to ensure that all
+peers
+    (a) meet event deadlines,
+    (b) advance in sync in real time, and
+    (c) can leave and join the network and remain symmetric.
+%
+We perform simulations varying the network latency and peer churn, as well as
+the frequency and deadlines of events.
+%
+We then measure
+    (a) the recurrence of time travels,
+    (b) the real-time pace of peers, and
+    (c) the recoverability from churns,
+such that they validate our goals above, respectively.
+
+Regarding (a), if a peer experiences a time travel, it means that it received
+an event after its deadline.
+We then count all time travel occurrences in our evaluation.
+%
+Regarding (b), we use an absolute simulation clock to track, in real time, the
+execution pace of each peer.
+% in comparison to the reference peer, which is the one most advanced in time.
+We then count the mismatches between the clocks in our evaluation.
+%
+Regarding (c), we make peers leave and join the network periodically, possibly
+creating partitions.
+We then assert that they all catch up in real time within a short period in our
+evaluation.
+
+\begin{figure*}
   \centering
@@
 
-In this section, we evaluate if our design meets the goals to ensure that all
-peers
-    (a) meet event deadlines,
-    (b) advance in sync in real time, and
-    (c) can leave and join the network and remain symmetric.
-%
-We perform simulations varying the network latency and peer churn, as well as
-the frequency and deadlines of events.
-%
-We then measure
-    (a) the recurrence of time travels,
-    (b) the real-time pace of peers, and
-    (c) the recoverability from churns,
-such that they validate our goals above, respectively.
-
-Regarding (a), if a peer experiences a time travel, it means that it received
-an event after its deadline.
-We then count all time travel occurrences in our evaluation.
-%
-Regarding (b), we use an absolute simulation clock to track, in real time, the
-execution pace of each peer.
-% in comparison to the reference peer, which is the one most advanced in time.
-We then count the mismatches between the clocks in our evaluation.
-%
-Regarding (c), we make peers leave and join the network periodically, possibly
-creating partitions.
-We then assert that they all catch up in real time within a short period in our
-evaluation.
-
-\begin{figure}[t]
+\begin{figure}
   \centering
@@
 
-\begin{figure*}[t]
+\begin{figure*}
   \centering
@@
 
-\bibliographystyle{IEEEtran}
-\bibliography{tpd-22}
-
-\begin{comment}
-\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{chico}}]{Francisco Sant'Anna}
-received his PhD degree in Computer Science from PUC-Rio, Brazil in
-2013.
-In 2016, he joined the Faculty of Computer Science at the Rio de Janeiro State
-University, Brazil.
-His research interests include Programming Languages and Concurrent \&
-Distributed Systems.
-\end{IEEEbiography}
-\end{comment}
+\bibliography{p2p-22}
 

### 2022-10-29 final/p2p-22.tex
...

--- 2022-10-28 p2p-22.tex
@@
 \newcommand{\nreps}[1] {\emph{#1~reps\xspace}}
-\newcommand{\code}[1]  {\texttt{\footnotesize{#1}}}
+\newcommand{\code}[1]  {\texttt{\small{#1}}}
 \newcommand{\Xon} {$1{\rightarrow}N$\xspace}
@@
 \abstract{
-In real-time networked applications, such as shared documents, users can
+In real-time networked applications, such as collaborative documents, users can
 interact remotely and yet share the exact same experience as if they were in a
@@
 Also, based on memory snapshots and deterministic simulation, a ``time machine''
-can rollback conflicting peers to resynchronize their state.
+can rollback conflicting peers to resynchronize their states.
 %
 We show that the middleware can handle applications with over 20 peers in a
-heterogeneous topology under high churn.
-For instance, in a simulation scenario of 25 events per minute with a network
-latency of $50ms$, the peers need to roll back and resynchronize only $3\%$ of
+heterogeneous topologies under high churns.
+For instance, in a simulation scenario with a network latency of 50ms and 25
+events per minute, the peers need to roll back and resynchronize only 3\% of
 the time.
@@
 Examples of these \emph{symmetric distributed applications}~\cite{gals} are
-shared documents, watch parties, and multi-player games.
+collaborative documents, watch parties, and multi-player games~\cite{TODO:collab}.
 %
@@
 only, such that execution can be equally reproduced in all nodes according to
-the shared timeline.
+the shared timeline~\cite{TODO:det_reprod}.
 However, current solutions depend on a central server to interconnect network
@@
 Peers in the application form a dynamic network graph and communicate only
-with direct neighbours, as in typical unstructured peer-to-peer networks.
-Events are flooded in the graph and are triggered locally with a small delay
+with direct neighbours, as in typical unstructured peer-to-peer
+networks~\cite{TODO:p2p}.
+Events are flooded in the graph and are triggered with a small deadline delay
 to compensate the network latency.
@@
 event deadlines with the desired performance.
-As an example, in a scenario of 25 events per minute with a network latency of
-$50ms$, the peers need to roll back only $3\%$ of the time.
+As an example, in a scenario with a network latency of $50ms$ and 25 events per
+minute, the peers need to roll back only $3\%$ of the time.
 
@@
         (b) how global time is determined, and
-        (c) how events are propagated and applied.
+        (c) how events are broadcast and applied.
     \label{fig.table}
@@
 As indicated in Figure~\ref{fig.table}, Croquet relies on a centralized
-network (a), in which nodes advance in sync (b), and in which event outcomes
-depend on the central server to be online (c).
+network (a), in which nodes advance in sync (b), and in which event broadcasts
+and outcomes depend on the central server to be online (c).
 
@@
 network (a), but in which nodes advance time independently (b), and in which
-event outcomes still depend on the central server to be online (c).
+event broadcasts and outcomes still depend on the central server to be online
+(c).
 
@@
 Hence, there is no notion of a timeline in which peers go through
-bit-identical steps, but they do eventually reach the same state.
+bit-identical steps, but they do eventually reach the same
+state~\cite{TODO:eventual}.
 %
@@
 On the other hand, they provide very restrictive APIs (the CRDTs themselves),
-and do not support real-time consensus among peers.
+and do not support real-time consensus among peers~\cite{TODO:consensus}.
 %
@@
 In this work, as indicated in Figure~\ref{fig.table}, our goal is to support
-peer-to-peer real-time symmetric execution (a), while still tolerating
+peer-to-peer (a) real-time symmetric execution, while still tolerating
 out-of-order (b) and offline (c) event generation.
@@
 can communicate events directly to each other.
-Events are delayed to be able to reach other peers in time.
+Event deadlines are delayed to be able to reach other peers in time.
 In the case a deadline is missed, the peer rolls back in time and then fast
@@
 %
-Regarding decentralized machines, Fusion%
-\footnote{Fusion: \url{https://doc.photonengine.com/en-us/fusion}}
-is a state synchronization networking library for the Unity game engine.
+Regarding decentralized machines, Fusion~\cite{fusion} is a state
+synchronization networking library for the Unity game engine.
 It provides networked tick-based simulation in two modes: client-server
@@
     \code{render(mem)} to output the current state.
-As we describe next, we use a similar approach by separating state modification
+As we describe next, we use a similar approach by separating state mutation
 from rendering~\cite{tml.alive}.
@@
 At any moment, if a user pauses the application, all peers are guaranteed to
-pause exactly at the same frame with the rectangle being ``bit-identically'' at
-the same position.
+pause exactly at the same frame, with the rectangle ``bit-identically'' at the
+same position.
 
@@
     The middleware orchestrates the peers, synchronizing their ticks locally
-    and broadcasting events with a $\Delta$ delay.
+    and broadcasting events with a deadline delay ($\Delta$).
     \label{fig.middleware}
@@
 The application generates asynchronous events that go through the middleware,
-which synchronizes them with a timestamp with a $\Delta$ delay scheduled to the
-future, such that all peers can satisfy.
+which synchronizes them with a timestamp with a deadline delay ($\Delta$)
+scheduled to the future, such that all peers can satisfy.
 The middleware controls the execution of the application by issuing ticks and
@@
 Nevertheless, a peer ahead of time may receive an event that should have been
-applied in the past, even considering the delay.
+applied in the past, even considering its $\Delta$.
 In this case, the middleware automatically rolls back the peer, applies the
@@
 As we detail further, rolling back a peer requires to simulate the whole
-application execution from the beginning.
+application execution and event occurrences from the beginning.
 As an optimization, the middleware takes periodic snapshots locally to avoid
@@
 %
+To avoid abrupt visual changes, the middleware makes smooth time transitions
+that take up to $1s$ to complete.
+%
 \item \textbf{Peer Churn:}
-Considering that we target peer-to-peer networks, it is important that the
-middleware can handle arrival and departure of peers seamlessly.
+Considering that we target dynamic peer-to-peer networks, it is important that
+the middleware can handle arrival and departure of peers seamlessly.
 Note that high churn may lead to intermittent network partitions.
@@
 This is trivial since all peers already need to hold the full event history (as
-described in item (a)), including their absolute timestamps in the shared
-timeline.
+described in item (a)).
+% including their absolute timestamps in the shared timeline.
 \end{enumerate}
@@
 {\footnotesize
+~
 \begin{verbatim}
@@
 \end{verbatim}
-}
-
-The middleware assumes that each peer assigns itself a unique identifier in a
-contiguous range.
+~
+}
+
+The middleware assumes that each peer assigns itself a unique identifier, such
+as TODO~\cite{TODO:hash}.
 The FPS must be the same in all peers to ensure bit-identical simulation.
@@
 pool in \code{G} with a custom allocator.
-In our guiding example of Figure~\ref{fig.move}, the moving rectangle
-application only requires to hold the current position and direction:
+In our guiding example of Figure~\ref{fig.move}, the application only needs to
+hold the rectangle position and direction:
 
 {\footnotesize
+~
 \begin{verbatim}
@@
 \end{verbatim}
+~
 }
@@
 {\footnotesize
+~
 \begin{verbatim}
 enum {
-    P2P_EVT_KEY = P2P_EVT_NEXT // next to predefined
+    P2P_EVT_KEY = P2P_EVT_NEXT // key is next to predefs
 };
 \end{verbatim}
+~
 }
@@
 In our example, we modify the state of the rectangle as follows:
-    (a) reset its position and speed on \code{START},
-    (b) increment its position based on the current speed on \code{TICK}, and
-    (c) modify its speed based on \code{KEY}:
+    on \code{START}, reset its position and speed;
+    on \code{TICK},  increment its position based on the current speed; and
+    on \code{KEY},   modify its speed:
 
 {\footnotesize
+~
 \begin{verbatim}
@@
                     break;
-                <...> // same for other keys
+                <...> // similar for other keys
             }
@@
 render the screen multiple times while travelling.
+The middleware renders the screen only after each frame in real time.
 %, which the middleware calls only on real-time frames.
@@
 As illustrated in Figure~\ref{fig.middleware}, events are broadcast between
-peers with a timestamp scheduled to the future with a $\Delta$ such that all
-peers are able to apply them in sync.
-%
-To prevent broadcast cycles, each peer keeps a vector with the highest
-timestamp it received from each other peer, such that lower numbers can be
-ignored when received.
+peers with a timestamp scheduled to the future with a deadline $\Delta$, such
+that all peers are able to apply them in sync.
+%
+To prevent broadcast cycles, each peer keeps a vector of the highest timestamps
+it received from each other peer, such that lower numbers can be ignored when
+received.
 Since we rely on TCP connections, it is not possible to receive out-of-order
@@
 %
-As mentioned in Section~\ref{sec.tml.api}, we assume that peers assign
-themselves unique identifiers in a contiguous range, such that they fit in a
-simple vector of timestamps.
-Therefore, this is not a severe limitation, and a separate mechanism could be
-used for assignments, with few modifications to the core of the middleware.
+%As mentioned in Section~\ref{sec.tml.api}, we assume that peers assign
+%themselves unique identifiers in a contiguous range, such that they fit in a
+%simple vector of timestamps.
+%Therefore, this is not a severe limitation, and a separate mechanism could be
+%used for assignments, with few modifications to the core of the middleware.
 
@@
 \end{verbatim}
-}
-
-The function receives the application callbacks as arguments \lin{1}
-(Section~\ref{sec.tml.api}).
+~
+}
+
+The function receives the application callbacks as arguments \lin{1}.
+%(Section~\ref{sec.tml.api}).
 We first (and last) call \code{cb\_ini} to initialize (and finalize) the
@@
 Regarding local inputs, we call \code{cb\_evt} to signal if the event should be
-broadcast \lin{26--30}, also transforming it to an application event
+broadcast \lin{26--30}, and transformed into an application event
 (Section~\ref{sec.tml.api.cb_evt}).
@@
 
-This section provided an overview of the middleware implementation to answer
+This section provides an overview of the middleware implementation to answer
 how it ensures the desired properties in concrete terms as follows:
     (a) peers meet event deadlines either travelling back and forth or in real
-        time (\linx{12--16} or \lin{18--22} / \ref{sec.tml.middleware.loop});
+        time (\linx{12--16} or \linx{18--22} / \ref{sec.tml.middleware.loop});
     (b) peers synchronize their clocks with periodic \code{SYNC} packets that
@@
     \label{fig.mult}
-Event delays in $ms$ considering the network latency, delay multiplier, and the
-average of 5 hops between peers from Figure~\ref{fig.topo}.
+Event $\Delta$s in $ms$ considering the network latency, deadline multiplier,
+and the average of 5 hops between peers from Figure~\ref{fig.topo}.
     }
@@
 We run the application main loop at $50~FPS$ ($20ms$ per frame) and calculate
-the middleware overhead at each iteration, which is $9us$ in the average,
-corresponding to less than $1/2000$ of the CPU time.
+the middleware overhead at each iteration.
+We found an overhead of $9us$ in the average, which corresponds to less than
+$1/2000$ of the CPU time.
 %
@@
     (ii)  the rate of events ($5-200~evt/sec$), and
-    (iii) the delay multiplier ($1-5x$).
-%
-Figure~\ref{fig.mult} shows the event delays considering the network latency
+    (iii) the deadline $\Delta$ multiplier ($1-5x$).
+%
+Figure~\ref{fig.mult} shows the event $\Delta$s considering the network latency
 and chosen multiplier.
-For instance, a $50ms$ latency with a $2x$ multiplier results in $500ms$ event
-delays, i.e., each event is triggered with this delay to compensate the given
-latency and fixed average of 5 hops.
-%
-These variations result in $108$ combinations.
+For instance, a $50ms$ latency with a $2x$ multiplier results in
+$\Delta=500ms$, i.e., each event is triggered with this deadline delay to
+compensate the given latency and fixed average of 5 hops.
+%
+All simulation variations result in $108$ combinations.
 We executed each combination $3$ times for $5$ minutes each.
@@
     \label{fig.baks}
-The $5x$, $2x$, and $x$ delay regions each contains $36$ measures for multiple
-configurations of event rate ($5-200~evt/min$ and network latency ($5-500ms$).
+The $5x$, $2x$, and $x$ $\Delta$ regions each contains $36$ measures for
+multiple configurations of network latencies ($5-500ms$) and event rates
+($5-200~evt/min$).
 %
@@
 The experiments uses $1$ process for each peer in a single Linux machine
-(\emph{i7/16GB}).
+(\emph{Ubuntu 21.04, i7/16GB}).
     }
@@
 We now detail the results and how we instrument the middleware implementation
-to measure the properties above:
+to measure the properties of interest above:
 
@@
 For instance, the three framed rectangles ($0.1$, $3.1$, and $7.2$) in the
-figure represent the scenarios with $50ms$ network delay, generating
+figure represent the scenarios with $50ms$ network latency, generating
 $25~evt/min$.
-Each rectangle uses a different delay multiplier, showing a good performance
+Each rectangle uses a different $\Delta$ multiplier, showing a good performance
 for $5x$, moderate for $2x$, and poor for $1x$.
@@
 \textbf{(b) Real-time pace of peers:}
-All peers output their current ticks continuously, allowing us to compare them
-in the shared timeline.
+All peers output their ticks continuously, allowing us to compare them in the
+shared timeline.
 Every time we see a greater tick never seen in the timeline before, we count
@@
 significant variance.
-Therefore, we omit the resulting table akin to Figure~\ref{fig.baks}), which
+Therefore, we omit the resulting table (akin to Figure~\ref{fig.baks}), which
 would not provide further insights.
@@
     recovering peers take $1.3$ seconds to catch up in time (goal c).
-Note that the middleware makes smooth time transitions in small steps that take
-$1$ second to complete, resulting in no more than 2 time travels to recover
-peers.
+As discussed in Section~\ref{sec.tml}.b, the middleware may take up to $1$
+second to catch up, resulting in no more than 2 time travels to recover peers.
 
@@
 %
-For instance, high event delays (column $5x$) make all scenarios viable,
-even with high latency and event frequency.
-As an application example, chats are not affected by high event delays.
-Nevertheless, the higher are the event delays, the less applications suit the
+For instance, high event deadlines (column $5x$) make all scenarios viable,
+even with high latencies and event frequencies.
+As an application example, chats are not affected by higher event deadlines.
+Nevertheless, the higher are the deadlines, the less applications suit the
 middleware.
 %
-For the lowest possible delay (column $1x$), we see that the middleware can
-either handle low latency with high event frequency (e.g., $5ms$ with
-$100evt/min$) or high lantency with low event frequency (e.g., $100ms$ with
+For the lowest possible $\Delta$ (column $1x$), we see that the middleware can
+either handle low latencies with high event frequencies (e.g., $5ms$ with
+$100evt/min$) or high lantencies with low event frequencies (e.g., $100ms$ with
 $5evt/min$).
 %
-The color patterns in the table provide the necessary information to confront
-an application with its feasibility or desired performance.
+In summary, the color patterns in the table provide the necessary information
+to confront an application with its feasibility or desired performance.
 
@@
 We propose a middleware to program \emph{symmetric peer-to-peer applications}
-in which interacting decentralized instances conform to identical behavior.
+in which interacting peers conform to identical behavior without a centralized
+coordination server.
 %
@@
 alternatives~\cite{gals,croquet}, being restricted to deterministic and
-stateless calls, and only supports pre-allocated memory.
-%
-Our main contribution is a ``time machine'' based on memory snapshots and
-deterministic simulation, allowing to decentralize the network timeline such
-that conflicting peers can rollback and resynchronize their state.
+stateless calls. %, and only supports pre-allocated memory.
+%
+Our main contribution is a time machine that supports rollbacks based on memory
+snapshots and deterministic simulation.
+This allows to decentralize the network timeline, since conflicting peers can
+resynchronize their state through time travels.
 %
 We assume that peers are non-malicious and that they can assign themselves
-unique identifiers in a contiguous range.
+unique identifiers.
 
 The middleware can handle applications with over 20 peers in a heterogeneous
-topology under high churn.
+topologies under high churns.
 %
@@
 %
-As a result, we present a feasibility table to indicate the application
-scenarios that the middleware can handle.
-For instance, in a simulation scenario of 25 events per minute with a network
-latency of $50ms$, the peers need to roll back and resynchronize only $3\%$ of
-the time.
+As a result, we present a performance table to indicate the feasibility of
+diverse application scenarios.
+For instance, in a simulation scenario with a network latency of 50ms and 25
+events per minute, the peers need to rollback only 3\% of the time.
 

### 2022-10-31 final/p2p-22.tex
...

@@
 \usepackage{verbatim}
-\usepackage{url}
 \usepackage{graphicx}
@@
 
-\author{\fnm{Francisco} \sur{Sant'Anna} \email{francisco@ime.uerj.br}}
-\affil{\orgdiv{Department of Computer Science}, \orgname{Rio de Janeiro State University}, \orgaddress{\country{Brazil}}}
+%\author{\fnm{Francisco} \sur{Sant'Anna} \email{francisco@ime.uerj.br}}
+%\affil{\orgdiv{Department of Computer Science}, \orgname{Rio de Janeiro State University}, \orgaddress{\country{Brazil}}}
 
@@
 In real-time networked applications, such as collaborative documents, users can
-interact remotely and yet share the exact same experience as if they were in a
-single machine.
+interact remotely and yet share the exact same experience as if they were using
+a single machine.
 %
@@
 Peers are allowed to join and leave the network at any time.
-Application developers must adhere to a restricted API, which is deterministic,
-stateless, and only supports pre-allocated memory.
+Application developers must adhere to a deterministic and stateless API.
+%, and only supports pre-allocated memory.
 %
@@
 Examples of these \emph{symmetric distributed applications}~\cite{gals} are
-collaborative documents, watch parties, and multi-player games~\cite{TODO:collab}.
+collaborative documents, watch parties, and multi-player games.
+%~\cite{TODO:collab}.
 %
@@
 only, such that execution can be equally reproduced in all nodes according to
-the shared timeline~\cite{TODO:det_reprod}.
+the shared timeline. %~\cite{TODO:det_reprod}.
 However, current solutions depend on a central server to interconnect network
@@
 with direct neighbours, as in typical unstructured peer-to-peer
-networks~\cite{TODO:p2p}.
-Events are flooded in the graph and are triggered with a small deadline delay
-to compensate the network latency.
+networks~\cite{p2p.survey}.
+Events are flooded in the graph and are applied with a small deadline delay to
+compensate the network latency.
 To deal with events received out of order or too late, a distributed time
@@
 %
-For the first goal, we identify the scenarios in which the middleware can meet
+For the first goal, we analyze the scenarios in which the middleware can meet
 event deadlines with the desired performance.
-As an example, in a scenario with a network latency of $50ms$ and 25 events per
-minute, the peers need to roll back only $3\%$ of the time.
+As an example, peers need to roll back only $3\%$ of the time under a scenario
+with a network latency of $50ms$ and 25 events per minute.
 
@@
 
-Croquet%
-\footnote{Croquet.io: https://croquet.io/}~\cite{croquet} guarantees
-bit-identical real-time behavior for users in collaborative distributed
-environments.
+Croquet~\cite{croquet,croquet.site} guarantees bit-identical real-time behavior
+for users in collaborative distributed environments.
 %
@@
 support late joins to a running session.
-The new node just needs to restore the latest snapshot and simulate the
+The new node just needs to request the latest snapshot and simulate the
 remaining events to reach the current running state.
@@
 the server, clients have their own independent local clocks.
-Event generation requires two round trips (\emph{when→ask→ack→tick}) and is
-delayed dynamically according to the slowest client.
-%
-On the one hand, ticks do not generate any traffic and clients have smooth
-frame transitions, even those with poor connections.
+Event generation requires two round trips (\emph{when→ask→ack→tick}), resulting
+in dynamic deadlines according to the slowest client.
+%
+On the one hand, ticks do not generate any traffic, and clients experience
+smooth frame transitions, even those with poor connections.
 On the other hand, events take longer to apply and clients may experience
@@
 
-In both solutions, the network advances together as a whole in real time, with
+In both solutions, the network advances together in real time as a whole, with
 a total order among events, which is determined by a central server that must
@@
 An antagonistic approach to deterministic reactions to events is to model the
-application with conflict-free replicated data types~\cite{crdts}.
+applications with Conflict-free Replicated Data Types~\cite{crdts}.
 CRDTs are designed in such a way that all operations are commutative, so that
@@
 Also, since operations need not to be ordered, they can be applied at the
-very first moment they are generated or received, even if the peers are
-offline.
-%
-Since operations must be commutative, it is not possible to timestamp them
-in a unique global shared clock.
-Hence, there is no notion of a timeline in which peers go through
-bit-identical steps, but they do eventually reach the same
-state~\cite{TODO:eventual}.
+very first moment they are generated or received, even if the originating peers
+are offline.
+%
+Note that it is not possible to timestamp commutative operations under a unique
+global shared clock.
+Hence, there is no notion of a timeline in which peers go through bit-identical
+steps, but they do eventually reach the same state~\cite{crdts.eventual}.
 %
@@
 On the other hand, they provide very restrictive APIs (the CRDTs themselves),
-and do not support real-time consensus among peers~\cite{TODO:consensus}.
+and do not support real-time consensus among peers~\cite{crdts.consensus}.
 %
@@
 %
-Regarding decentralized machines, Fusion~\cite{fusion} is a state
+Regarding decentralized solutions, Fusion~\cite{fusion} is a state
 synchronization networking library for the Unity game engine.
@@
 Regarding single-user local machines, the video game Braid%
-\footnote{Braid: \url{http://braid-game.com/}}
+\footnote{Braid: \burl{http://braid-game.com/}}
 is designed on the assumption that players have unrestricted rollbacks as part
@@
 
-The middleware is written in C and relies on the library \emph{SDL}%
-\footnote{SDL: \url{http://www.libsdl.org}}
+Our middleware is written in C and relies on the library \emph{SDL}%
+\footnote{SDL: \burl{http://www.libsdl.org}}
 to deal with the network, timers, and input events, such as mouse clicks and
@@
 The full source code%
-\footnote{TODO}
+\footnote{Code: \burl{http://github.com/TODO-hidden-blind-review}}
 is under 500 lines of code.
@@
         mirrors of a single application.
-        Video: \url{http://youtube.com/TODO}
+        Video: \burl{http://youtube.com/TODO-hidden-blind-review}
         \label{fig.move}
@@
 We guide our discussion through the didactic example of Figure~\ref{fig.move},
-in which multiple users use the arrow keys to control the direction of a
-moving rectangle.
+in which multiple users control the direction of a moving rectangle using the
+arrow keys.
 The six instances are connected in a peer-to-peer network, such that most
@@
 pause exactly at the same frame, with the rectangle ``bit-identically'' at the
-same position.
+very same position.
 
@@
 synchronized events.
+Note that ticks need not to be broadcast, only $\Delta$ events.
 
@@
 Nevertheless, a peer ahead of time may receive an event that should have been
-applied in the past, even considering its $\Delta$.
+applied in the past. %, even considering its $\Delta$.
 In this case, the middleware automatically rolls back the peer, applies the
@@
 
-The middleware assumes that each peer assigns itself a unique identifier, such
-as TODO~\cite{TODO:hash}.
+The middleware assumes that each peer assigns itself a unique identifier (e.g.,
+a combination its MAC \& IP addresses).
+%, such as TODO~\cite{TODO:hash}.
 The FPS must be the same in all peers to ensure bit-identical simulation.
+
+Note that the use of callbacks is a common technique for event-driven
+programming in Games, GUIs, and interactive applications in general, sharing
+recognition with related techniques such as \emph{MVC},
+\emph{Publish-Subscribe}, and \emph{Observer} patterns~\cite{meyer,nystrom}.
+%
+Furthermore, the default run-to-completion semantics of callbacks is key to
+ensure the desired determinism, since unlike threads, they run sequentially and
+atomically from start to end~\cite{events,threads}.
+
+%As pointed by Berry and Benveniste, the advantages of deterministic systems should
+%be obvious: there is no reason a programmer should want his/hers
+%programs to behave in some non-deterministic manner [2].
+%A. Benveniste and G. Berry. The Synchronous Approach to Reactive and Real-
+%Time Systems. Proceedings of the IEEE, 79(9):1270–1282, 1991.
+%see also twelve
 
@@
 states.
-If dynamic allocation is required, it is necessary to hold a finite memory
-pool in \code{G} with a custom allocator.
 In our guiding example of Figure~\ref{fig.move}, the application only needs to
@@
 
+If dynamic allocation is required, it is necessary to hold a finite memory
+pool in \code{G} with a custom allocator.
+This is a limitation of C, since languages with serialization and
+meta-programming offer better mechanisms to take memory snapshots.
+Another alternative is to provide a serialization callback, which we left for
+future work.
+
 \subsubsection{Application Events}
@@
 %
-The middleware predefines two events for all applications:
+The middleware predefines three events for all applications:
+    \code{P2P\_EVT\_SYNC} represents synchronization events
+        (Section \ref{sec.tml}.b).
     \code{P2P\_EVT\_START} represents the beginning of the simulation, and
@@
 should affect the simulation state \code{G} deterministically.
-An occurring event may be passed as argument to the callback, which must never
+Occurring events are passed as arguments to the callback, which must never
 perform any side effects, such as stateful calls or rendering frames.
@@
 render the screen multiple times while travelling.
-The middleware renders the screen only after each frame in real time.
+The middleware renders the screen after each frame in real time.
 %, which the middleware calls only on real-time frames.
@@
 As illustrated in Figure~\ref{fig.middleware}, events are broadcast between
-peers with a timestamp scheduled to the future with a deadline $\Delta$, such
-that all peers are able to apply them in sync.
-%
-To prevent broadcast cycles, each peer keeps a vector of the highest timestamps
-it received from each other peer, such that lower numbers can be ignored when
-received.
+peers with a deadline $\Delta$ scheduled to the future, such that all peers are
+able to apply them in sync.
+%
+To prevent broadcast cycles, each peer keeps a collection of the highest
+timestamps it received from each other peer, such that lower numbers can be
+ignored when received.
 Since we rely on TCP connections, it is not possible to receive out-of-order
 timestamps attached to a given source peer.
-When the middleware receives a higher number from a given peer, it updates its
-vector position and triggers a broadcast to neighbours.
+When the middleware receives a higher number from a given peer, it updates this
+number and triggers a broadcast to neighbours.
 %
@@
 Then, we call \code{cb\_sim} \lin{3}, passing the event \code{START} to start
-the application in real time
-(Sections~\ref{sec.tml.api.events}~and~\ref{sec.tml.api.cb_sim}).
+the application in real time.
+%(Sections~\ref{sec.tml.api.events}~and~\ref{sec.tml.api.cb_sim}).
 The main loop \lin{5--39} checks continuously for network packets \lin{6--23},
@@
 \code{SYNC} from the future \lin{7--10}, then we fast forward the application
-to catch up in time (Section~\ref{sec.tml}.b).
+to catch up in time. % (Section~\ref{sec.tml}.b).
 If we receive an event that should have been applied in the past \lin{12--16},
-then we travel back and forth (Section~\ref{sec.tml}.a).
-Otherwise, we just apply the event in real time \lin{18--22}
-(Section~\ref{sec.tml.api.cb_sim}).
+then we travel back and forth. % (Section~\ref{sec.tml}.a).
+Otherwise, we just apply the event in real time \lin{18--22}.
+%(Section~\ref{sec.tml.api.cb_sim}).
 We detail the function \code{p2p\_travel} in
@@
 Regarding local inputs, we call \code{cb\_evt} to signal if the event should be
-broadcast \lin{26--30}, and transformed into an application event
-(Section~\ref{sec.tml.api.cb_evt}).
+broadcast \lin{26--30}, and transformed into an application event.
+%(Section~\ref{sec.tml.api.cb_evt}).
 %
@@
 The middleware also takes periodic snapshots of the application state to
-optimize time travelling (Section~\ref{sec.tml.api.state}).
+optimize time travelling. % (Section~\ref{sec.tml.api.state}).
 
@@
     (a) peers meet event deadlines either travelling back and forth or in real
-        time (\linx{12--16} or \linx{18--22} / \ref{sec.tml.middleware.loop});
+        time (Section \ref{sec.tml.middleware.loop}: \linx{12--16} or
+        \linx{18--22});
     (b) peers synchronize their clocks with periodic \code{SYNC} packets that
-        fast forward late peers
-        (\linx{7--10} / \ref{sec.tml.middleware.loop}); and
+        fast forward late peers (Section \ref{sec.tml.middleware.loop}:
+        \linx{7--10}); and
     (c) peers can join the network at any time and still catch up in time by
@@
 the middleware overhead at each iteration.
-We found an overhead of $9us$ in the average, which corresponds to less than
-$1/2000$ of the CPU time.
+We found a negligible overhead of $9us$ in the average, which corresponds to
+less than $1/2000$ of the CPU time.
 %
@@
 %
-Figure~\ref{fig.mult} shows the event $\Delta$s considering the network latency
-and chosen multiplier.
+Figure~\ref{fig.mult} shows the $\Delta$s considering the network latency and
+chosen multiplier.
 For instance, a $50ms$ latency with a $2x$ multiplier results in
 $\Delta=500ms$, i.e., each event is triggered with this deadline delay to
-compensate the given latency and fixed average of 5 hops.
-%
+compensate the given latency and fixed average of 5 hops
+($5 \times 50 \times 2 = 500$).
+
 All simulation variations result in $108$ combinations.
@@
 shared timeline.
-Every time we see a greater tick never seen in the timeline before, we count
-smaller ticks from other peers, which constitute time violations.
+Every time we see a greater tick never seen in the timeline before, we start to
+count smaller ticks from other peers, which constitute time violations.
 As an example, if all peers sum $1000$ of such smaller ticks over $100k$
@@
 Then, to evaluate the network correctness under such high churn, we count how
-much time a recovering peer takes to catch up in real time:
+much time a recovering peer takes to catch up in real time, as follows:
 When a peer becomes online, we take the current maximum tick considering all
@@
 As an application example, chats are not affected by higher event deadlines.
-Nevertheless, the higher are the deadlines, the less applications suit the
+Nevertheless, the higher is the deadline, the less applications suit the
 middleware.
@@
 %
-We assume that peers are non-malicious and that they can assign themselves
-unique identifiers.
+%We assume that peers are non-malicious and that they can assign themselves
+%unique identifiers.
 
 The middleware can handle applications with over 20 peers in a heterogeneous
-topologies under high churns.
+topologies and under high churns.
 %
@@
 events.
-We than measure the recurrence of time travels in the network as a
-representative of quality.
+We than measure the recurrence of time travels in the network as a quality
+proxy.
 %

### 2022-10-31 final/p2p-22.tex
...

@@
 Regarding single-user local machines, the video game Braid%
-\footnote{Braid: \burl{http://braid-game.com/}}
+\footnote{Braid: \url{http://braid-game.com/}}
 is designed on the assumption that players have unrestricted rollbacks as part
@@
 Our middleware is written in C and relies on the library \emph{SDL}%
-\footnote{SDL: \burl{http://www.libsdl.org}}
+\footnote{SDL: \url{http://www.libsdl.org}}
 to deal with the network, timers, and input events, such as mouse clicks and
@@
 The full source code%
-\footnote{Code: \burl{http://github.com/TODO-hidden-blind-review}}
+\footnote{Code: \url{http://github.com/TODO-hidden-blind-review}}
 is under 500 lines of code.
@@
         mirrors of a single application.
-        Video: \burl{http://youtube.com/TODO-hidden-blind-review}
+        Video: \url{http://youtube.com/TODO-hidden-blind-review}
         \label{fig.move}

### 2022-11-07 final/p2p-22.tex
...

@@
 
-%\author{\fnm{Francisco} \sur{Sant'Anna} \email{francisco@ime.uerj.br}}
-%\affil{\orgdiv{Department of Computer Science}, \orgname{Rio de Janeiro State University}, \orgaddress{\country{Brazil}}}
+\author{\fnm{Francisco} \sur{Sant'Anna} \email{francisco@ime.uerj.br}}
+\affil{\orgdiv{Department of Computer Science}, \orgname{Rio de Janeiro State University}, \orgaddress{\country{Brazil}}}
 
@@
 
+\section*{Declarations}
+
+\subsection*{Ethics Approval}
+
+Not applicable
+
+\subsection*{Conflict of Interest}
+
+The author declares no conflicts of interest.
+
+\subsection*{Data Availability}
+
+Not applicable
+
+\subsection*{Author Contribution}
+
+Francisco Sant'Anna is the single author and wrote the full manuscript.
+
+\subsection*{Funding}
+
+This study was partially funded by FAPERJ (grant E-26/211.329/2019 (251640)).
+
+\subsection*{Consent to publish}
+
+The author hereby expresses consent to publish.
+
 \bibliography{p2p-22}

### 2022-11-29 final/p2p-22.tex
- performance comparison

@@
 Figure~\ref{fig.table} compares three selected works regarding these aspects.
+%
+At the end of this section, we also discuss related work on local and
+distributed time machines.
+
+\subsection{Symmetric Distributed Applications}
 
@@
 
-We now discuss related work on time machines.
-%
-Regarding decentralized solutions, Fusion~\cite{fusion} is a state
+Many aspects may affect the performance of symmetric distributed applications,
+such as
+    the network latency,
+    the number of nodes in the network,
+    the rate of user events,
+    and the application frame rate.
+%
+It is expected that centralized solutions exhibit a better performance in
+comparison to peer-to-peer applications, since nodes are only 2 hops aways from
+each other.
+%
+The performance of Croquet depends exclusively on the network latency, which
+restricts the application to a maximum FPS.
+For instance, since all frame ticks need to be transmitted over the network,
+a latency of 50ms supports at most 20 FPS.
+Also, the higher is the FPS, the higher is the network traffic and server load,
+which may also degrade performance.
+%
+GALS detaches latency from FPS, since each node has its own timeline.
+However, the protocol requires two roundtrips to the server, which degrades
+performance considerably.
+The authors claim to achieve a responsiveness of 350ms in applications with
+25 nodes at 25 FPS, where 350ms represents the time it takes for all nodes to
+apply an event since its initial occurrence.
+%
+As we discuss in Section~\ref{sec.eval}, the performance of our midlleware is
+measured in terms of rollbacks, i.e., the more the peers need to resynchronize
+and travel back in time, the lower is the performance of the application.
+In a similar scenario with 21 peers at 50 FPS, and average of 5 hops with 50ms
+latency, we show 3.1\% of rollbacks with responsiveness of 500ms, which we
+classify as moderate performance (Figure~\ref{fig.baks}, framed rectangle in
+the center).
+As expected, the peer-to-peer performance is lower in comparison to centralized
+solutions, but still viable in many scenarios, as detalied in
+Section~\ref{sec.eval}.
+%
+Regarding CRDTs, recall that they only provide eventual consistency with no
+notion of a shared timeline in which peers go through identical steps.
+Therefore, measuring the performance of event responsiveness, maximum FPS, or
+clock synchronization is meaningless, given that events can be applied locally
+and broadcasted using a best-effort policy.
+
+\subsection{Time Machines}
+
+Regarding decentralized time machines, Fusion~\cite{fusion} is a state
 synchronization networking library for the Unity game engine.
@@
 %
-Regarding single-user local machines, the video game Braid%
+Regarding single-user local time machines, the video game Braid%
 \footnote{Braid: \url{http://braid-game.com/}}

### 2023-03-09 tpds-23/tpds-23.tex
...

--- 2022-11-29 final/p2p-22.tex
@@
-\documentclass[sn-mathphys,iicol]{sn-jnl}% Math and Physical Sciences Reference Style 
-%\documentclass[default,iicol]{sn-jnl}% Default with double column layout
-
-\jyear{2022}%
-
-\raggedbottom
-%%\unnumbered% uncomment this for unnumbered level heads
+%\documentclass[journal]{IEEEtran}
+\documentclass[10pt,journal,compsoc]{IEEEtran}
 
 \usepackage{verbatim}
+\usepackage{url}
 \usepackage{graphicx}
@@
 \usepackage{xspace}
-\newcommand{\FC}       {Freechains\xspace}
 \newcommand{\reps}     {\emph{reps}\xspace}
@@
 
-\title[Article Title]{
-    Symmetric Peer-to-Peer Applications
-}
-
-\author{\fnm{Francisco} \sur{Sant'Anna} \email{francisco@ime.uerj.br}}
-\affil{\orgdiv{Department of Computer Science}, \orgname{Rio de Janeiro State University}, \orgaddress{\country{Brazil}}}
-
-\abstract{
+\title{
+    A Middleware for Symmetric Peer-to-Peer Applications
+}
+
+\author{
+    Francisco Sant'Anna~\IEEEmembership{Department of Computer Science, Rio de Janeiro State University}
+}
+
+\IEEEtitleabstractindextext{%
+\begin{abstract}
 In real-time networked applications, such as collaborative documents, users can
@@
 %
-In this work, we propose a middleware for \emph{symmetric peer-to-peer
+In this work, we propose a middleware for these \emph{symmetric peer-to-peer
 applications} in which decentralized instances can broadcast asynchronous
@@
 %
-We show that the middleware can handle applications with over 20 peers in a
+We show that the middleware can handle applications with over 20 peers in
 heterogeneous topologies under high churns.
-For instance, in a simulation scenario with a network latency of 50ms and 25
-events per minute, the peers need to roll back and resynchronize only 3\% of
-the time.
-}
-
-\keywords{
-collaboration, determinism, peer-to-peer, time machine
-}
-
-%%\pacs[JEL Classification]{D8, H51}
-
-%%\pacs[MSC Classification]{35A01, 65L10, 65L12, 65L20, 65L70}
+For instance, in a scenario with a network latency of 50ms and 25 events per
+minute, the peers need to roll back only 3\% of the time.
+\end{abstract}
+
+\begin{IEEEkeywords}
+Bitcoin, blockchains, CRDT, distributed consensus, peer-to-peer,
+publish-subscribe, reputation system, VCS
+\end{IEEEkeywords}}
 
 \maketitle
+
+% TOTAL: 12 pages
 
@@
 \label{sec.introduction}
+
+ 6. ***How does the middleware handle consistency when advancing or rollbacking in  time to synchronize?***
+About #6, this is a valid criticism and we should include a brief discussion about consistency models and position our work on this spectrum.
+The facts that peers share a global clock and go through bit-identical steps may characterize strict consistency.
+However, since the global clocks may diverge temporarily, we must rely on rollbacks.
+The key to preserve consistency in the presence of rollbacks is to restrict our programming API to deterministic and stateless calls only.
+This way, it is not possible to diverge from other peers, for example, by re-executing stateful file operations multiple times (or at different times).
+
+&gt; A apresentação da solução entra em um nível de detalhamento muito baixo...
+  ...na decisão da topologia utilizada na simulação não fica muito claro se  essa topologia representa algum cenário específico.
+
+Sobre a topologia escolhida, optamos por misturar 3 arranjos abstratos:
+
+1. linha contínua
+2. ciclo fracamente conectado
+3. ciclo totalmente conectado
+
+Entendemos que casos concretos que sejam subconjuntos dos arranjos acima terão desempenho igual ou superior ao exibido na Figura 7.
+Por exemplo, uma linha contínua de até 15 nós é um subconjunto da topologia escolhida, assim como dois ciclos conectados separados por um nó.
+Assim, pretendemos atender a um número maior de cenários, mesmo que isso dependa de uma interpretação mais apurada por parte dos interessados.
+De qualquer maneira, a escolha é arbitrária e não representa um cenário concreto.
+
+&gt; Figures 1 and 2 create an expectation of approaches comparison; I suggest writing in the motivation why there will be no clear comparison in the
+  experimental results.
+We will be more explicit about this in Sections 2/4, since it is not possible to reproduce our experiment in centralized solutions (the topology is inherently peer-to-peer).
+We only discuss the performance of related works based on their claims and also on theoretical upper limits."
+
+Os autores mencionam que em cenário com latência de 50ms e 25 eventos por minuto, os pares precisaram retroceder 3\% do tempo.
+Eles devem deixar claro porque isso é uma vantagem e qual a média de período de tempo aceitável para retrocesso.
+- Nos trabalhos relacionados, apenas propostas centralizadas são mencionadas e comparadas. Não há propostas descentralizadas para comparação também? Sugiro expandir essa seção.
+- O design experimental precisa ser melhor descrito. Apenas o emulador das latências é descrito (NetEm). Como foi realizada (ou qual ferramenta foi usada) a simulação dos experimentos?
+- Apenas aplicações com deadline mais curtos adequam-se melhor ao middleware.  Os autores precisam explicar no texto os resultados dos experimentos com maior latência também (100ms, 250ms e 500ms).
+- Os autores precisam deixar claro o benefício de usar uma solução descentralizada desde que o seu desempenho é menor que as soluções centralizadas.
 
@@
 nodes and determine a shared timeline.
-
-In this work, we propose \emph{symmetric peer-to-peer applications}, which
-do not rely on central servers for coordination.
+Therefore, applications may experience failures, unless users are permanently
+connected, with no partitions in the network, particularly from the server.
+In contrast, a decentralized solution can be resilient to partitions, and
+ultimately, even work completely offline, as required by local-first
+software~\cite{local}.
+
+In this work, we propose a middleware for \emph{symmetric peer-to-peer
+applications}, which do not rely on central servers for coordination.
 Peers in the application form a dynamic network graph and communicate only
@@
 %
-As results, we show that the last two goals are met even under extreme
+As results, we show that goals (b) and (c) are met even under extreme
 conditions: peers exhibit a time mismatch of under $0.2\%$ and take less than
@@
 %
-For the first goal, we analyze the scenarios in which the middleware can meet
+For the goal (a), we analyze the scenarios in which the middleware can meet
 event deadlines with the desired performance.
-As an example, peers need to roll back only $3\%$ of the time under a scenario
-with a network latency of $50ms$ and 25 events per minute.
+As an example, peers need to roll back only $3\%$ of the time under a network
+latency of $50ms$ and 25 events per minute.
 
@@
     CRDTs, and this work.
-    The vertical arrows in parallel represent the timelines in nodes in the
-    network.
+    The vertical arrows in parallel represent the nodes timelines.
+    %in nodes in the network.
     The arrows crossing nodes represent communication between them.
@@
 In this section, we revisit existing solutions for symmetric distributed
-applications.
+applications, which are not necessarily peer to peer.
 We focus on
@@
 stateless calls only.
-
-Many aspects may affect the performance of symmetric distributed applications,
-such as
-    the network latency,
-    the number of nodes in the network,
-    the rate of user events,
-    and the application frame rate.
-%
-It is expected that centralized solutions exhibit a better performance in
-comparison to peer-to-peer applications, since nodes are only 2 hops aways from
-each other.
-%
-The performance of Croquet depends exclusively on the network latency, which
-restricts the application to a maximum FPS.
-For instance, since all frame ticks need to be transmitted over the network,
-a latency of 50ms supports at most 20 FPS.
-Also, the higher is the FPS, the higher is the network traffic and server load,
-which may also degrade performance.
-%
-GALS detaches latency from FPS, since each node has its own timeline.
-However, the protocol requires two roundtrips to the server, which degrades
-performance considerably.
-The authors claim to achieve a responsiveness of 350ms in applications with
-25 nodes at 25 FPS, where 350ms represents the time it takes for all nodes to
-apply an event since its initial occurrence.
-%
-As we discuss in Section~\ref{sec.eval}, the performance of our midlleware is
-measured in terms of rollbacks, i.e., the more the peers need to resynchronize
-and travel back in time, the lower is the performance of the application.
-In a similar scenario with 21 peers at 50 FPS, and average of 5 hops with 50ms
-latency, we show 3.1\% of rollbacks with responsiveness of 500ms, which we
-classify as moderate performance (Figure~\ref{fig.baks}, framed rectangle in
-the center).
-As expected, the peer-to-peer performance is lower in comparison to centralized
-solutions, but still viable in many scenarios, as detalied in
-Section~\ref{sec.eval}.
-%
-Regarding CRDTs, recall that they only provide eventual consistency with no
-notion of a shared timeline in which peers go through identical steps.
-Therefore, measuring the performance of event responsiveness, maximum FPS, or
-clock synchronization is meaningless, given that events can be applied locally
-and broadcasted using a best-effort policy.
 
@@
 
-In Figure~\ref{fig.middleware}, a peer is depicted as an application that uses
-an API to communicate with the middleware.
+In Figure~\ref{fig.middleware}, a peer is depicted as an application using an
+API to communicate with the middleware.
 The application itself is not aware of the network and has no direct access to
@@
 The middleware assumes that each peer assigns itself a unique identifier (e.g.,
-a combination its MAC \& IP addresses).
+a combination of its MAC \& IP addresses).
 %, such as TODO~\cite{TODO:hash}.
@@
 
-Note that the use of callbacks is a common technique for event-driven
+Note that using callbacks is a common technique for event-driven
 programming in Games, GUIs, and interactive applications in general, sharing
-recognition with related techniques such as \emph{MVC},
-\emph{Publish-Subscribe}, and \emph{Observer} patterns~\cite{meyer,nystrom}.
+similarities with \emph{MVC}, \emph{Publish-Subscribe}, and \emph{Observer}
+patterns~\cite{meyer,nystrom}.
 %
@@
 \textbf{(b) Real-time pace of peers:}
-All peers output their ticks continuously, allowing us to compare them in the
-shared timeline.
+All peers output their ticks continuously, which we compare in the shared
+timeline.
 Every time we see a greater tick never seen in the timeline before, we start to
@@
 $ lua5.3 test2-sheet.lua
+$
 \end{comment}
@@
 Our results show that goals (b) and (c) are met in all scenarios consistently,
-and we consider them nearly optimal:
+which are nearly optimal:
 Regardless of the network latency, event rate, and delay multiplier,
@@
 
+\subsection{Discussion}
+
+local-first
+Apesar dos rollbacks quando a rede se reconecta serem inevitáveis, os seus efeitos podem ser imperceptíveis para os usuários finais (e.g., quando há pouca
+atividade ou dependência entre eventos).
+
+O principal benefício é que a aplicação continua funcionando mesmo com a
+presença de partições na rede, o que inclui a queda do (que seria o) servidor
+em uma solução centralizada.
+
+Levando essa possibilidade ao extremo, um peer ainda poderia funcionar
+normalmente mesmo que totalmente offline por um determinado período (como é
+comum em pesquisas recentes em &quot;local-first software&quot;).
+
+Many aspects may affect the performance of symmetric distributed applications,
+such as
+    the network latency,
+    the number of nodes in the network,
+    the rate of user events,
+    and the application frame rate.
+%
+It is expected that centralized solutions exhibit a better performance in
+comparison to peer-to-peer applications, since nodes are only 2 hops away from
+each other.
+%
+However, the main benefit of our work is that applications remain functional
+even under network partitions, which includes failures in (would be) servers.
+Taking this possibility to an extreme, a peer can still work normally even
+completely offline for a period of time, as desired by local-first
+software~\cite{local}:
+Even though rollbacks are inevitable when reconnecting, their effects may be
+unnoticeable for final users if there is low activity or dependency between
+events.
+%
+As pointed out in column (c) of Figure~\ref{fig.table}, centralized solutions
+require that clients are permanently connected with the server, which controls
+the shared timeline of events.
+
+The performance of Croquet~\cite{croquet} depends exclusively on the network
+latency, which restricts the application to a maximum FPS.
+For instance, since all frame ticks need to be transmitted over the network,
+a latency of 50ms supports at most 20 FPS.
+Also, the higher is the FPS, the higher is the network traffic and server load,
+which may also degrade performance.
+%
+GALS~\cite{gals} detaches latency from FPS, since each node has its own
+timeline.
+However, the protocol requires two round trips to the server, which degrades
+performance considerably.
+The authors claim to achieve a responsiveness of 350ms in applications with
+25 nodes at 25 FPS, where 350ms represents the time it takes for all nodes to
+apply an event since its initial occurrence.
+
+The performance of our middleware is measured in terms of rollbacks, i.e., the
+more the peers need to resynchronize and travel back in time, the lower is the
+performance of the application.
+In a similar scenario with 21 peers at 50 FPS, and average of 5 hops with 50ms
+latency, we show 3.1\% of rollbacks with responsiveness of 500ms, which we
+classify as moderate performance (Figure~\ref{fig.baks}, framed rectangle in
+the center).
+As expected, the peer-to-peer performance is lower in comparison to centralized
+solutions, but still viable in many scenarios.
+
+Regarding CRDTs, recall that they only provide eventual consistency with no
+notion of a shared timeline in which peers go through identical steps.
+Therefore, measuring the performance of event responsiveness, maximum FPS, or
+clock synchronization is meaningless, given that events can be applied locally
+and broadcast using a best-effort policy.
+
 %\section{Application}
@@
 
-The middleware can handle applications with over 20 peers in a heterogeneous
+The middleware can handle applications with over 20 peers in heterogeneous
 topologies and under high churns.
@@
 
-\section*{Declarations}
-
-\subsection*{Ethics Approval}
-
-Not applicable
-
-\subsection*{Conflict of Interest}
-
-The author declares no conflicts of interest.
-
-\subsection*{Data Availability}
-
-Not applicable
-
-\subsection*{Author Contribution}
-
-Francisco Sant'Anna is the single author and wrote the full manuscript.
-
-\subsection*{Funding}
-
-This study was partially funded by FAPERJ (grant E-26/211.329/2019 (251640)).
-
-\subsection*{Consent to publish}
-
-The author hereby expresses consent to publish.
-
-\bibliography{p2p-22}
+\bibliographystyle{IEEEtran}
+\bibliography{tpds-23}
+
+%\begin{comment}
+\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{chico}}]{Francisco Sant'Anna}
+received his PhD degree in Computer Science from PUC-Rio, Brazil in
+2013.
+In 2016, he joined the Faculty of Computer Science at the Rio de Janeiro State
+University, Brazil.
+His research interests include Programming Languages and Concurrent \&
+Distributed Systems.
+\end{IEEEbiography}
+%\end{comment}
 

### 2023-03-10 tpds-23/tpds-23.tex
...

@@
 
- 6. ***How does the middleware handle consistency when advancing or rollbacking in  time to synchronize?***
-About #6, this is a valid criticism and we should include a brief discussion about consistency models and position our work on this spectrum.
-The facts that peers share a global clock and go through bit-identical steps may characterize strict consistency.
-However, since the global clocks may diverge temporarily, we must rely on rollbacks.
-The key to preserve consistency in the presence of rollbacks is to restrict our programming API to deterministic and stateless calls only.
-This way, it is not possible to diverge from other peers, for example, by re-executing stateful file operations multiple times (or at different times).
-
-&gt; A apresentação da solução entra em um nível de detalhamento muito baixo...
-  ...na decisão da topologia utilizada na simulação não fica muito claro se  essa topologia representa algum cenário específico.
-
-Sobre a topologia escolhida, optamos por misturar 3 arranjos abstratos:
-
-1. linha contínua
-2. ciclo fracamente conectado
-3. ciclo totalmente conectado
-
-Entendemos que casos concretos que sejam subconjuntos dos arranjos acima terão desempenho igual ou superior ao exibido na Figura 7.
-Por exemplo, uma linha contínua de até 15 nós é um subconjunto da topologia escolhida, assim como dois ciclos conectados separados por um nó.
-Assim, pretendemos atender a um número maior de cenários, mesmo que isso dependa de uma interpretação mais apurada por parte dos interessados.
-De qualquer maneira, a escolha é arbitrária e não representa um cenário concreto.
-
-&gt; Figures 1 and 2 create an expectation of approaches comparison; I suggest writing in the motivation why there will be no clear comparison in the
-  experimental results.
-We will be more explicit about this in Sections 2/4, since it is not possible to reproduce our experiment in centralized solutions (the topology is inherently peer-to-peer).
-We only discuss the performance of related works based on their claims and also on theoretical upper limits."
-
-Os autores mencionam que em cenário com latência de 50ms e 25 eventos por minuto, os pares precisaram retroceder 3\% do tempo.
-Eles devem deixar claro porque isso é uma vantagem e qual a média de período de tempo aceitável para retrocesso.
-- Nos trabalhos relacionados, apenas propostas centralizadas são mencionadas e comparadas. Não há propostas descentralizadas para comparação também? Sugiro expandir essa seção.
-- O design experimental precisa ser melhor descrito. Apenas o emulador das latências é descrito (NetEm). Como foi realizada (ou qual ferramenta foi usada) a simulação dos experimentos?
-- Apenas aplicações com deadline mais curtos adequam-se melhor ao middleware.  Os autores precisam explicar no texto os resultados dos experimentos com maior latência também (100ms, 250ms e 500ms).
-- Os autores precisam deixar claro o benefício de usar uma solução descentralizada desde que o seu desempenho é menor que as soluções centralizadas.
+\begin{comment}
+Os autores mencionam que em cenário com latência de 50ms e 25 eventos por
+minuto, os pares precisaram retroceder 3\% do tempo.
+Eles devem deixar claro porque isso é uma vantagem e qual a média de período de
+tempo aceitável para retrocesso.
+
+- O design experimental precisa ser melhor descrito.
+Apenas o emulador das latências é descrito (NetEm).
+Como foi realizada (ou qual ferramenta foi usada) a simulação dos experimentos?
+
+- Apenas aplicações com deadline mais curtos adequam-se melhor ao middleware.
+Os autores precisam explicar no texto os resultados dos experimentos com maior
+latência também (100ms, 250ms e 500ms).
+\end{comment}
 
@@
 To deal with events received out of order or too late, a distributed time
-machine can rollback peers to a previous state and reapply events in order and
-in time.
+machine can rollback peers to previous states and reapply the events in order,
+such that a single consistent timeline is established across the network.
 Our main contribution is the design of a middleware to ensure that all peers
@@
 
+In this work, as indicated in Figure~\ref{fig.table}, our goal is to support
+peer-to-peer (a) real-time symmetric execution, while still tolerating
+out-of-order (b) and offline (c) event generation.
+%
+As detailed in Figure~\ref{fig.algos}, peers have independent timelines and
+can communicate events directly to each other.
+Event deadlines are delayed to be able to reach other peers in time.
+In the case a deadline is missed, the peer rolls back in time and then fast
+forwards execution until it catches real-time behavior again.
+%
+Like Croquet and GALS, peers compute events in sequence using deterministic and
+stateless calls only.
+
+\subsection{Conflict-free Replicated Data Types}
+
 An antagonistic approach to deterministic reactions to events is to model the
@@
 is an example of decentralized CRDTs~\cite{p2p.automerge,p2p.pushpin}.
-
-In this work, as indicated in Figure~\ref{fig.table}, our goal is to support
-peer-to-peer (a) real-time symmetric execution, while still tolerating
-out-of-order (b) and offline (c) event generation.
-%
-As detailed in Figure~\ref{fig.algos}, peers have independent timelines and
-can communicate events directly to each other.
-Event deadlines are delayed to be able to reach other peers in time.
-In the case a deadline is missed, the peer rolls back in time and then fast
-forwards execution until it catches real-time behavior again.
-%
-Like Croquet and GALS, peers compute events in sequence using deterministic and
-stateless calls only.
 
@@
 The main job of the middleware is to deal with the uncertainties of the
-network in order to preserve the symmetric behavior across peers.
+network in order to provide a consistent timeline and preserve the symmetric
+behavior across peers.
 As described in the Introduction, the middleware ensures that all peers
@@
 
-This section provides an overview of the middleware implementation to answer
+This section provided an overview of the middleware implementation to answer
 how it ensures the desired properties in concrete terms as follows:
@@
 
+The fact that peers share a global timeline of event, and that they go through
+bit-identical steps ensures consistency in the network.
+However, since the clocks may diverge temporarily, we must rely on rollbacks.
+Therefore, the key to preserve consistency in the presence of rollbacks is to
+restrict our programming API to deterministic and stateless calls only.
+This way, it becomes impossible to diverge from other peers by, for example,
+re-executing stateful file operations multiple times (or at different times).
+
 %\newpage
@@
 hops away).
-%
+We opted to mix three abstract arrangements:
+    (i) straight lines, and (ii) weakly and (iii) strongly connected cycles.
+Even though this topology is arbitrary, we can expect that concrete scenarios
+with subsets of these arrangements will have at least the same performance of
+our experiments.
+For instance, a straight line of 15 peers is a subset of our topology, as well
+as two cycles separated by a single peer.
+
 We run the application main loop at $50~FPS$ ($20ms$ per frame) and calculate
@@
 \subsection{Discussion}
-
-local-first
-Apesar dos rollbacks quando a rede se reconecta serem inevitáveis, os seus efeitos podem ser imperceptíveis para os usuários finais (e.g., quando há pouca
-atividade ou dependência entre eventos).
-
-O principal benefício é que a aplicação continua funcionando mesmo com a
-presença de partições na rede, o que inclui a queda do (que seria o) servidor
-em uma solução centralizada.
-
-Levando essa possibilidade ao extremo, um peer ainda poderia funcionar
-normalmente mesmo que totalmente offline por um determinado período (como é
-comum em pesquisas recentes em &quot;local-first software&quot;).
 

### 2023-03-10 tpds-23/tpds-23.tex
...

@@
 \label{sec.introduction}
-
-\begin{comment}
-Os autores mencionam que em cenário com latência de 50ms e 25 eventos por
-minuto, os pares precisaram retroceder 3\% do tempo.
-Eles devem deixar claro porque isso é uma vantagem e qual a média de período de
-tempo aceitável para retrocesso.
-
-- O design experimental precisa ser melhor descrito.
-Apenas o emulador das latências é descrito (NetEm).
-Como foi realizada (ou qual ferramenta foi usada) a simulação dos experimentos?
-
-- Apenas aplicações com deadline mais curtos adequam-se melhor ao middleware.
-Os autores precisam explicar no texto os resultados dos experimentos com maior
-latência também (100ms, 250ms e 500ms).
-\end{comment}
 
@@
 latency of $50ms$ and 25 events per minute.
+%
+We consider this overhead negligible for soft real-time applications that do
+not depend on instantaneous network feedback, such as chats, watch parties, and
+turn-based games.
 
@@
 enum {
-    P2P_EVT_KEY = P2P_EVT_NEXT // key is next to predefs
+    P2P_EVT_KEY = P2P_EVT_NEXT // KEY after predefs
 };
@@
 
-We run the application main loop at $50~FPS$ ($20ms$ per frame) and calculate
+\begin{comment}
+- O design experimental precisa ser melhor descrito.
+Apenas o emulador das latências é descrito (NetEm).
+Como foi realizada (ou qual ferramenta foi usada) a simulação dos experimentos?
+\end{comment}
+
+We run the application main loop at 50 FPS ($20ms$ per frame) and calculate
 the middleware overhead at each iteration.
@@
     \label{fig.baks}
+The squares show the percentage of time travels.
+%
 The $5x$, $2x$, and $x$ $\Delta$ regions each contains $36$ measures for
@@
 As an application example, chats are not affected by higher event deadlines.
-Nevertheless, the higher is the deadline, the less applications suit the
+Nevertheless, the lower is the deadline, the less applications suit the
 middleware.
@@
 either handle low latencies with high event frequencies (e.g., $5ms$ with
-$100evt/min$) or high lantencies with low event frequencies (e.g., $100ms$ with
+$100evt/min$) or high latencies with low event frequencies (e.g., $100ms$ with
 $5evt/min$).
+%
+The middleware does not perform well under higher network latencies (e.g., over
+$250ms$).
+This is expected since peers that are multiple hops away take too long to
+synchronize past events and may generate future events in the meantime, which
+causes time travels.
 %

### 2023-03-12 tpds-23/tpds-23.tex
...

@@
 %
-In this work, we propose a middleware for these \emph{symmetric peer-to-peer
+In this work, we propose a middleware for \emph{symmetric peer-to-peer
 applications} in which decentralized instances can broadcast asynchronous
@@
 For instance, in a scenario with a network latency of 50ms and 25 events per
-minute, the peers need to roll back only 3\% of the time.
+minute, the peers need to roll back only 3\% of the time to remain fully
+symmetric.
 \end{abstract}
@@
 \begin{IEEEkeywords}
-Bitcoin, blockchains, CRDT, distributed consensus, peer-to-peer,
-publish-subscribe, reputation system, VCS
+collaboration, determinism, peer-to-peer, time machine
 \end{IEEEkeywords}}
@@
 event occurrences must somehow be synchronized back across devices.
+%
+This problem relates to consistency in distributed systems, and is also known
+as \emph{state melding}~\cite{melding} in the context of virtual worlds, which
+adds the requirement of a time deadline to reach consistency.
 
@@
 However, current solutions depend on a central server to interconnect network
-nodes and determine a shared timeline.
+nodes and determine the shared timeline.
 Therefore, applications may experience failures, unless users are permanently
 connected, with no partitions in the network, particularly from the server.
-In contrast, a decentralized solution can be resilient to partitions, and
-ultimately, even work completely offline, as required by local-first
-software~\cite{local}.
+In contrast, a decentralized solution could be resilient to partitions, and
+ultimately, even work offline as promoted by local-first software~\cite{local}.
 
@@
 networks~\cite{p2p.survey}.
+%
 Events are flooded in the graph and are applied with a small deadline delay to
 compensate the network latency.
-To deal with events received out of order or too late, a distributed time
-machine can rollback peers to previous states and reapply the events in order,
-such that a single consistent timeline is established across the network.
+To deal with events received out of order or too late, a time machine can
+rollback a peer to a previous state and reapply the events in order, such that
+a single consistent timeline is established across the network.
+%
 Our main contribution is the design of a middleware to ensure that all peers
@@
     (c) can leave and join the network and remain symmetric.
+%
+Our middleware provides \emph{perceptive consistency}~\cite{melding}, which
+ensures that an event triggered by any peer at time $t$ is perceived by all
+peers simultaneously at time $t+\Delta$.
+%
+As the main limitation, the middleware cannot ensure \emph{instantaneity} for
+events in most scenarios, which would require $\Delta$ to be less than a
+perceptive threshold~\cite{melding}.
+We also that assume peers are non malicious in the sense that they do not lie
+about the time in which events occurred locally.
 
@@
     (c) how events are propagated and applied.
-Figure~\ref{fig.table} compares three selected works regarding these aspects.
+Figure~\ref{fig.table} compares three selected works regarding these aspects
+against our work.
+%
+One of these approaches relies on CRDTs~\cite{crdts}, and is discussed in
+separate.
 %
@@
 As detailed in Figure~\ref{fig.algos}, a centralized \emph{reflector} issues
-periodic ticks, such that all nodes in the network advance together according
-to a synchronized global shared clock.
+fixed periodic ticks (e.g., every $50ms$), such that all nodes in the network
+advance together according to a synchronized global shared clock.
 If a user generates an event, it is sent back to the reflector, which
 broadcasts the event in the next tick, which all nodes apply in sync.
+%
+This mechanism is similar to \emph{bucket synchronization}~\cite{gautier}, in
+which events are assigned to time buckets to be applied later.
 %
@@
 deterministic and stateless calls.
-This way, it is guaranteed that all clients reach the same state, and
+This way, it is guaranteed that all clients reach the exact same state, and
 observe bit-identical steps.
@@
 In this work, as indicated in Figure~\ref{fig.table}, our goal is to support
-peer-to-peer (a) real-time symmetric execution, while still tolerating
-out-of-order (b) and offline (c) event generation.
+    (a) peer-to-peer real-time symmetric execution, while still
+    (b) tolerating out-of-order, and
+    (c) offline event generation.
 %
@@
 %
+This technique is also known as \emph{log-based
+rollback-recovery}~\cite{rollback}, but applied in the context of real-time
+collaborative applications.
+%
 Like Croquet and GALS, peers compute events in sequence using deterministic and
 stateless calls only.
+
+As it is the case for the discussed work, we consider that middlewares for
+symmetric applications, which provide bit-identical behavior, must comply with
+\emph{perceptive consistency}~\cite{melding}, as follows:
+
+``\textbf{Perceptive consistency}~\cite{qin,bouillot}.
+If any operation is executed on process $i$ at time $t$, it must be executed on
+process $j$ at $t$ as well.
+At a high-level point of view, it indicates that for a group of users
+interacting with each other, if a user takes an action, then the action will be
+perceived by everyone simultaneously (though the action might be perceived at a
+time later than the time at which it was taken by the user). It is also
+referred to as absolute consistency~\cite{qin}.''
 
@@
 %
-As indicated in Figure~\ref{fig.table}, CRDTs work on decentralized networks
-(a), in which peers advance independently (b), and in which operations can be
-applied immediately, even while offline (c).
-%
-Automerge along with its accompanying Hypermerge peer-to-peer infrastructure
-is an example of decentralized CRDTs~\cite{p2p.automerge,p2p.pushpin}.
+As indicated in Figure~\ref{fig.table}, CRDTs work on
+    (a) decentralized networks, in which
+    (b) peers advance independently, and in which
+    (c) operations can be applied immediately, even while offline.
+%
+Automerge~\cite{p2p.automerge} along with its accompanying
+Hypermerge~\cite{p2p.pushpin} peer-to-peer infrastructure is an example of
+decentralized CRDTs.
+
+CRDTs provide \emph{delayed consistency}~\cite{melding}, as follows:
+
+``\textbf{Delayed consistency}~\cite{qin}.
+This is a relaxation of perceptive consistency.
+In this model, for any operation issued by host $i$, say at time $t$, it takes
+effect instantaneously on host $i$.
+On a remote host $j$, it is executed at time $t + (i,j)$, where $(i,j)$ is the
+phase different between processes $i$ and $j$, and it is a parameter specific
+to these two processes.
+Unlike perceptive consistency, this model does not force all processes to
+execute the same operation at the same time.
+Rather, it makes sure that the state of objects operated by one process be
+viewed at another process with a consistent delay.''
 
@@
 
-Regarding decentralized time machines, Fusion~\cite{fusion} is a state
-synchronization networking library for the Unity game engine.
+Regarding decentralized time machines, the \emph{Time Warp}
+mechanism~\cite{timewarp} rollbacks the memory state of a replica when it
+receives an event from the past, and also broadcasts ``anti-messages'' for all
+events that were incorrectly applied.
+Our work assumes that all events are correct, and reapplies them relying on
+deterministic simulation.
+
+Fusion~\cite{fusion} is a state synchronization networking library for the
+Unity game engine.
 It provides networked tick-based simulation in two modes: client-server
@@
 central server.
-%
+
 Regarding single-user local time machines, the video game Braid%
@@
     \caption[XXX] {
+        \textbf{An application example:}
         A moving rectangle is controlled by peers connected indirectly.
@@
 very same position.
+%
+This is only possible because the middleware ensures that all peers shares the
+same timeline of events, in which only deterministic operations are applied.
 
@@
   \caption{
+    \textbf{The middleware architecture:}
     Applications generate asynchronous events and communicate through an API
@@
 which synchronizes them with a timestamp with a deadline delay ($\Delta$)
-scheduled to the future, such that all peers can satisfy.
+scheduled to the future, such that all peers can satisfy (also known as
+\emph{local lag}~\cite{mauve}]).
 The middleware controls the execution of the application by issuing ticks and
@@
 \item \textbf{Event Deadlines:}
-Due to the inherent latency of networks, peers generate events with a
-deliberate $\Delta$ so that they can reach other peers in time to be applied in
-sync.
+Due to the inherent latency of networks, a peer (through the middleware)
+generates events with a deliberate $\Delta$ so that they can reach other peers
+in time to be applied in sync.
 %We first assume a predetermined delay for the sake of simplicity.
 Nevertheless, a peer ahead of time may receive an event that should have been
-applied in the past. %, even considering its $\Delta$.
+applied in the past, which characterizes an optimistic approach to choose
+$\Delta$, since it assumes that inconsistencies may occur~\cite{melding}.
+%, even considering its $\Delta$.
+%
 In this case, the middleware automatically rolls back the peer, applies the
@@
 
+As a limitation of our work, regarding items (a) and (b), we assume that peers
+never lie about their local clocks, which would affect the event timestamps,
+allowing a peer to modify the global timeline with nonsensical information.
+%
+For this reason, we assume a permissioned peer-to-peer network with an external
+mechanism for authentication.
+
 \subsection{The Programming API}
@@
 
-Note that using callbacks is a common technique for event-driven
-programming in Games, GUIs, and interactive applications in general, sharing
-similarities with \emph{MVC}, \emph{Publish-Subscribe}, and \emph{Observer}
-patterns~\cite{meyer,nystrom}.
+Note that using callbacks is a already common technique for event-driven
+programming, such as in Games, GUIs, and interactive applications in general,
+sharing similarities with \emph{MVC}, \emph{Publish-Subscribe}, and
+\emph{Observer} patterns~\cite{meyer,nystrom}.
 %
@@
 meta-programming offer better mechanisms to take memory snapshots.
-Another alternative is to provide a serialization callback, which we left for
-future work.
+Another alternative is to provide a serialization callback, which we left out
+for future work.
 
@@
 static immutable globals that can live outside the simulation memory, such as
-the window and image textures in our example:
+the window and image textures of our example:
 
@@
 
+The simulation callback is the most sensitive API of the proposed middleware,
+since it is responsible for preserving the consistency of the global
+application state \code{G}, even under multiple executions from scratch.
+%
+Since the code is written in C, which is unaware of the semantics of symmetric
+applications, note that it is the responsibility of the programmer to ensure
+that the state is completely reproducible.
+%
+As discussed in Section~\ref{sec.related}, the typical approach, which we
+chose, is to rely on deterministic and stateless calls only.
+
 \subsubsection{Rendering Callback}
@@
 
-Time travelling is also the reason why \code{cb\_sim} must be separated from
-the rendering callback \code{cb\_ren}~\cite{tml.js}, which would otherwise
-render the screen multiple times while travelling.
-The middleware renders the screen after each frame in real time.
+The restriction for stateless calls is the reason why the rendering callback
+\code{cb\_ren}~\cite{tml.js} must be separated from \code{cb\_sim}, which would
+otherwise render the screen multiple times while time travelling.
+Instead, the middleware calls \code{cb\_ren} to render the screen after each
+frame event, but only in real time.
 %, which the middleware calls only on real-time frames.
@@
 
-The callback \code{cb\_evt} is called by the middleware in real time, whenever
-a local SDL event occurs.
+The callback \code{cb\_evt} is invoked by the middleware in only real time,
+whenever a local SDL event occurs.
 This callback has two goals:
-    (a) map and broadcast application events, and
-    (b) provide instant feedback to the user.
+    (a) to map and broadcast application events, and
+    (b) to provide instant feedback to the user.
 %As mentioned previously,
-Not all local low-level events need to broadcast application events in the
-network.
+Not all local low-level events need to be broadcast in the network, only those
+mapped as application events, as described in Section~\ref{sec.tml.api.events}.
 The callback is allowed to perform side effects, such as modifying the network
 topology, or exhibiting instant feedback on the screen.
-In our example, we only generate application events for the 5 key events of
-interest, but we also exhibit the pressed key in real time as a visual
+In our example, we only generate application events for the 5 key of interest,
+but we also exhibit the pressed key in real time as a visual
 feedback:
@@
 
-We now detail how the middleware orchestrates an application and ensures that
-all peers
+We now detail how the middleware orchestrates an application to ensure that all
+peers
     (a) meet event deadlines,
@@
 %
-To prevent broadcast cycles, each peer keeps a collection of the highest
-timestamps it received from each other peer, such that lower numbers can be
-ignored when received.
+To prevent broadcast cycles, each peer keeps a set of the highest timestamps it
+received from each other peer, such that lower numbers can be ignored when
+received.
 Since we rely on TCP connections, it is not possible to receive out-of-order
@@
 
-The middleware main loop \code{p2p\_loop} is responsible for calling the
+The middleware main loop \code{p2p\_loop} is responsible for invoking the
 application callbacks, and also controlling its timeline.
@@
 
-The fact that peers share a global timeline of event, and that they go through
-bit-identical steps ensures consistency in the network.
-However, since the clocks may diverge temporarily, we must rely on rollbacks.
+Two facts ensure consistency in the network:
+    (a) peers share a global timeline of events, and
+    (b) they go through bit-identical steps
+%
+However, since the clocks may diverge temporarily, we must rely on simulation
+rollbacks.
 Therefore, the key to preserve consistency in the presence of rollbacks is to
@@
 less than $1/2000$ of the CPU time.
+This means that an application using the middleware would perform all network
+synchronization and still have almost 100\% of the CPU time available for its
+actual domain logic.
 %
@@
     and the application frame rate.
-%
+
 It is expected that centralized solutions exhibit a better performance in
@@
 events.
-%
+
 As pointed out in column (c) of Figure~\ref{fig.table}, centralized solutions
-require that clients are permanently connected with the server, which controls
-the shared timeline of events.
+require that clients are permanently connected with the server, which is
+responsible to control the shared timeline of events.
 
@@
 %
+%To our knowledge, it is the first middleware to support bit-identical real-time
+%behavior in decentralized peer-to-peer networks.
+%
 Our main contribution is a time machine that supports rollbacks based on memory
@@
 %unique identifiers.
+%
+The middleware provides \emph{perceptive consistency}~\cite{melding}, which
+adds a time deadline to reach consistency, ensuring that an event triggered by
+any peer at time $t$ is perceived by all peers simultaneously at time
+$t+\Delta$.
 

### 2023-03-12 tpds-23/tpds-23.tex
...

@@
 The full source code%
-\footnote{Code: \url{http://github.com/TODO-hidden-blind-review}}
+\footnote{Code: \url{http://github.com/fsantanna/p2p-tml/}}
 is under 500 lines of code.
@@
         mirrors of a single application.
-        Video: \url{http://youtube.com/TODO-hidden-blind-review}
+        Video: \url{https://youtu.be/zkl0vSOGino}
         \label{fig.move}
@@
 $5evt/min$).
+Note that the middleware cannot ensure \emph{instantaneity} for events in most
+scenarios, which would require a very low $\Delta$~\cite{melding}.
 %
@@
 
-%\begin{comment}
+\begin{comment}
 \begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{chico}}]{Francisco Sant'Anna}

### 2023-03-12 tpds-23/tpds-23.tex
...

@@
 A common approach towards symmetric applications is to introduce a middleware
-to orchestrate events and time in the network~\cite{gals,croquet}.
+to orchestrate events and time in the network~\cite{gals,croquet2}.
 This way, applications developers can rely on middleware primitives to trigger
@@
 events in most scenarios, which would require $\Delta$ to be less than a
-perceptive threshold~\cite{melding}.
+user-observable threshold~\cite{melding}.
 We also that assume peers are non malicious in the sense that they do not lie
@@
 
-Croquet~\cite{croquet,croquet.site} guarantees bit-identical real-time behavior
-for users in collaborative distributed environments.
+Croquet~\cite{croquet,croquet2,croquet.site} guarantees bit-identical real-time
+behavior for users in collaborative distributed environments.
 %
@@
 decentralized CRDTs.
+
+The main advantage of CRDTs is that they can be designed in such a way that
+concurrent modifications to shared state can preserve intent~\cite{peritext},
+regardless of the order in which they are applied.
+%
+This is not the case for the symmetric application middlewares described in the
+previous section.
+Nevertheless, programmers can still use CRDTs on top of a middleware, since
+they are nothing more than data structures and merging algorithms.
 
@@
 
-The performance of Croquet~\cite{croquet} depends exclusively on the network
+The performance of Croquet~\cite{croquet2} depends exclusively on the network
 latency, which restricts the application to a maximum FPS.
@@
 The middleware API shares similar limitations with centralized
-alternatives~\cite{gals,croquet}, being restricted to deterministic and
+alternatives~\cite{gals,croquet2}, being restricted to deterministic and
 stateless calls. %, and only supports pre-allocated memory.

### 2023-03-17 tpds-23/tpds-23.tex
...

@@
 %
-This problem relates to consistency in distributed systems, and is also known
-as \emph{state melding}~\cite{melding} in the context of virtual worlds, which
-adds the requirement of a time deadline to reach consistency.
+This problem relates to consistency in distributed systems, but with the
+additional requirement of a time deadline to reach consistency~\cite{melding}.
 
@@
 %
-Our middleware provides \emph{perceptive consistency}~\cite{melding}, which
-ensures that an event triggered by any peer at time $t$ is perceived by all
-peers simultaneously at time $t+\Delta$.
+Our middleware ensures \emph{perceptive consistency}~\cite{melding}, in which
+an event triggered by any peer at time $t$ is perceived by all peers
+simultaneously at time $t+\Delta$.
 %
@@
 user-observable threshold~\cite{melding}.
-We also that assume peers are non malicious in the sense that they do not lie
+We also assume that peers are non malicious in the sense that they do not lie
 about the time in which events occurred locally.
@@
 As an example, peers need to roll back only $3\%$ of the time under a network
-latency of $50ms$ and 25 events per minute.
+latency of 50ms and 25 events per minute.
 %
@@
 As detailed in Figure~\ref{fig.algos}, a centralized \emph{reflector} issues
-fixed periodic ticks (e.g., every $50ms$), such that all nodes in the network
+fixed periodic ticks (e.g., every 50ms), such that all nodes in the network
 advance together according to a synchronized global shared clock.
@@
 
-As it is the case for the discussed work, we consider that middlewares for
-symmetric applications, which provide bit-identical behavior, must comply with
-\emph{perceptive consistency}~\cite{melding}, as follows:
+As it is the case for the discussed work in this section, we consider that
+middlewares for symmetric applications must comply with \emph{perceptive
+consistency}~\cite{melding}, as follows:
 
@@
 %
-Automerge~\cite{p2p.automerge} along with its accompanying
-Hypermerge~\cite{p2p.pushpin} peer-to-peer infrastructure is an example of
-decentralized CRDTs.
+Automerge~\cite{p2p.automerge} along with its accompanying Hypermerge
+peer-to-peer infrastructure~\cite{p2p.pushpin} is an example of decentralized
+CRDTs.
 
@@
 previous section.
-Nevertheless, programmers can still use CRDTs on top of a middleware, since
-they are nothing more than data structures and merging algorithms.
-
-CRDTs provide \emph{delayed consistency}~\cite{melding}, as follows:
+Nevertheless, programmers could still use CRDTs on top of a middleware to
+preserve intent, since they are nothing more than data structures and merging
+algorithms.
+
+Contrasting with perceptive consistency, CRDTs provide \emph{delayed
+consistency}~\cite{melding}, as follows:
 
@@
 
-Regarding single-user local time machines, the video game Braid%
+Regarding local single-user time machines, the video game Braid%
 \footnote{Braid: \url{http://braid-game.com/}}
@@
 \footnote{Code: \url{http://github.com/fsantanna/p2p-tml/}}
-is under 500 lines of code.
+is around 500 lines of code.
 
@@
 which synchronizes them with a timestamp with a deadline delay ($\Delta$)
-scheduled to the future, such that all peers can satisfy (also known as
-\emph{local lag}~\cite{mauve}]).
+scheduled to the future (aka \emph{local lag}~\cite{mauve}]), such that all
+peers can satisfy.
 The middleware controls the execution of the application by issuing ticks and
@@
 the middleware can handle arrival and departure of peers seamlessly.
-Note that high churn may lead to intermittent network partitions.
+Note that high churns may lead to intermittent network partitions.
 Given our unstructured approach, nothing needs to be done when a peer leaves
@@
 As a limitation of our work, regarding items (a) and (b), we assume that peers
-never lie about their local clocks, which would affect the event timestamps,
+never lie about their local clocks, which would affect event timestamps,
 allowing a peer to modify the global timeline with nonsensical information.
 %
-For this reason, we assume a permissioned peer-to-peer network with an external
+Therefore, we assume a permissioned peer-to-peer network with an external
 mechanism for authentication.
@@
 
-The middleware expects to take full control of the application execution in
-order to manipulate its timeline and disseminate events in the network.
+The middleware expects to take full control of the application in order to
+manipulate its timeline and disseminate events in the network.
 For this reason, the basic programming API is to call a \code{loop} function
@@
 
-Note that using callbacks is a already common technique for event-driven
+Note that using callbacks is already a common technique for event-driven
 programming, such as in Games, GUIs, and interactive applications in general,
@@
 Occurring events are passed as arguments to the callback, which must never
-perform any side effects, such as stateful calls or rendering frames.
+perform any side effects (outside \code{G}), such as stateful calls or
+rendering frames.
 In our example, we modify the state of the rectangle as follows:
@@
 The restriction for stateless calls is the reason why the rendering callback
-\code{cb\_ren}~\cite{tml.js} must be separated from \code{cb\_sim}, which would
+\code{cb\_ren} must be separated from \code{cb\_sim}~\cite{tml.js}, which would
 otherwise render the screen multiple times while time travelling.
@@
 
-The callback \code{cb\_evt} is invoked by the middleware in only real time,
+The callback \code{cb\_evt} is invoked by the middleware only in real time,
 whenever a local SDL event occurs.
@@
 topology, or exhibiting instant feedback on the screen.
-In our example, we only generate application events for the 5 key of interest,
+In our example, we only generate application events for the 5 keys of interest,
 but we also exhibit the pressed key in real time as a visual
@@
 how it ensures the desired properties in concrete terms as follows:
-    (a) peers meet event deadlines either travelling back and forth or in real
-        time (Section \ref{sec.tml.middleware.loop}: \linx{12--16} or
-        \linx{18--22});
-    (b) peers synchronize their clocks with periodic \code{SYNC} packets that
+    (a) peers meet event deadlines either in real time or travelling back and
+        forth (Section \ref{sec.tml.middleware.loop}: \linx{18--22} or
+        \linx{12--16});
+    (b) peers synchronize their clocks with periodic \code{SYNC} packets which
         fast forward late peers (Section \ref{sec.tml.middleware.loop}:
@@
 restrict our programming API to deterministic and stateless calls only.
-This way, it becomes impossible to diverge from other peers by, for example,
+This way, it becomes impossible to diverge from other peers by inadvertently
 re-executing stateful file operations multiple times (or at different times).
@@
 
-We run the application main loop at 50 FPS ($20ms$ per frame) and calculate
+We run the application main loop at 50 FPS (20ms per frame) and calculate
 the middleware overhead at each iteration.
@@
 less than $1/2000$ of the CPU time.
-This means that an application using the middleware would perform all network
-synchronization and still have almost 100\% of the CPU time available for its
-actual domain logic.
+This means that an application using the middleware would perform all
+synchronization upkeep and still have almost 100\% of the CPU time available
+for its actual domain logic.
 %
@@
 As discussed in Section~\ref{sec.tml}.b, the middleware may take up to $1$
-second to catch up, resulting in no more than 2 time travels to recover peers.
+second to catch up, resulting in no more than 2 time travels to recover a peer.
 
@@
 The middleware does not perform well under higher network latencies (e.g., over
-$250ms$).
+250ms).
 This is expected since peers that are multiple hops away take too long to
 synchronize past events and may generate future events in the meantime, which
-causes time travels.
+results in time travels.
 %
@@
 Taking this possibility to an extreme, a peer can still work normally even
-completely offline for a period of time, as desired by local-first
-software~\cite{local}:
+while completely offline for a period of time:
+%, as desired by local-first software~\cite{local}:
 Even though rollbacks are inevitable when reconnecting, their effects may be
-unnoticeable for final users if there is low activity or dependency between
+unnoticeable to final users if there is low activity or dependency between
 events.
@@
 snapshots and deterministic simulation.
-This allows to decentralize the network timeline, since conflicting peers can
-resynchronize their state through time travels.
+This allows us to decentralize the network timeline, since conflicting peers
+can resynchronize their state through time travels.
 %
@@
 The middleware provides \emph{perceptive consistency}~\cite{melding}, which
-adds a time deadline to reach consistency, ensuring that an event triggered by
-any peer at time $t$ is perceived by all peers simultaneously at time
-$t+\Delta$.
+adds a deadline to reach consistency, ensuring that an event triggered by any
+peer at time $t$ is perceived by all peers simultaneously at time $t+\Delta$.
 

### 2024-07-30 spe-24/spe.tex
- spe-24

--- 2022-10-28 p2p-22.tex
@@
-\documentclass[sn-mathphys,iicol]{sn-jnl}% Math and Physical Sciences Reference Style 
-%\documentclass[default,iicol]{sn-jnl}% Default with double column layout
-
-\jyear{2022}%
+\documentclass[sn-mathphys,iicol]{sn-jnl}
+
+\jyear{2024}
 
 \raggedbottom
-%%\unnumbered% uncomment this for unnumbered level heads
 
@@
 For instance, in a simulation scenario of 25 events per minute with a network
-latency of $50ms$, the peers need to roll back and resynchronize only $3\%$ of
+latency of 50ms, the peers need to roll back and resynchronize only 3\% of
 the time.
@@
 }
-
-%%\pacs[JEL Classification]{D8, H51}
-
-%%\pacs[MSC Classification]{35A01, 65L10, 65L12, 65L20, 65L70}
 
@@
 The full source code%
-\footnote{TODO}
-is under 500 lines of code.
+\footnote{https://github.com/fsantanna/p2p-tml}
+is around 500 lines of code.
 
@@
         mirrors of a single application.
-        Video: \url{http://youtube.com/TODO}
+        Video: \url{https://youtu.be/zkl0vSOGino}
         \label{fig.move}
@@
 
-\bibliography{p2p-22}
+\bibliography{spe}
 

### 2024-10-17 rita-24/response.md
...

# Response Letter

We would like to thank the reviewers for the thoughtful comments, questions,
and criticism.

We accommodate all clarifications and answers in a final version of the paper.
To help the reviewers find the modifications of interest, we provide a link
with a *diff* between the versions, as follows:

https://

Based on the overall comments, our focus on the final version was to make the
limitations of our work more explicit, and to improve the section on related
work.

We also address each individual comment from all reviewers as follows...

- usar argumento de que centralizados precisam de bcast
- nosso caso é p2p, nem existe possibilidade de bcast pois nos nao tem
  conhecimento da rede
- possibilidade vem da existencia da time machine, que nao existe nos centralizados

In particular, the dependency on a central server

- target small P2P networks
    - examples / topology extrapolate the target

- pq exec local?
we use an absolute simulation clock to track

The topology has an average of 5 hops between peers
    - ja é bastante radical

We use NetEm [9] to simulate the network latency with
a normal distribution.

The experiments take around 40 hours to complete.

40s online and 20s offline in the
average, including peers 9-11 in the middle of the network.

## Reviewer A

Regarding the experimental evaluation:

> <...>
> However, the experimental evaluation could be more extensive and rigorous.
> <...>
> The proposed approach is technically sound and the experimental results are
> promising.
> <...>
> b) The experimental evaluation could be more extensive, including more
> diverse scenarios and a comparison with existing approaches
> <...>
> The evaluation uses a fixed network topology. How would the middleware's
> performance be affected in scenarios with highly dynamic topologies, where
> peers frequently join, leave, or experience intermittent connectivity?
> Have you considered any mechanisms to adapt to such changes?
> <...>
> The evaluation focuses on simulations with up to 21 peers. How do you
> envision the middleware scaling to much larger networks, and what challenges
> might arise in real-world deployments with potentially hundreds or thousands
> of peers?

Regarding the section on related work:

> a) The related work section could be expanded to include more recent work on
> peer-to-peer systems and distributed consensus algorithms;

Regarding the limiations:

> <...>
> c) The paper could discuss the limitations of the proposed approach and
> potential future work.
> <...>
> The paper assumes non-malicious peers. What security measures or
> modifications would be necessary to protect against malicious peers
> attempting to disrupt the synchronization or inject false events?

Regarding the middleware & protocol:

> While the API description is generally clear, some parts could be further
> clarified, particularly regarding the handling of dynamic memory allocation
> and the interaction between the application and the middleware during time
> travel.
> <...>
> Suggestion: if possible please consider providing a more formal description
> of the time synchronization protocol, including any guarantees or assumptions
> about clock drift and network delays.

## Reviewer J

Regarding the section on related work (comments J.1-J.6):

### Comment J.1

> i) This can do with a subtitles to make it easier to follow the discussion.
> There appears to be two major parts in this section: (1) the discussion on
> the three selected works (Croquet, GALS, CRDTs) in comparison to this work
> and this Work and (2) related works on time machines. It would be beneficial
> to clearly delineate these two subsection, preferably with subtitles.

As suggested, we split Section 2 in two:

- 2.1. Symmetric Distributed Applications
- 2.2. Software Time Machines

### Comment J.2

> ii) The introduction on the related works makes it appear like there are
> ideally 3 subsections: (a) how the network is organized, (b) how global time
> is determined, and (c) how events are propagated and applied.
> However, this three aspects are actually related to the discussion on the
> selected works compared to the proposed solution.

To make clear that we have only two subsections, we now briefly introduce 2.1
and 2.2 with paragraph at the beginning of 2 (before 2.1 and 2.2):

### Comment J.3

> This section first revisits existing solutions for symmetric distributed
> applications, namely Croquet [2], GALS [1], and CRDTs [3]. Then, we discuss
> software time machines that are built for either local or networked
> applications.

We now link the three aspects (a,b,c) with the selected works at the beginning
of 2.1:

> Even though existing solutions aim to provide symmetric behavior for
> distributed applications, they diverge in three key aspects as follows:
> a. How the network is organized.
> b. How global time is determined.
> c. How events are propagated and applied.
> Figure 1 compares three selected works regarding these aspects.
> <...>

### Comment J.4

> --> There is a need to reorganize the related works section, with a proper
> introduction showing the two core aspects of the discussion, (1) and (2) as
> indicated in i), forming the introduction to the section.

Done.

### Comment J.5

> iii) The discussion does not have a very smooth transition. The punchline of
> the related works section should ideally be the proposal of this work, which
> is brought out more clearly when discussed at the end of the related works to
> show how all the gaps identified will be solved.

To guide our discussion, we explain how the text is organized at the beginning
of the section:

> <...>
> Figure 1 compares three selected works regarding these aspects. We now
> discuss these works with respect to the three aspects, while comparing with
> our approach. At the end of this section, we also provide initial details on
> how our middleware works.

### Comment J.6

> iv) The discussion on related work on time machines is rather abrupt.
> It does not have a proper introduction to the problem of time machines in
> relation to this work (or in general for that matter) but directly proceeds
> into an analysis of Fusion and Braid.
> --> What is aimed by this section? Answering this can help find a way of
> introducing the time machines aspect well to the reader.

It was indeed abrupt and out of context.
We now start discussing (a) what a software time machine is, (b) what are the
difficulties to xxx, and (c) what are the common implementation techniques.
Then, we

Regarding peer-to-peer networks, applications, and protocols (comments J.1-J.6):

### Comment J.7

> <...>
> However, in a discussion on p2p application, or even systems, it is necessary
> to highlight the underlying infrastructure to dispel any doubts.
> In this regard, the reviewer would have expected to see the following
> addressed:

> i) As a proof-of-concept, it will be important to have p2p
> model/framework to give a clear picture to an expert reader and explain to a
> lay reader the place of this middleware in the overall p2p application
> model/framework. Has this been considered?

> ii) This proposal uses broadcasting to synchronize peers. What is the
> implication of this assuming a large p2p network? Has this been considered?

> iii) How are the events, as well as the memory snapshots, stored for
> persistence in the p2p network, that is, is there a storage mechanisms used
> so that an offline who joins immediately picks up from where it had dropped off?

> iv) The simulations were done on the same machine. This means that there is
> no actual considerations due to the effect of network challenges. Is this the
> case?

> v) Although this was considering the middleware only, it would be important
> to talk about the impact in terms of network efficiency if any due to the
> middleware.

Regarding grammar issues:

### Comment J.X

> Figure 2 - GALS --> check the spelling for client

OK.

> <...>
> This should be ...among all peers to be the “correct real time”,...

OK.

### 2024-10-18 rita-24/response.md
...

@@
 
+-------------------------------------------------------------------------------
+
 ## Reviewer A
@@
 
+-------------------------------------------------------------------------------
+
 Regarding the section on related work:
@@
 > peer-to-peer systems and distributed consensus algorithms;
+
+-------------------------------------------------------------------------------
 
@@
 > attempting to disrupt the synchronization or inject false events?
+
+-------------------------------------------------------------------------------
 
@@
 
-Regarding the section on related work (comments J.1-J.6):
+Regarding the section on related work (comments J.1-J.5):
 
@@
 
-- 2.1. Symmetric Distributed Applications
-- 2.2. Software Time Machines
+- 2.1. Symmetric Distributed Applications: discussion on the selected works,
+    and comparison with our proposal.
+- 2.2. Software Time Machines: expanded and improved discussion on time
+    machines.
 
@@
 
-To make clear that we have only two subsections, we now briefly introduce 2.1
+Now, to make clear that we have only two subsections, we briefly introduce 2.1
 and 2.2 with paragraph at the beginning of 2 (before 2.1 and 2.2):
-
-### Comment J.3
 
@@
 
-### Comment J.4
+### Comment J.3
 
@@
 
-Done.
+Done, as detailed in J.1 and J.2.
 
-### Comment J.5
+### Comment J.4
 
@@
 
-### Comment J.6
+### Comment J.5
 
@@
 
-It was indeed abrupt and out of context.
-We now start discussing (a) what a software time machine is, (b) what are the
-difficulties to xxx, and (c) what are the common implementation techniques.
-Then, we
+The discussion was indeed abrupt and out of context, so we reshaped the new
+subsection 2.2. We included two new parapgrahs at the very beginning, in which
+we discuss
+    (a) what a software time machine is,
+    (b) in which scenarios it appears,
+    (c) what are the challenges to overcome, and
+    (d) what are the common implementation techniques.
 
-Regarding peer-to-peer networks, applications, and protocols (comments J.1-J.6):
+-------------------------------------------------------------------------------
+
+Regarding the protocol, middleware, and peer-to-peer concepts (comments
+J.7-J.11):
 
 ### Comment J.7
-
-> <...>
-> However, in a discussion on p2p application, or even systems, it is necessary
-> to highlight the underlying infrastructure to dispel any doubts.
-> In this regard, the reviewer would have expected to see the following
-> addressed:
 
@@
 
+### Comment J.8
+
 > ii) This proposal uses broadcasting to synchronize peers. What is the
@@
 
+The use of the term "broadcast" was inaccurate in the paper.
+Now, every occurrence of "broadcast" was substituted by "trigger" or
+"propagation" or "dissemination", depending on the context.
+
+As was described in the Introduction,
+
+> Peers in the application form a dynamic network graph and communicate only
+> with direct neighbours, as in typical unstructured peer-to-peer networks.
+> Events are flooded in the graph and are triggered locally with a small delay
+> to compensate the network latency.
+
+What we meant is that the same effect of a broadcast is reached after the
+events are flooded in the network, but only through direct neighbour
+communication.
+Nevertheless, we no longer use this loosened notion of a broadcast.
+
+We now also emphasize this distinction from centralized solutions (which do
+rely on broadcasts) in the Related Work section:
+
+> Note how the central server is the only participant with knowledge about
+> clients, which never communicate directly.
+> This implies that the server must never fail, and also that periodic full
+> broadcasts are the only possibility of spanning the whole network.
+> In this work, we argue that in a peer-to-peer alternative any node can fail,
+> and that broadcasts can be replaced by optimistic flooding.
+> <...>
+> Note, that it is not necessary that all peers communicate directly with all
+> other peers, but only indirectly via unstructured flooding.
+> <...>
+
+Section "3.2.1 Event Broadcasting" was renamed to "3.2.1 Event Dissemination"
+with the introduction rephrased as follows:
+
+> <...>, events *propagate* between peers with a timestamp scheduled to the
+> future with an extra delta ∆ such that all peers are able to apply them in
+> sync. To prevent *dissemination* cycles, each peer <...>
+
+To conclude, there are no full broadcasts in the protocol.
+Regardless of the network size, peers only communicate with direct neighbours.
+In our experiments, flooding takes 5 hops in the average, with a maximum of 14
+hops (between nodes 0 and 20).
+As an example, with a *100ms* network latency, it takes *500ms* in the average
+to flood an event in the P2P graph.
+We detail these costs in Figure 6.
+Therefore, we do consider the implication of hops and latency in the protocol.
+
+### Comment J.9
+
 > iii) How are the events, as well as the memory snapshots, stored for
 > persistence in the p2p network, that is, is there a storage mechanisms used
-> so that an offline who joins immediately picks up from where it had dropped off?
+> so that an offline who joins immediately picks up from where it had dropped
+> off?
+
+The full history of both events and memory snapshots are stored in all peers.
+Therefore, an offline peer that joins
+
+### Comment J.10
 
@@
 
+### Comment J.11
+
 > v) Although this was considering the middleware only, it would be important
@@
 
-Regarding grammar issues:
+-------------------------------------------------------------------------------
 
-### Comment J.X
+Regarding grammar issues (comments J.12-J.13):
+
+### Comment J.12
 
@@
 
-OK.
+Fixed.
+
+### Comment J.13
 
@@
 
-OK.
+Fixed.

### 2024-10-18 rita-24/response.md
...

@@
 
-Based on the overall comments, our focus on the final version was to make the
-limitations of our work more explicit, and to improve the section on related
-work.
+Based on the overall comments, our focus on the final version was to improve
+the section on related work, and to make the limitations of our proposal more
+explicit.
+For that matter, we expanded the section on related work splitting it in two
+subsections
+    "2.1 Symmetric Distributed Applications" and
+    "2.2 Software Time Machines".
+We also included a new section "3.3 Middleware Limitations" to discuss the
+limitations of our work.
 
@@
 
-Regarding the limiations:
+Regarding the limitations:
 
@@
 
-- 2.1. Symmetric Distributed Applications: discussion on the selected works,
+- 2.1 Symmetric Distributed Applications: discussion on the selected works,
     and comparison with our proposal.
-- 2.2. Software Time Machines: expanded and improved discussion on time
+- 2.2 Software Time Machines: expanded and improved discussion on time
     machines.
@@
 The discussion was indeed abrupt and out of context, so we reshaped the new
-subsection 2.2. We included two new parapgrahs at the very beginning, in which
+subsection 2.2. We included two new paragraphs at the very beginning, in which
 we discuss
@@
 The full history of both events and memory snapshots are stored in all peers.
-Therefore, an offline peer that joins
+There are no differences among peers with respect to their roles: any peer can
+fail and any peer can offer full recovery.
+Therefore, an offline peer that (re)joins the network already holds past
+snapshots, and can fully recover from any direct neighbour.
+
+We added a remark at the beginning of Section 3 "The Middleware & Programming
+API", when describing the middleware architecture with Figure 4:
+
+> Note that all peers execute the exactly same application and middleware, with
+> no differences with respect to their roles and physical resources.
+
+We now also explicit that snapshots are replicated, further in the same
+section:
+
+> As an optimization, the middleware takes periodic snapshots locally *at all
+> peers* to avoid full simulation.
+> <...>
+> In addition, a peer that rejoins the network recovers from a past snapshot it
+> already holds, thus avoiding full simulation.
 
@@
 
+The evaluation requires a synchronized global clock that can measure
+differences in the order of milliseconds.
+We use Linux's NetEm to emulate network latency, but do not consider other
+network challenges.
+That being said, latency is the only network dependent parameter that we
+evaluate.
+
+We
+
 ### Comment J.11
@@
 > middleware.
+
+I'm not sure if I understand this comment correctly.
+In Section 4, we discuss the negligible CPU impact the middleware has over the
+application as a whole, which is 9us every 20ms.
+Regarding the network, the middleware only receives and forwards event packets
+among neighbours, in the same way typical flooding protocols behave.
+
+We now include a discussion about packet forwarding at the end of Section
+"3.2.1 Event Dissemination":
+
+> Considering the peer-to-peer network as a whole, each event packet is
+> replicated to all neighbours of all peers.
+> This results in a theoretical limit of quadratic messages if all peers are
+> connected to all peers.
+> In contrast, as discussed in Section 2, a centralized solution only requires
+> a packet to be sent to a server, which then broadcasts it to all clients.
 

### 2024-10-22 rita-24/response.md
...

@@
 explicit.
-For that matter, we expanded the section on related work splitting it in two
+
+For that matter, we expanded Section "2. Related Work" by splitting it in two
 subsections
-    "2.1 Symmetric Distributed Applications" and
-    "2.2 Software Time Machines".
-We also included a new section "3.3 Middleware Limitations" to discuss the
-limitations of our work.
+    "2.1. Symmetric Distributed Applications" and
+    "2.2. Software Time Machines".
+
+We also expanded Section "3.2.4. Middleware Summary" to discuss its
+limitations, and added a paragraph at the end of Section "4. Evaluation" with
+the same goal.
+As detailed next, we discuss and justify the following limitations: peer
+identifiers, deterministic API, static memory, snapshots footprint,
+non-malicious peers, and network size.
 
@@
 
-- usar argumento de que centralizados precisam de bcast
-- nosso caso é p2p, nem existe possibilidade de bcast pois nos nao tem
-  conhecimento da rede
-- possibilidade vem da existencia da time machine, que nao existe nos centralizados
+-------------------------------------------------------------------------------
 
-In particular, the dependency on a central server
+EVALUATION
 
@@
     - examples / topology extrapolate the target
-
-- pq exec local?
-we use an absolute simulation clock to track
-
-The topology has an average of 5 hops between peers
+- The topology has an average of 5 hops between peers
     - ja é bastante radical
-
-We use NetEm [9] to simulate the network latency with
-a normal distribution.
-
-The experiments take around 40 hours to complete.
-
-40s online and 20s offline in the
-average, including peers 9-11 in the middle of the network.
+- The experiments take around 40 hours to complete.
+- 40s online and 20s offline in the average, including peers 9-11 in the middle of the network.
 
@@
 
+### Comment A.TODO
+
 > <...>
@@
 > potential future work.
+
+### Comment A.TODO
+
 > <...>
@@
 
+We assume non-malicious peers in the network.
+This is a fundamental limitation also shared by the alternatives discussed in
+related work.
+Even with CRDTs, nothing prevents a malicious peer to inject 1M events per
+minute to break the application.
+
+We now point this limitation in the Introduction, along with a reposition to
+target small networks, implying more control over participants:
+
+> We target small peer-to-peer networks, in which nodes are a only few hops
+> away from each other.
+> This ensures that events can span the whole network in a reasonable time to
+> preserve the real-time behavior of applications.
+> We also assume that peers are non-malicious in the sense that they do not
+> generate erroneous events.
+
+In Section "3.2.4. Middleware Summary", we now also explain how this limitation
+emerges in our API:
+
+> As detailed in Section 3.2.1, note that the API to disseminate an event in
+> the network is as simple as a variable assignment. Therefore, a malicious
+> peer could SPAM or inject erroneous events to break the application.
+> Again, this is a fundamental limitation that is also shared with the
+> alternatives discussed in Section 2.
+
 -------------------------------------------------------------------------------
@@
 
+### Comment A.TODO
+
 > While the API description is generally clear, some parts could be further
 > clarified, particularly regarding the handling of dynamic memory allocation
+> <...>
+
+In Section "3.2.4. Middleware Summary", we now also illustrate how to
+accommodate dynamic allocation over static snapshots:
+
+> Regarding the static limitation, we mentioned the possibility of a custom
+> allocator, which is illustrated as follows:
+
+```
+struct {
+    // FIXED REGION
+    int x,  y;   // position
+    int dx, dy;  // direction speed
+    // VARIABLE REGION
+    char heap[HEAP_SIZE];
+} G;
+
+T* obj = my_malloc(G.heap, <size>);
+<...>
+my_free(obj);
+```
+
+> The variable region is still technically static, but is dynamically reshaped
+> internally with a custom allocator `my\_alloc` & `my\_free`.
+
+### Comment A.TODO
+
+> <...>
 > and the interaction between the application and the middleware during time
 > travel.
+> <...>
+
+The application is frozen during time travel.
+
+In Section "3.2.3. The Time Machine", we now include the paragraph as follows:
+
+> The step delay is dynamically adjusted such that the whole time travel loop
+> takes at most 1 second to complete.
+> Note that during this period, the application is frozen, and new inputs are
+> enqueued for further processing.
+
+### Comment A.TODO
+
 > <...>
@@
 > about clock drift and network delays.
+
+Even though we consider that this suggestion is relevant, we believe that a
+formal description of the protocol with its implications requires a dedicated
+research on its own.
 
@@
 We now also emphasize this distinction from centralized solutions (which do
-rely on broadcasts) in the Related Work section:
+rely on broadcasts) in Section "2. Related Work":
 
@@
 
-Section "3.2.1 Event Broadcasting" was renamed to "3.2.1 Event Dissemination"
+Section "3.2.1. Event Broadcasting" was renamed to "3.2.1. Event Dissemination"
 with the introduction rephrased as follows:
@@
 
-We added a remark at the beginning of Section 3 "The Middleware & Programming
+We added a remark at the beginning of Section "3. The Middleware & Programming
 API", when describing the middleware architecture with Figure 4:
@@
 We now include a discussion about packet forwarding at the end of Section
-"3.2.1 Event Dissemination":
+"3.2.1. Event Dissemination":
 

### 2024-10-22 rita-24/response.md
...

@@
 
+```
 https://
+```
 
@@
     "2.2. Software Time Machines".
-
 We also expanded Section "3.2.4. Middleware Summary" to discuss its
@@
 the same goal.
-As detailed next, we discuss and justify the following limitations: peer
-identifiers, deterministic API, static memory, snapshots footprint,
-non-malicious peers, and network size.
 
 We also address each individual comment from all reviewers as follows...
-
--------------------------------------------------------------------------------
-
-EVALUATION
-
-- target small P2P networks
-    - examples / topology extrapolate the target
-- The topology has an average of 5 hops between peers
-    - ja é bastante radical
-- The experiments take around 40 hours to complete.
-- 40s online and 20s offline in the average, including peers 9-11 in the middle of the network.
+(Some comments might be reordered.)
 
@@
 
-Regarding the experimental evaluation:
+Regarding the section on related work:
 
-> <...>
-> However, the experimental evaluation could be more extensive and rigorous.
-> <...>
-> The proposed approach is technically sound and the experimental results are
-> promising.
-> <...>
-> b) The experimental evaluation could be more extensive, including more
-> diverse scenarios and a comparison with existing approaches
-> <...>
-> The evaluation uses a fixed network topology. How would the middleware's
-> performance be affected in scenarios with highly dynamic topologies, where
-> peers frequently join, leave, or experience intermittent connectivity?
-> Have you considered any mechanisms to adapt to such changes?
-> <...>
-> The evaluation focuses on simulations with up to 21 peers. How do you
-> envision the middleware scaling to much larger networks, and what challenges
-> might arise in real-world deployments with potentially hundreds or thousands
-> of peers?
-
--------------------------------------------------------------------------------
-
-Regarding the section on related work:
+### Comment A.1
 
@@
 > peer-to-peer systems and distributed consensus algorithms;
+
+Besides the structural modifications, we now also refer to the following works:
+
+- [2022] "An experiment in live collaborative programming on the croquet shared
+   experience platform"
+    - Yoshiki Ohshima, ..., David A. Smith
+    - The last author is the first author of the original Croquet paper from
+      2005.
+    - They reimplement the original Croquet in JavaScript to run in modern
+      computers. In their Section 2, they describe the network architecture
+      that we discuss in our Figure 2.
+    - They also refer to CRDTs and derivatives as the most recent alternative
+      with respect to real-time collaborative protocols (they cite the same
+      papers we do).
+- [2023] "Collaborative Live Modelling by Language-Agnostic Versioning"
+    - Joeri Exelmans, ..., Hans Vangheluwe
+    - This work uses the term "synchronous collaboration"
+
+Most literature on distributed consensus algorithms focus on fault tolerance
+(Raft/Paxos/SMR/PBFT), but not on real-time symmetric behavior.
 
@@
 
-### Comment A.TODO
+### Comment A.2
 
@@
 
-### Comment A.TODO
+We now discuss the limitations of our work as follows...
+
+- 1. Introduction
+
+> We target small peer-to-peer networks, in which nodes are a only few hops
+> away from each other. This ensures that events can span the whole network in
+> a reasonable time to preserve the real-time behavior of applications. We also
+> assume that peers are non-malicious in the sense that they do not generate
+> erroneous events.
+
+- 3.2.1. Event Dissemination:
+
+> Considering the peer-to-peer network as a whole, each event packet is
+> replicated to all neighbours of all peers. This results in a theoretical
+> limit of quadratic messages if all peers are connected to all peers. In
+> contrast, as discussed in Section~\ref{sec.related.sym}, a centralized
+> solution only requires a packet to be sent to a server, which then broadcasts
+> it to all clients.
+
+- 3.2.4. Middleware Summary
+
+> Contiguous unique identifiers:
+> Deterministic API:
+> Non-malicious peers:
+> Static memory footprint:
+
+(Details in the paper or link above.)
+
+### Comment A.3
 
@@
 
-### Comment A.TODO
+### Comment A.4
 
@@
 
-### Comment A.TODO
+### Comment A.5
 
@@
 
-### Comment A.TODO
+### Comment A.6
 
@@
 research on its own.
+
+-------------------------------------------------------------------------------
+
+Regarding the experimental evaluation:
+
+### Comment A.7
+
+> <...>
+> However, the experimental evaluation could be more extensive and rigorous.
+> <...>
+> The proposed approach is technically sound and the experimental results are
+> promising.
+> <...>
+> b) The experimental evaluation could be more extensive, including more
+> diverse scenarios and a comparison with existing approaches
+
+> <...>
+> The evaluation uses a fixed network topology. How would the middleware's
+> performance be affected in scenarios with highly dynamic topologies, where
+> peers frequently join, leave, or experience intermittent connectivity?
+> Have you considered any mechanisms to adapt to such changes?
+
+Even though the network topology is fixed, it includes three typical variations
+of straight bus lines, rings, and connected topologies.
+They all coexist in the same deployment with peers leaving and joining the
+network.
+In item "(c) Correctness of unstable peers", we also make all 21 nodes to
+switch between online and offline for the same amount of time, including the
+peers in the middle of the network.
+This periodically creates full partitions in the peer-to-peer network, making
+the two opposites incommunicable for a considerable period of time.
+
+> <...>
+> The evaluation focuses on simulations with up to 21 peers. How do you
+> envision the middleware scaling to much larger networks, and what challenges
+> might arise in real-world deployments with potentially hundreds or thousands
+> of peers?
+
+We now position the scope of our work to small peer-to-peer networks only:
+
+> We target small peer-to-peer networks, in which nodes are a only few hops
+> away from each other. This ensures that events can span the whole network in
+> a reasonable time to preserve the real-time behavior of applications.
+
+As suggested in this paragraph, our design is not appropriate for too many
+hops, since each hop increases the event latency, which harms real-time
+behavior.
 
@@
 > <...>, events *propagate* between peers with a timestamp scheduled to the
-> future with an extra delta ∆ such that all peers are able to apply them in
+> future with an extra delta such that all peers are able to apply them in
 > sync. To prevent *dissemination* cycles, each peer <...>
@@
 
-We
-
 ### Comment J.11

### 2024-10-24 rita-24/response.md
...

@@
 
-> a) The related work section could be expanded to include more recent work on
+> The related work section could be expanded to include more recent work on
 > peer-to-peer systems and distributed consensus algorithms;
 
-Besides the structural modifications, we now also refer to the following works:
+We now expanded the end of Section "2.1 Symmetric Distributed Applications" to
+put our work into the perspective of classical distributed consensus
+algorithms:
+
+> Finally, note that classical distributed consensus mechanisms focus on fault
+> tolerance [8], but not on real-time symmetric behavior. These algorithms
+> typically rely on leaders [9], which require an election period that does not
+> cope with our symmetric real-time requirements. In this sense, these
+> algorithms could be considered to decentralize servers in Croquet or GALS,
+> but not symmetric peers with high churn rates, as we propose. In contrast,
+> our proposal requires a time machine to reapply reordered operations, since
+> we cannot guarantee strong consistency [10].
+
+We now include references on consensus mechanisms:
+
+- [2019] "Concurrency: the works of Leslie Lamport"
+    - Leslie Lamport, Robert Shostak, and Marshall Pease
+- [2020] "Paxos vs Raft: Have we reached consensus on distributed consensus?"
+    - Heidi Howard, Richard Mortier
+- [2017] "Verifying strong eventual consistency in distributed systems"
+    - Victor Gomes, ..., Alastair Beresford
+
+We also included a new reference for Croquet that has been published recently:
 
@@
     - They also refer to CRDTs and derivatives as the most recent alternative
-      with respect to real-time collaborative protocols (they cite the same
-      papers we do).
-- [2023] "Collaborative Live Modelling by Language-Agnostic Versioning"
-    - Joeri Exelmans, ..., Hans Vangheluwe
-    - This work uses the term "synchronous collaboration"
-
-Most literature on distributed consensus algorithms focus on fault tolerance
-(Raft/Paxos/SMR/PBFT), but not on real-time symmetric behavior.
+      with respect to real-time collaborative protocols (they roughly cite the
+      same papers we do).
 
@@
 > <...>
-> c) The paper could discuss the limitations of the proposed approach and
+> The paper could discuss the limitations of the proposed approach and
 > potential future work.
@@
 > <...>
-> However, the experimental evaluation could be more extensive and rigorous.
-> <...>
-> The proposed approach is technically sound and the experimental results are
-> promising.
-> <...>
-> b) The experimental evaluation could be more extensive, including more
-> diverse scenarios and a comparison with existing approaches
+> The experimental evaluation could be more extensive, including more
+> diverse scenarios *(see A.8)* and a comparison with existing approaches
+
+At the end of Section "4. Evaluation", we compare the existing approaches
+discussed in Section "2.1. Symmetric Distributed Applications" with respect to
+the three aspects that we consider in our evaluation (i.e., recurrence of time
+travels, real-time pace of peers, correctness of unstable peers):
+
+### Comment A.8
 
@@
 the two opposites incommunicable for a considerable period of time.
+
+### Comment A.9
 

### 2024-10-24 rita-24/response.md
...

@@
 
-> Finally, note that classical distributed consensus mechanisms focus on fault
-> tolerance [8], but not on real-time symmetric behavior. These algorithms
-> typically rely on leaders [9], which require an election period that does not
-> cope with our symmetric real-time requirements. In this sense, these
-> algorithms could be considered to decentralize servers in Croquet or GALS,
-> but not symmetric peers with high churn rates, as we propose. In contrast,
-> our proposal requires a time machine to reapply reordered operations, since
-> we cannot guarantee strong consistency [10].
+> "Finally, note that classical distributed consensus mechanisms focus on
+> fault tolerance [8], but not on real-time symmetric behavior. These
+> algorithms typically rely on leaders [9], which require an election period
+> that does not cope with our symmetric real-time requirements. In this sense,
+> these algorithms could be considered to decentralize servers in Croquet or
+> GALS, but not to decentralize symmetric peers with high churn rates, as we
+> propose in this work. In contrast, our proposal requires a time machine to
+> reapply reordered operations, since we cannot guarantee strong consistency
+> [10]."
 
@@
 
-- 1. Introduction
+- 1. Introduction:
 
-> We target small peer-to-peer networks, in which nodes are a only few hops
+> "We target small peer-to-peer networks, in which nodes are only a few hops
 > away from each other. This ensures that events can span the whole network in
@@
 > assume that peers are non-malicious in the sense that they do not generate
-> erroneous events.
+> erroneous events."
 
@@
 
-> Considering the peer-to-peer network as a whole, each event packet is
+> "Considering the peer-to-peer network as a whole, each event packet is
 > replicated to all neighbours of all peers. This results in a theoretical
 > limit of quadratic messages if all peers are connected to all peers. In
-> contrast, as discussed in Section~\ref{sec.related.sym}, a centralized
-> solution only requires a packet to be sent to a server, which then broadcasts
-> it to all clients.
+> contrast, as discussed in Section 2.1, a centralized solution only requires a
+> packet to be sent to a server, which then broadcasts it to all clients."
 
-- 3.2.4. Middleware Summary
+- 3.2.4. Middleware Summary:
 
-> Contiguous unique identifiers:
-> Deterministic API:
-> Non-malicious peers:
-> Static memory footprint:
-
-(Details in the paper or link above.)
+We discuss the middleware limiations as follows:
+contiguous unique identifiers, deterministic API, non-malicious peers, and
+static memory footprint.
+The text with details are in the paper or link above.
 
@@
 
-> We target small peer-to-peer networks, in which nodes are a only few hops
+> "We target small peer-to-peer networks, in which nodes are a only few hops
 > away from each other.
@@
 > We also assume that peers are non-malicious in the sense that they do not
-> generate erroneous events.
+> generate erroneous events."
 
@@
 
-> As detailed in Section 3.2.1, note that the API to disseminate an event in
+> "As detailed in Section 3.2.1, note that the API to disseminate an event in
 > the network is as simple as a variable assignment. Therefore, a malicious
@@
 > Again, this is a fundamental limitation that is also shared with the
-> alternatives discussed in Section 2.
+> alternatives discussed in Section 2."
 
@@
 
-> Regarding the static limitation, we mentioned the possibility of a custom
-> allocator, which is illustrated as follows:
+> "Regarding the static limitation, we mentioned the possibility of a custom
+> allocator, which is illustrated as follows:"
 
@@
 
-> The variable region is still technically static, but is dynamically reshaped
-> internally with a custom allocator `my\_alloc` & `my\_free`.
+> "The variable region is still technically static, but is dynamically reshaped
+> internally with a custom allocator `my\_alloc` & `my\_free`."
 
@@
 
-> The step delay is dynamically adjusted such that the whole time travel loop
+> "The step delay is dynamically adjusted such that the whole time travel loop
 > takes at most 1 second to complete.
 > Note that during this period, the application is frozen, and new inputs are
-> enqueued for further processing.
+> enqueued for further processing."
 
@@
 
-> We target small peer-to-peer networks, in which nodes are a only few hops
+> "We target small peer-to-peer networks, in which nodes are a only few hops
 > away from each other. This ensures that events can span the whole network in
-> a reasonable time to preserve the real-time behavior of applications.
+> a reasonable time to preserve the real-time behavior of applications."
 
@@
 
-> This section first revisits existing solutions for symmetric distributed
+> "This section first revisits existing solutions for symmetric distributed
 > applications, namely Croquet [2], GALS [1], and CRDTs [3]. Then, we discuss
 > software time machines that are built for either local or networked
-> applications.
+> applications."
 
@@
 
-> Even though existing solutions aim to provide symmetric behavior for
+> "Even though existing solutions aim to provide symmetric behavior for
 > distributed applications, they diverge in three key aspects as follows:
@@
 > c. How events are propagated and applied.
-> Figure 1 compares three selected works regarding these aspects.
+> Figure 1 compares three selected works regarding these aspects."
 > <...>
@@
 > <...>
-> Figure 1 compares three selected works regarding these aspects. We now
+> "Figure 1 compares three selected works regarding these aspects. We now
 > discuss these works with respect to the three aspects, while comparing with
 > our approach. At the end of this section, we also provide initial details on
-> how our middleware works.
+> how our middleware works."
 
@@
 
-> Peers in the application form a dynamic network graph and communicate only
+> "Peers in the application form a dynamic network graph and communicate only
 > with direct neighbours, as in typical unstructured peer-to-peer networks.
 > Events are flooded in the graph and are triggered locally with a small delay
-> to compensate the network latency.
+> to compensate the network latency."
 
@@
 
-> Note how the central server is the only participant with knowledge about
+> "Note how the central server is the only participant with knowledge about
 > clients, which never communicate directly.
@@
 > In this work, we argue that in a peer-to-peer alternative any node can fail,
-> and that broadcasts can be replaced by optimistic flooding.
+> and that broadcasts can be replaced by optimistic flooding."
 > <...>
-> Note, that it is not necessary that all peers communicate directly with all
-> other peers, but only indirectly via unstructured flooding.
+> "Note, that it is not necessary that all peers communicate directly with all
+> other peers, but only indirectly via unstructured flooding."
 > <...>
@@
 
-> <...>, events *propagate* between peers with a timestamp scheduled to the
+> <...>, "events *propagate* between peers with a timestamp scheduled to the
 > future with an extra delta such that all peers are able to apply them in
-> sync. To prevent *dissemination* cycles, each peer <...>
+> sync. To prevent *dissemination* cycles, each peer" <...>
 
@@
 
-> Note that all peers execute the exactly same application and middleware, with
-> no differences with respect to their roles and physical resources.
+> "Note that all peers execute the exactly same application and middleware,
+> with no differences with respect to their roles and physical resources."
 
@@
 
-> As an optimization, the middleware takes periodic snapshots locally *at all
-> peers* to avoid full simulation.
+> "As an optimization, the middleware takes periodic snapshots locally *at all
+> peers* to avoid full simulation."
 > <...>
-> In addition, a peer that rejoins the network recovers from a past snapshot it
-> already holds, thus avoiding full simulation.
+> "In addition, a peer that rejoins the network recovers from a past snapshot
+> it already holds, thus avoiding full simulation."
 
@@
 
-> Considering the peer-to-peer network as a whole, each event packet is
+> "Considering the peer-to-peer network as a whole, each event packet is
 > replicated to all neighbours of all peers.
@@
 > In contrast, as discussed in Section 2, a centralized solution only requires
-> a packet to be sent to a server, which then broadcasts it to all clients.
+> a packet to be sent to a server, which then broadcasts it to all clients."
 

### 2024-10-25 rita-24/response.md
...

@@
     "2.2. Software Time Machines".
-We also expanded Section "3.2.4. Middleware Summary" to discuss its
-limitations, and added a paragraph at the end of Section "4. Evaluation" with
-the same goal.
+We also rewrote Section "3.2.4. Middleware Summary" to discuss its limitations.
 
-We also address each individual comment from all reviewers as follows...
+Next, we address each individual comment from all reviewers as follows...
 (Some comments might be reordered.)
@@
 
-Regarding the section on related work:
+Regarding the section on related work (comment A.1):
 
@@
 
-We now include references on consensus mechanisms:
+We now include recent references on consensus mechanisms:
 
@@
 
-Regarding the limitations:
+Regarding the limitations (comments A.2-A.3):
 
@@
 > replicated to all neighbours of all peers. This results in a theoretical
-> limit of quadratic messages if all peers are connected to all peers. In
+> limit of quadratic messages if all peers were connected to all peers. In
 > contrast, as discussed in Section 2.1, a centralized solution only requires a
@@
 
-We discuss the middleware limiations as follows:
+We discuss the middleware limitaions as follows:
 contiguous unique identifiers, deterministic API, non-malicious peers, and
 static memory footprint.
-The text with details are in the paper or link above.
+The text is too long to fit here and can be found in the paper or link above.
 
@@
 
-Regarding the middleware & protocol:
+Regarding the middleware & protocol (comments A.4-A.6):
 
@@
 > "The variable region is still technically static, but is dynamically reshaped
-> internally with a custom allocator `my\_alloc` & `my\_free`."
+> internally with the custom allocator `my_alloc` & `my_free`."
 
@@
 formal description of the protocol with its implications requires a dedicated
-research on its own.
+work on its own.
 
@@
 
-Regarding the experimental evaluation:
+Regarding the experimental evaluation (comments A.7-A.9):
 
@@
 
-At the end of Section "4. Evaluation", we compare the existing approaches
+At the end of Section "4. Evaluation", we now compare the existing approaches
 discussed in Section "2.1. Symmetric Distributed Applications" with respect to
 the three aspects that we consider in our evaluation (i.e., recurrence of time
-travels, real-time pace of peers, correctness of unstable peers):
+travels, real-time pace of peers, correctness of unstable peers).
+The text is too long to fit here and can be found in the paper or link above.
 
@@
 Even though the network topology is fixed, it includes three typical variations
-of straight bus lines, rings, and connected topologies.
+of (a) straight bus lines, (b) rings, and (c) connected topologies.
 They all coexist in the same deployment with peers leaving and joining the
 network.
-In item "(c) Correctness of unstable peers", we also make all 21 nodes to
-switch between online and offline for the same amount of time, including the
-peers in the middle of the network.
-This periodically creates full partitions in the peer-to-peer network, making
-the two opposites incommunicable for a considerable period of time.
+In item "(c) Correctness of unstable peers", we make all 21 nodes to switch
+between online and offline for the same amount of time, including the peers in
+the middle of the network.
+This churn rate of 50% periodically creates full partitions in the peer-to-peer
+network, making the two opposites incommunicable for a considerable amount of
+time.
 
@@
 
-As suggested, we split Section 2 in two:
+As suggested, we now split Section 2 in two:
 
@@
 Now, to make clear that we have only two subsections, we briefly introduce 2.1
-and 2.2 with paragraph at the beginning of 2 (before 2.1 and 2.2):
+and 2.2 with a paragraph at the beginning of Section 2 (before 2.1 and 2.2):
 
@@
 
-To guide our discussion, we explain how the text is organized at the beginning
-of the section:
+To guide our discussion, we now explain how the text is organized at the
+beginning of Section 2.1:
 
@@
     (d) what are the common implementation techniques.
+The text is too long to fit here and can be found in the paper or link above.
 
@@
 > model/framework. Has this been considered?
+
+We now include a paragraph at the beginning of Section "3. The Middleware &
+Programming API" describing the P2P framework or, in other words, what we
+exactly mean by P2P:
+
+> "The peers form a dynamic unstructured peer-to-peer network [3], and
+> communicate only indirectly to each other. Events are flooded in the network
+> graph via gossiping, i.e.: when a peer generates an event, it communicates to
+> its neighbours, which communicate to their neighbours, and so on. Note that
+> all peers execute the exactly same application and middleware, with no
+> differences with respect to their roles and physical resources. Therefore, we
+> can describe our peer-to-peer network as follows: (a) all peers have the same
+> role and run the same software; (b) any peer can join or leave at any time;
+> (c) events are only communicated with direct neighbours."
 
@@
 > "Note how the central server is the only participant with knowledge about
-> clients, which never communicate directly.
+> clients, which never communicate directly among themselves.
 > This implies that the server must never fail, and also that periodic full

### 2024-12-06 rita-24/response.md
...
- versao final

@@
 ```
-https://
+https://github.com/fsantanna-no/p2p-tml-paper/compare/690245...main
 ```
@@
 
-We discuss the middleware limitaions as follows:
+We discuss the middleware limitations as follows:
 contiguous unique identifiers, deterministic API, non-malicious peers, and
@@
 time.
+Therefore, our unstructured design does provide the necessary mechanisms to
+adapt to such changes, since peers joins and leaves only impact direct
+neighbours.
+Our evaluation takes high churn rates into consideration.
 
@@
 > <...>
-> "Figure 1 compares three selected works regarding these aspects. We now
-> discuss these works with respect to the three aspects, while comparing with
-> our approach. At the end of this section, we also provide initial details on
-> how our middleware works."
+> "Figure 1 compares three selected works regarding these aspects, which we
+> discuss and contrast with our approach next. At the end of this section, we
+> also provide initial details on how our middleware works.
 
@@
 > "The peers form a dynamic unstructured peer-to-peer network [3], and
-> communicate only indirectly to each other. Events are flooded in the network
-> graph via gossiping, i.e.: when a peer generates an event, it communicates to
-> its neighbours, which communicate to their neighbours, and so on. Note that
-> all peers execute the exactly same application and middleware, with no
-> differences with respect to their roles and physical resources. Therefore, we
-> can describe our peer-to-peer network as follows: (a) all peers have the same
-> role and run the same software; (b) any peer can join or leave at any time;
-> (c) events are only communicated with direct neighbours."
+> communicate mostly indirectly to each other. Events are flooded in the
+> network graph via gossiping, i.e.: when a peer generates an event, it
+> communicates to its neighbours, which communicate to their neighbours, and so
+> on. Note that all peers execute the exactly same application and middleware,
+> with no differences with respect to their roles and physical resources.
+> Therefore, we can describe our peer-to-peer network as follows: (a) all peers
+> have the same role and run the same software; (b) any peer can join or leave
+> at any time; (c) events are only communicated with direct neighbours."
 
@@
 The evaluation requires a synchronized global clock that can measure
-differences in the order of milliseconds.
+differences in the order of a few milliseconds.
+This can be done reliably in a single machine, but not trivially across the
+network with the precision that we needed.
 We use Linux's NetEm to emulate network latency, but do not consider other
@@
 That being said, latency is the only network dependent parameter that we
-evaluate.
+currently evaluate.
 

### 2024-12-06 rita-24/rita.tex
- versao final

--- 2024-07-30 spe-24/spe.tex
@@
-\documentclass[sn-mathphys,iicol]{sn-jnl}
-
-\jyear{2024}
-
-\raggedbottom
-
-\usepackage{verbatim}
+\documentclass[fleqn,10pt]{SelfArx}
+\usepackage[english]{babel}
+\usepackage{lipsum}
 \usepackage{url}
-\usepackage{graphicx}
-
-\usepackage{xspace}
-\newcommand{\FC}       {Freechains\xspace}
-\newcommand{\reps}     {\emph{reps}\xspace}
-\newcommand{\onerep}   {\emph{1~rep}\xspace}
-\newcommand{\nreps}[1] {\emph{#1~reps\xspace}}
+\usepackage{balance}
+
 \newcommand{\code}[1]  {\texttt{\footnotesize{#1}}}
-\newcommand{\Xon} {$1{\rightarrow}N$\xspace}
-\newcommand{\Xno} {$1{\leftarrow}N$\xspace}
-\newcommand{\Xnn} {$N{\leftrightarrow}N$\xspace}
-\newcommand{\Xoo} {$1{\leftrightarrow}1$\xspace}
-\newcommand{\Xo}  {$1{\hookleftarrow}$\xspace}
 \newcommand{\lin}[1]{(\emph{ln. #1}\xspace)}
@@
 
-\renewcommand{\theenumi}{\alph{enumi}}
-
-\hyphenation{off-line}
-
-\begin{document}
-
-\title[Article Title]{
-    Symmetric Peer-to-Peer Applications
-}
-
-\author{\fnm{Francisco} \sur{Sant'Anna} \email{francisco@ime.uerj.br}}
-\affil{\orgdiv{Department of Computer Science}, \orgname{Rio de Janeiro State University}, \orgaddress{\country{Brazil}}}
-
-\abstract{
+\usepackage{xspace}
+\usepackage{hyperref}
+\usepackage{mathptmx}
+\usepackage{graphicx}
+\usepackage{times}
+\usepackage{textcomp}
+\usepackage{gensymb} % degree. ie 360º
+\usepackage{amsmath}
+\usepackage{comment}
+\usepackage{multirow}
+\usepackage[framemethod=tikz]{mdframed}
+\usepackage{graphicx}
+\usepackage[labelfont=bf]{subcaption}
+
+\usepackage{tcolorbox}% http://ctan.org/pkg/tcolorbox
+\definecolor{mycolor}{rgb}{0,0,0}% Rule colour
+\makeatletter
+\newcommand{\mybox}[1]{%
+	\setbox0=\hbox{#1}%
+	\setlength{\@tempdima}{\dimexpr\wd0+13pt}%
+	\begin{tcolorbox}[colframe=mycolor,boxrule=0.5pt,arc=4pt,
+		left=6pt,right=6pt,top=6pt,bottom=6pt,boxsep=0pt,width=\@tempdima]
+		#1
+	\end{tcolorbox}
+}
+\makeatother
+\usepackage[resetlabels,labeled]{multibib}
+\newcites{A}{Appendix}
+\usepackage[num]{abntex2cite}
+\citebrackets[]
+%----------------------------------------------------------------------------------------
+%	COLUMNS
+%----------------------------------------------------------------------------------------
+\setlength{\columnsep}{0.55cm} % Distance between the two columns of text
+\setlength{\fboxrule}{0.75pt} % Width of the border around the abstract
+%----------------------------------------------------------------------------------------
+%	COLORS
+%----------------------------------------------------------------------------------------
+\definecolor{color1}{RGB}{0,0,90}
+\definecolor{color2}{RGB}{0,20,20}
+%----------------------------------------------------------------------------------------
+%	HYPERLINKS
+%----------------------------------------------------------------------------------------
+\usepackage{hyperref} % Required for hyperlinks
+\hypersetup{hidelinks,colorlinks,breaklinks=true,urlcolor=color2,citecolor=color1,linkcolor=color1,bookmarksopen=false,pdftitle={Title},pdfauthor={Author}}
+%----------------------------------------------------------------------------------------
+%	ARTICLE INFORMATION
+%----------------------------------------------------------------------------------------
+\JournalInfo{\color{gray}\normalsize\sffamily\bfseries Revista de Informática Teórica e Aplicada - RITA - ISSN 2175-2745\\ Vol.~XX, Num.~XX~(2018)~11-XX} % Journal information
+\Archive{\mybox{RESEARCH ARTICLE}} % Type of Article: Researh, Review, Tutorial, Study Case
+
+\PaperTitle{Symmetric Peer-to-Peer Applications}
+\ShortTitle{Symmetric Peer-to-Peer Applications}
+
+\Titulo{Aplicações Peer-to-Peer Simétricas}
+
+\PaperVol{XX} % Volume 
+\PaperNum{X} % Number 
+\PaperAno{YYYY} % Year of Publication
+
+\Authors{Francisco Sant'Anna\textsuperscript{1}*}
+\affiliation{\textsuperscript{1}\textit{Department of Computer Science, Rio de Janeiro State University (UERJ), Brazil}}
+\affiliation{*\textbf{Corresponding author}: francisco@ime.uerj.br}
+
+\Keywords{collaboration --- determinism --- peer-to-peer --- time machine }
+
+\newcommand{\keywordname}{Keywords} % Defines the keywords heading name
+
+\PalavrasChave{colaboração --- determinismo --- peer-to-peer --- máquina do tempo} 
+\Doi{http://dx.doi.org/10.22456/2175-2745.XXXX} % DOI
+\DateR{dd/mm/yyyy} % Received
+\DateA{dd/mm/yyyy} % Accepted
+
+%----------------------------------------------------------------------------------------
+%	ABSTRACT
+%----------------------------------------------------------------------------------------
+
+\Abstract{
 In real-time networked applications, such as shared documents, users can
@@
 In this work, we propose a middleware for \emph{symmetric peer-to-peer
-applications} in which decentralized instances can broadcast asynchronous
-events and yet conform to identical behavior.
+applications} in which decentralized instances can trigger asynchronous events
+and yet conform to identical behavior.
 %
@@
 
-\keywords{
-collaboration, determinism, peer-to-peer, time machine
-}
-
-\maketitle
-
-\section{Introduction}
+\Resumo{
+Em aplicações de tempo real em rede, tais como documentos compartilhados, os
+usuários podem interagir remotamente e ainda obterem a mesma experiência, como
+se estivessem em uma única máquina.
+%
+Neste trabalho, nós propomos um middleware para \emph{aplicações peer-to-peer
+simétricas} nas quais instâncias descentralizadas podem disparar eventos
+assíncronos e ainda obterem comportamento idêntico.
+%
+Os pares podem entrar e sair da rede a qualquer momento.
+Os desenvoledores de aplicações devem aderir a uma API restritiva, que deve ser
+determinística, sem estados internos, e também com alocação prévia de memória.
+%
+O middleware é responsável por sincronizar os eventos em uma linha de tempo
+global e compartilhada em toda a rede.
+Além disso, com base em cópias da memória e simulação determinística, uma
+``máquina do tempo'' pode retroceder pares em conflito para ressincronizar o
+seu estado.
+%
+Nós mostramos que o middleware suporta aplicações com mais de 20 pares em uma
+topologia heterogênea e sob alta taxa de desconexões.
+Como exemplo, em uma simulação com 25 eventos por segundo em uma rede com
+latência de 50ms, os pares precisam retroceder e ressincronizar somente 3\% do
+tempo.
+}
+
+%----------------------------------------------------------------------------------------
+
+\begin{document}
+
+%\setcounter{page}{11}
+
+\flushbottom % Makes all text pages the same height
+\maketitle % Print the title and abstract box
+\thispagestyle{empty} % Removes page numbering from the first page
+%\tableofcontents % Print the contents section
+
+\section{Introduction} % The \section*{} command stops section numbering
 \label{sec.introduction}
@@
 Peers in the application form a dynamic network graph and communicate only
-with direct neighbours, as in typical unstructured peer-to-peer networks.
+with direct neighbours, as in typical unstructured peer-to-peer
+networks~\cite{p2p.survey}.
 Events are flooded in the graph and are triggered locally with a small delay
 to compensate the network latency.
-To deal with events received out of order or too late, a distributed time
-machine can rollback peers to a previous state and reapply events in order and
+To deal with events received out of order or too late, a time machine can
+rollback peers locally to a previous state and reapply events in order and
 in time.
@@
     (c) can leave and join the network and remain symmetric.
+We target small peer-to-peer networks, in which nodes are only a few hops away
+from each other.
+This ensures that events can span the whole network in a reasonable time to
+preserve the real-time behavior of applications.
+We also assume that peers are non-malicious in the sense that they do not
+generate erroneous events.
 
@@
 In Section~\ref{sec.related}, we revisit existing solutions for symmetric
-distributed applications.
+distributed applications and software time machines.
 In Section~\ref{sec.tml}, we detail the architecture of our middleware, its
@@
 \label{sec.related}
+
+This section first revisits existing solutions for symmetric distributed
+applications, namely Croquet~\cite{croquet,croquet.new}, GALS~\cite{gals}, and
+CRDTs~\cite{crdts}.
+Then, we discuss software time machines that are built for either local or
+networked applications.
+
+\subsection{Symmetric Distributed Applications}
+\label{sec.related.sym}
 
@@
 
-In this section, we revisit existing solutions for symmetric distributed
-applications.
-We focus on
-    (a) how the network is organized,
-    (b) how global time is determined, and
-    (c) how events are propagated and applied.
-Figure~\ref{fig.table} compares three selected works regarding these aspects.
+Even though existing solutions aim to provide symmetric behavior for
+distributed applications, they diverge in three key aspects as follows:
+
+\textbf{a.} How the network is organized.
+
+\textbf{b.} How global time is determined.
+
+\textbf{c.} How events are propagated and applied.
+
+Figure~\ref{fig.table} compares three selected works regarding these aspects,
+which we discuss and contrast with our approach next.
+At the end of this section, we also provide initial details on how our
+middleware works.
 
@@
 depend on the central server to be online (c).
+%
+Note how the central server is the only participant with knowledge about
+clients, which never communicate directly among themselves.
+This implies that the server must never fail, and also that periodic full
+broadcasts are the only possibility of spanning the whole network.
+In this work, we argue that in a peer-to-peer alternative any node can fail,
+and that broadcasts can be replaced by optimistic flooding.
 
@@
 the server, clients have their own independent local clocks.
-Event generation requires two round trips (\emph{when→ask→ack→tick}) and is
+Event generation requires two round trips (\emph{when → ask → ack → tick}) and is
 delayed dynamically according to the slowest client.
@@
 can communicate events directly to each other.
+Note, that it is not necessary that all peers communicate directly with all
+other peers, but only indirectly via unstructured flooding.
 Events are delayed to be able to reach other peers in time.
@@
 
-We now discuss related work on time machines.
-%
+Finally, note that classical distributed consensus mechanisms focus on fault
+tolerance~\cite{lamport}, but not on real-time symmetric behavior.
+These algorithms typically rely on leaders~\cite{consensus.paxos.raft}, which
+require an election period that does not cope with our symmetric real-time
+requirements.
+In this sense, these algorithms could be considered to decentralize servers in
+Croquet or GALS, but not to decentralize symmetric peers with high churn rates,
+as we propose in this work.
+In contrast, our proposal requires a time machine to reapply reordered
+operations, since we cannot guarantee strong consistency~\cite{p2p.sec}.
+
+\subsection{Software Time Machines}
+\label{sec.related.time}
+
+%We now discuss related work on decentralized and centralized time machines.
+
+A software time machine is a mechanism that allows to regress or advance the
+execution of an application.
+%
+Time travelling is a established research topic for debugging
+tools~\cite{tml.review}.
+In this context, it is possible to stop a program after a failure and revert
+its execution in order to understand the bug causes.
+%
+In our context of real-time symmetric applications, time travelling is
+essential to re-synchronize the execution of mirrored instances that diverge
+due to asynchronous events.
+
+Fundamentally, it is not possible to reverse the execution of most programs.
+The reason is because even trivial operations, such as `printf` and `abs` are
+not invertible.
+Therefore, time machines rely on techniques that either \emph{record} the
+entire application states in a log, or \emph{reconstruct} them from
+deterministic inputs~\cite{tml.review}.
+%
+In this work, we mix both techniques by taking periodic memory snapshots and
+simulating deterministic inputs in between them.
+
 Regarding decentralized machines, Fusion%
@@
 full rollbacks when clients diverge.
-However, the shared mode is less powerful and only supports eventual
-consistency among clients.
-We presume that continuous memory snapshots would be too costly without a
-central server.
-%
+As discussed in Section~\ref{sec.related.sym}, this approach is similar to how
+Croquet handles late joins to a running session.
+The shared mode is less powerful and only supports eventual consistency among
+clients.
+We presume that the reason is because continuous memory snapshots would be too
+costly without a central server.
+
 Regarding single-user local machines, the video game Braid%
@@
 Like Fusion, the implementation is based on continuous memory snapshots,
-instead of deterministic simulation as we propose in this work.
+instead of deterministic simulation, as we propose in this work.
 %
@@
 moving rectangle.
-The six instances are connected in a peer-to-peer network, such that most
-peers can only communicate indirectly to each other.
 %
 The application is symmetric in the sense that if any user presses a key, a
-corresponding event is broadcast and applied in all peers simultaneously, as
-if they were a single mirrored machine.
+corresponding event propagates in the network, and all peers apply it
+simultaneously, as if they were a single mirrored machine.
 At any moment, if a user pauses the application, all peers are guaranteed to
@@
     The middleware orchestrates the peers, synchronizing their ticks locally
-    and broadcasting events with a $\Delta$ delay.
+    and disseminating events with a $\Delta$ delay.
     \label{fig.middleware}
@@
 which synchronizes them with a timestamp with a $\Delta$ delay scheduled to the
-future, such that all peers can satisfy.
+future, such that all peers can (optimistically) satisfy.
 The middleware controls the execution of the application by issuing ticks and
 synchronized events.
+
+The peers form a dynamic unstructured peer-to-peer network~\cite{p2p.survey},
+and communicate mostly indirectly to each other.
+Events are flooded in the network graph via gossiping, i.e.: when a peer
+generates an event, it communicates to its neighbours, which communicate to
+their neighbours, and so on.
+Note that all peers execute the exactly same application and middleware, with
+no differences with respect to their roles and physical resources.
+Therefore, we can describe our peer-to-peer network as follows:
+    (a) all peers have the same role and run the same software;
+    (b) any peer can join or leave at any time;
+    (c) events are only communicated with direct neighbours.
 
@@
 Due to the inherent latency of networks, peers generate events with a
-deliberate $\Delta$ so that they can reach other peers in time to be applied in
-sync.
+deliberate delta $\Delta$ so that they can reach other peers in time to be
+applied in sync.
 %We first assume a predetermined delay for the sake of simplicity.
@@
 application execution from the beginning.
-As an optimization, the middleware takes periodic snapshots locally to avoid
-full simulation.
+As an optimization, the middleware takes periodic snapshots locally at all
+peers to avoid full simulation.
 %
@@
 polling inaccuracies).
-The middleware assumes the maximum time among all peers the be the
+The middleware assumes the maximum time among all peers to be the
 ``correct real time'', so that other peers must fast forward to catch up with
@@
 the middleware proceeds as follows:
-    (a) peers broadcast \code{SYNC} events periodically with their local
+    (a) peers disseminate \code{SYNC} events periodically with their local
         times, and
@@
 timeline.
+In addition, a peer that rejoins the network recovers from a past snapshot it
+already holds, thus avoiding full simulation.
 \end{enumerate}
@@
 
-When an application generates events, they need to be broadcast to the network
+When an application generates events, they need to propagate in the network
 so that all peers behave the same.
@@
 This callback has two goals:
-    (a) map and broadcast application events, and
+    (a) map and propagate application events, and
     (b) provide instant feedback to the user.
 %As mentioned previously,
-Not all local low-level events need to broadcast application events in the
+Not all local low-level events need to correspond to application events in the
 network.
@@
 
-\subsubsection{Event Broadcasting}
+\subsubsection{Event Dissemination}
 \label{sec.tml.middleware.events}
 
-As illustrated in Figure~\ref{fig.middleware}, events are broadcast between
-peers with a timestamp scheduled to the future with a $\Delta$ such that all
-peers are able to apply them in sync.
-%
-To prevent broadcast cycles, each peer keeps a vector with the highest
+As illustrated in Figure~\ref{fig.middleware}, events propagate between peers
+with a timestamp scheduled to the future with an extra delta $\Delta$ such that
+all peers are able to apply them in sync.
+%
+To prevent dissemination cycles, each peer keeps a vector with the highest
 timestamp it received from each other peer, such that lower numbers can be
@@
 When the middleware receives a higher number from a given peer, it updates its
-vector position and triggers a broadcast to neighbours.
+vector position and triggers an event propagation to neighbours.
 %
@@
 simple vector of timestamps.
-Therefore, this is not a severe limitation, and a separate mechanism could be
-used for assignments, with few modifications to the core of the middleware.
 
@@
 }
+
+Considering the peer-to-peer network as a whole, each event packet is
+replicated to all neighbours of all peers.
+This results in a theoretical limit of quadratic messages if all peers were
+connected to all peers.
+In contrast, as discussed in Section~\ref{sec.related.sym}, a centralized
+solution only requires a packet to be sent to a server, which then broadcasts
+it to all clients.
 
@@
 25      if (<next-local-input>) {
-26        // broadcast local event
+26        // propagate local event
 27        p2p_evt evt;
 28        if (cb_evt(&evt)) {
-29          p2p_bcast(<future>, &evt);
+29          p2p_propagate(<future>, &evt);
 30        }
@@
 Regarding local inputs, we call \code{cb\_evt} to signal if the event should be
-broadcast \lin{26--30}, also transforming it to an application event
+disseminated \lin{26--30}, also transforming it into an application event
 (Section~\ref{sec.tml.api.cb_evt}).
@@
 
-\subsubsection{Summary}
+The step delay is dynamically adjusted such that the whole time travel loop
+takes at most 1 second to complete.
+Note that during this period, the application is frozen, and new inputs are
+enqueued for further processing.
+
+\subsubsection{Middleware Summary}
 \label{sec.tml.middleware.summary}
@@
 
+As discussed in Section~\ref{sec.tml.api}, the middleware imposes some
+restrictions, which are discussed as follows:
+
+\begin{itemize}
+%
+\item \textbf{Contiguous unique identifiers:}
+We assume that peers have unique identifiers in a contiguous range, which
+allows us to hold the peers database in a vector.
+Nevertheless, it is possible to generate UUIDs based local information
+only~\cite{p2p.id}, and to use a hash table as database instead.
+%
+\item \textbf{Deterministic API:}
+Since simulations must be equally reproduced in all peers, non-deterministic
+APIs are forbidden.
+This is a fundamental limitation that cannot be easily overcame.
+Note that the alternatives discussed in Section~\ref{sec.related} either impose
+the same limitation~\cite{croquet,gals}, or can be even more restrictive
+allowing only CRDTs~\cite{crdts}.
+%
+\item \textbf{Non-malicious peers:}
+As detailed in Section~\ref{sec.tml.api.cb_evt}, note that the API to
+disseminate an event in the network is as simple as a variable assignment.
+Therefore, a malicious peer could SPAM or inject erroneous events to break the
+application.
+Again, this is a fundamental limitation that is also shared with the
+alternatives discussed in Section~\ref{sec.related}.
+%
+\item \textbf{Static memory footprint:}
+The middleware takes periodic snapshots of the whole application state in order
+to optimize time machine simulations and resynchronize peers.
+This mechanism imposes two challenges:
+first, the snapshot must be small and determined, which implies a contiguous
+static region;
+second, the history of snapshots must be stored in all peers.
+Currently, we allocate periodic snapshots assuming that the memory does not
+deplete until the application terminates.
+Alternative policies could favor recent snapshots or simply impose a limit that
+reject clients that are too late.
+Regarding the static limitation, we mentioned the possibility of a custom
+allocator, which is illustrated as follows:
+\end{itemize}
+
+{\footnotesize
+\begin{verbatim}
+struct {
+    // FIXED REGION
+    int x,  y;   // position
+    int dx, dy;  // direction speed
+    // VARIABLE REGION
+    char heap[HEAP_SIZE];
+} G;
+
+T* obj = my_malloc(G.heap, <size>);
+<...>
+my_free(obj);
+\end{verbatim}
+}
+
+The variable region is still technically static, but is dynamically reshaped
+internally with the custom allocator \code{my\_alloc} \& \code{my\_free}.
+
 %\newpage
@@
 
-\begin{comment}
-$ vi test.c: // #define TEST_OFFLINE
-$ sudo ./test-ana.sh 2>&1 | tee all.log
-$ ./test-sheet.sh > sheet.log
-$ ./test-sheet-2.sh > sheet-2.log
-$ lua5.3 test-sheet-2.lua sheet-2.log > /tmp/x
-$ gedit /tmp/x
-
-$ vi test.c: #define TEST_OFFLINE
-$ sudo ./test2-ana.sh 2>&1 | tee all2.log
-$ lua5.3 test2-sheet.lua
-\end{comment}
-
 Our results show that goals (b) and (c) are met in all scenarios consistently,
@@
 either handle low latency with high event frequency (e.g., $5ms$ with
-$100evt/min$) or high lantency with low event frequency (e.g., $100ms$ with
+$100evt/min$) or high latency with low event frequency (e.g., $100ms$ with
 $5evt/min$).
@@
 
-%\section{Application}
-%\label{sec.app}
+Considering the metrics we evaluate, we now compare our work with the existing
+solutions of Section~\ref{sec.related.sym}:
+
+\textbf{(a) Recurrence of time travels:}
+In Croquet, the centralized reflector issues periodic ticks, which represents
+the only authority over time.
+On the one hand, clients never need to travel back in time, but on the other
+hand, the reflector represents a single point of failure.
+%
+GALS imposes the same tradeoffs, relying on a central server to provide a
+single authority over time.
+%
+CRDT operations can be applied in any order, which also excludes time travels
+entirely.
+However, the notion of a shared real-time timeline cannot exist, which
+precludes bit-identical behavior across peers.
+
+\textbf{(b) Real-time pace of peers:}
+Croquet stipulates a global tick period which all clients must comply with.
+However, clients with poor connections experience constant freezes and cannot
+advance in real time.
+As a concrete example, a peer that experiences a latency of $500ms$, while the
+reflector issues ticks every $100ms$, will freeze for $400ms$, even if no
+events occur in the network.
+%
+GALS supports independent timelines for clients, but at the cost of applying
+events with a greater delay in comparison to Croquet.
+%
+With CRDTs, although peers advance independently, the notion of real time does
+not exist.
+
+\textbf{(c) Correctness of unstable peers:}
+Since clients are connected directly with the server, Croquet and GALS have no
+issues with unstable clients.
+When reconnecting, a client just needs to synchronize pending events from the
+server and apply them in order.
+However, as discussed before, the server represents a single point of failure.
+%
+CRDTs are also immune to unstable peers, since peers just need to synchronize
+events and apply them in any order, but again missing the notion of real-time
+behavior.
 
@@
 We show that peers can
-    (a) broadcast events and meet deadlines globally, 
+    (a) disseminate events and meet deadlines globally, 
     (b) advance in sync in real time, and
@@
 
-\bibliography{spe}
-
+%----------------------------------------------------------------------------------------
+%	REFERENCE LIST
+%----------------------------------------------------------------------------------------
+
+\phantomsection
+
+\makeatletter
+\renewcommand\@biblabel[1]{{\parbox{0.7cm}{[#1]}}}
+\makeatother
+% References
+\renewcommand{\refname}{References}
+\bibliography{rita}
+
+\balance
 \end{document}
